---
title: "Processing_Environmental_Data"
author: "Tessa Rehill"
date: "2024-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r}
library(readxl)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(readr)
library(patchwork)
```

#HOBO LOGGER DATA 
```{r}
#Set the path to your main folder
main_folder <- "./raw_data/HOBO"

#Get a list of site folders within the main folder
site_folders <- list.dirs(main_folder, recursive = FALSE)

#Loop through each site folder
for (site_folder in site_folders) {
  #Get a list of Excel files in the current site folder
  excel_files <- list.files(site_folder, pattern = "\\.xlsx$", full.names = TRUE)
  
  #Read and process each Excel file
  site_data <- map_dfr(excel_files, function(file) {
    data <- read_excel(file)
    
    #Rename columns
    colnames(data) <- c('n', 'date', 'temperature_c', 'light_lux')
    
    #Return the data
    return(data)
  })
  
  #Create the output file path
  output_file <- file.path(site_folder, "combined_dataset.csv")
  
  #Save the combined data for the site to the associated folder
  write.csv(site_data, output_file, row.names = FALSE)
  
  cat(paste("Combined dataset saved in:", output_file, "\n"))
}


#Read each combined file and add another column called site 
ek<- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/HOBO/ED_KING/combined_dataset.csv") 
  ek$Site <- "EK"
dx<- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/HOBO/DIXON/combined_dataset.csv") 
  dx$Site <- "DX"
es <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/HOBO/ELLIS_S/combined_dataset.csv") 
  es$Site <- "ES"
en <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/HOBO/ELLIS_N/combined_dataset.csv") 
  en$Site <- "EN"
tr <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/HOBO/TAYLOR/combined_dataset.csv") 
  tr$Site <- "TR"
nn <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/HOBO/NANAT/combined_dataset.csv") 
  nn$Site <- "NN"

#Combine all site files and add to processed data folder 
hobo_combined <- rbind(ek, dx, es, en, tr, nn)

#Remove anamolies (likely when taking them out)
hobo_combined <- hobo_combined %>%
  filter(temperature_c <= 20)

write.csv(hobo_combined, 
"/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/HOBO/combined_hobo.csv", 
          row.names = FALSE)

#read file back in 
hobo <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/HOBO/combined_hobo.csv")
```

##Light lux over time
```{r}
#Read and preprocess the data
hobo <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/HOBO/combined_hobo.csv")

hobo$date <- as.Date(hobo$date)

hobo_filtered <- hobo %>% #only choose months of study period
  filter(format(date, "%m") %in% c("03", "04", "05", "06", "07"))  
  

#Calculate daily mean light lux by site and date
hobo_daily_mean <- hobo_filtered %>%
  group_by(Site, date) %>%
  summarize(daily_mean_light_lux = mean(light_lux, na.rm = TRUE), .groups = "drop")

#Define a threshold for gaps (e.g., 7 days)
gap_threshold <- 7

#Identify gaps and set light to NA for gaps
hobo_daily_mean <- hobo_daily_mean %>%
  arrange(Site, date) %>%
  group_by(Site) %>%
  mutate(
    time_diff = c(NA, diff(date)),
    daily_mean_light_lux = ifelse(time_diff > gap_threshold, NA, daily_mean_light_lux)  # Break lines at gaps
  ) %>%
  ungroup()  # Make sure to ungroup after modifying data


#Plot the daily mean of light availability over time by site
plot_light <- ggplot(hobo_daily_mean, aes(x = date, y = daily_mean_light_lux, color = Site)) +
  geom_line(aes(group = Site), na.rm = TRUE) +  # Avoid gaps and handle missing values
  labs(
    x = "Date",
    y = "Daily Mean Light (lux)",
    color = "Site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(
    values = c(
      "EK" = "#FF5733",
      "EN" = "#4da607",
      "ES" = "grey",
      "NN" = "#62a4ea",
      "DX" = "#eacf62",
      "TR" = "#076fa6"
    )
  )

plot_light 

#save plot 
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/environmental_supps/meanlight.jpg", plot = plot_light, width = 9, height = 6, dpi = 300)

#save 

start_date <- as.Date("2024-03-01")
end_date <- as.Date("2024-08-01")

light <- plot_light + scale_x_date(limits = c(start_date, end_date)) + theme(axis.title.x = element_blank(), legend.position = "none", axis.text.x = element_blank())

light
```
##Determine flux (light) range by site for supplemental table 
```{r}
hobo_summary <- hobo_daily_mean %>%
  group_by(Site) %>%
  summarise(
    min_light = min(daily_mean_light_lux, na.rm = TRUE),
    max_light = max(daily_mean_light_lux, na.rm = TRUE),
    mean_light = mean(daily_mean_light_lux, na.rm = TRUE)
  )

hobo_summary

#save table 
write.csv(hobo_summary, "/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/Range_supp/light_summary_supp.csv", row.names = FALSE)

```

#ENV LOGGER DATA
```{r}
#Set the path to your main folder
main_folder <- "./raw_data/ENV"

#Get a list of site folders within the main folder
site_folders <- list.dirs(main_folder, recursive = FALSE)

#Loop through each site folder
for (site_folder in site_folders) {
  #Get a list of Excel files in the current site folder
  excel_files <- list.files(site_folder, pattern = "\\.csv$", full.names = TRUE)
  
  #Read and process each Excel file
  site_data <- map_dfr(excel_files, function(file) {
    data <- read_csv(file)
    
    #Rename columns
    colnames(data) <- c('time', 'temperature_c')
    
    #Return the data
    return(data)
  })
  
  #Create the output file path
  output_file <- file.path(site_folder, "combined_dataset.csv")
  
  #Save the combined data for the site to the associated folder
  write.csv(site_data, output_file, row.names = FALSE)
  
  cat(paste("Combined dataset saved in:", output_file, "\n"))
}


#Read each combined file and add another column called site 
ek<- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/ENV/ED_KING/combined_dataset.csv") 
  ek$Site <- "EK"
dx<- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/ENV/DIXON/combined_dataset.csv") 
  dx$Site <- "DX"
es <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/ENV/ELLIS_S/combined_dataset.csv") 
  es$Site <- "ES"
en <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/ENV/ELLIS_N/combined_dataset.csv") 
  en$Site <- "EN"
tr <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/ENV/TAYLOR/combined_dataset.csv") 
  tr$Site <- "TR"
nn <-read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/ENV/NANAT/combined_dataset.csv") 
  nn$Site <- "NN"

#Combine all site files and add to processed data folder 
env_combined <- rbind(ek, dx, es, en, tr, nn)

#Remove anomalies (likely when taking loggers out of ocean )
env_combined <- env_combined %>%
  filter(temperature_c <= 20)

write.csv(env_combined, 
"/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/combined_env.csv", 
          row.names = FALSE)


```

#TILT CURRENT METERS 
```{r}

#Set the path to your main folder
main_folder <- "/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters"

#Get a list of site folders within the main folder
site_folders <- list.dirs(main_folder, recursive = FALSE)

#Loop through each site folder
for (site_folder in site_folders) {
  #Get a list of Excel files in the current site folder
  excel_files <- list.files(site_folder, pattern = "\\.csv$", full.names = TRUE)
  
  #Read and process each Excel file
  site_data <- map_dfr(excel_files, function(file) {
    data <- read_csv(file)
    
    #Rename columns
    colnames(data) <- c('date', 'speed_mpers', 'heading', 'speed_upper', 'speed_lowe', 'tilt_rad','direction','batt_volts', 'temp_c')
    
    #Return the data
    return(data)
  })
  
  #Create the output file path
  output_file <- file.path(site_folder, "combined_dataset.csv")
  
  #Save the combined data for the site to the associated folder
  write.csv(site_data, output_file, row.names = FALSE)
  
  cat(paste("Combined dataset saved in:", output_file, "\n"))
}

column_names <- colnames(read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/EK/combined_dataset.csv", nrows = 1))

ek <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/EK/combined_dataset.csv", 
               skip = 136792, col.names = column_names)
ek$Site <- "EK"

dx <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/DX/combined_dataset.csv", 
               skip = 136792, col.names = column_names)
dx$Site <- "DX"

es <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/ES/combined_dataset.csv", 
               skip = 136792, col.names = column_names)
es$Site <- "ES"

en <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/EN/combined_dataset.csv", 
               skip = 136792, col.names = column_names)
en$Site <- "EN"

tr <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/TR/combined_dataset.csv", 
               skip = 136792, col.names = column_names)
tr$Site <- "TR"

nn <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/raw_data/Tilt_meters/NN/combined_dataset.csv", 
               skip = 136792, col.names = column_names)
nn$Site <- "NN"


#Combine all site files and add to processed data folder 
tilt_combined <- rbind(ek, dx, es, en, tr, nn)


write.csv(tilt_combined, 
"/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/combined_tilt.csv", 
          row.names = FALSE)
```

##Plotting tilt data 
```{r}
tilt <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/combined_tilt.csv")

#convert time to date format
tilt$date <- as.Date(tilt$date)

#filter and process the data
tilt_filtered <- tilt %>%
  filter(speed_mpers < 0.35) %>%
  filter(format(date, "%m") %in% c("03", "04", "05", "06"))  %>% 
  filter(!(date >= "2024-06-24" & date <= "2024-06-28"))  %>% 
  filter(!(date >= "2024-05-15" & date <= "2024-05-17")) %>% 
  filter(!(date >= "2024-07-01" & date <= "2024-12-01"))

  
#smooth the data by calculating daily mean and max variables for each site
tilt_summary <- tilt_filtered %>%
  group_by(Site, date, speed_mpers) %>%
  summarise(mean_speed_mpers = mean(speed_mpers), 
            max_speed_mpers = max(speed_mpers), 
            mean_max_speed = mean(speed_upper), 
            max_max_speed = max(speed_upper))
                                    

write.csv(tilt_summary, "/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/tilt_summary.csv", row.names = FALSE)

# Plot the smoothed data
plot <- ggplot(tilt_summary, aes(x = date, y = mean_speed_mpers, color = Site)) + 
  geom_line() +  # Line plot for temperature over time
  labs(
    title = "Daily Mean Speed Over Time by Site",
    x = "Date",
    y = "Mean Daily Speed (m/s)",
    color = "Site"
  ) +
  theme_minimal() +  # Apply a clean theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for better readability
  ) +
  scale_color_manual(
    values = c(
      "EK" = "#FF5733",
      "EN" = "#4da607",
      "ES" = "grey",
      "NN" = "#62a4ea",
      "DX" = "#eacf62",
      "TR" = "#076fa6"
    )
  ) +
  facet_wrap(~Site, scales = "free_y")  # Create separate panels for each site with individual y-scales

plot


#Smooth the data by calculating daily mean for each site
tilt_plot <- tilt_summary %>%
  group_by(Site, date) %>%
  summarise(mean_speed_mpers = mean(mean_speed_mpers, na.rm = TRUE)) %>%
  ungroup()


# lot the daily mean over time by site
plot_current <- ggplot(tilt_plot, aes(x = date, y = mean_speed_mpers, color = Site)) +
  geom_line(aes(group = Site), na.rm = TRUE) +  # Avoid gaps and handle missing values
  labs(
    x = "Date",
    y = "Daily Mean Current Speed (m/s)",
    color = "Site"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(
    values = c(
      "EK" = "#FF5733",
      "EN" = "#4da607",
      "ES" = "grey",
      "NN" = "#62a4ea",
      "DX" = "#eacf62",
      "TR" = "#076fa6"
    )
  )

plot_current 

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/environmental_supps/current.jpg", plot = plot_current, width = 9, height = 6, dpi = 300)


start_date <- as.Date("2024-03-01")
end_date <- as.Date("2024-08-01")

current <- plot_current + scale_x_date(limits = c(start_date, end_date)) + theme(axis.title.x = element_blank(), legend.position = "none")

current
```
###Determine tilt_plot range by site for supplemental table 
```{r}
tilt_summary <- tilt_plot %>%
  group_by(Site) %>%
  summarise(
    min_current = min(mean_speed_mpers, na.rm = TRUE),
    max_current = max(mean_speed_mpers, na.rm = TRUE),
    mean_current = mean(mean_speed_mpers, na.rm = TRUE)
  )

tilt_summary

#save table 
write.csv(tilt_summary, "/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/Range_supp/current_summary_supp.csv", row.names = FALSE)

```

#Plotting temperature (using both ENV and HOBO)
##read in data and combine + summarize across both dataset 
```{r}
hobo <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/HOBO/combined_hobo.csv")
env <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/combined_env.csv")

hobo <- hobo %>%
  select(-n, -light_lux) %>%
  rename(time = date)

combined_data <- rbind(hobo, env)

#summarize
#Convert time to date format
combined_data$date <- as.Date(combined_data$time)

#Filter and process the data
env_filtered <- combined_data %>%
  filter(format(date, "%m") %in% c("03", "04", "05", "06", "07"))  %>% 
  arrange(Site, date) %>%                 # Sort by Site and date
  group_by(Site) %>%                      # Group by site
  mutate(
    time_diff = c(NA, diff(date)),        # Calculate time differences between dates
    temperature_c = ifelse(time_diff > 7, NA, temperature_c) # Set temperature to NA for gaps > 7 days
  ) %>%
  ungroup() %>% #removing dates within monitoring periods 
  filter(!(date >= "2024-06-24" & date <= "2024-06-28")) %>% # Remove dates between June 24 and June 28
  filter(!(date >= "2024-03-31" & date <= "2024-04-06")) %>% # Remove dates between March 31 and April 6
  filter(!(date >= "2024-05-12" & date <= "2024-05-18")) %>% # Remove dates between March 31 and April 6
  filter(!(date >= "2024-08-04" & date <= "2024-08-10"))
  
#Smooth the data by calculating daily mean temperature for each site
env_smoothed <- env_filtered %>%
  group_by(Site, date) %>%
  summarise(temperature_c = mean(temperature_c, na.rm = TRUE)) %>%
  ungroup()

#Plot the smoothed data
plot <- ggplot(env_smoothed, aes(x = date, y = temperature_c, color = Site)) + 
  geom_line() +  # Line plot for temperature over time
  labs(
    title = "Daily Mean Temperature Over Time by Site",
    x = "Date",
    y = "Temperature (°C)",
    color = "Site"
  ) +
  theme_minimal() +  # Apply a clean theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for better readability
  ) +
  scale_color_manual(
    values = c(
      "EK" = "#FF5733",
      "EN" = "#4da607",
      "ES" = "grey",
      "NN" = "#62a4ea",
      "DX" = "#eacf62",
      "TR" = "#076fa6"
    )
  ) +
  facet_wrap(~Site, scales = "free_y")  # Create separate panels for each site with individual y-scales

plot 


ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/environmental_supps/meantemp_bysite.jpg", plot = plot, width = 9, height = 6, dpi = 300)


#Plot temperature vs. time, colored by site (all on the same graph)
plot_temp <- plot <- ggplot(env_smoothed, aes(x = date, y = temperature_c, color = Site)) + 
  geom_line() +  # Line plot for temperature over time
  labs(
    x = "Date",
    y = "Daily Mean Temperature (°C)",
    color = "Site"
  ) +
  theme_minimal() +  # Apply a clean theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for better readability
  ) +
  scale_color_manual(
    values = c(
      "EK" = "#FF5733",
      "EN" = "#4da607",
      "ES" = "grey",
      "NN" = "#62a4ea",
      "DX" = "#eacf62",
      "TR" = "#076fa6"
    )
  )  # Set custom colors for sites

plot_temp 

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/environmental_supps/meantemp.jpg", plot = plot_temp, width = 9, height = 6, dpi = 300)


start_date <- as.Date("2024-03-01")
end_date <- as.Date("2024-08-01")

temp <- plot_temp + 
  scale_x_date(limits = c(start_date, end_date)) + 
  theme(axis.title.x = element_blank(), axis.text.x = element_blank())
temp

```

###Determine temperature range by site for supplemental table 
```{r}
env_summary <- env_smoothed %>%
  group_by(Site) %>%
  summarise(
    min_temp = min(temperature_c, na.rm = TRUE),
    max_temp = max(temperature_c, na.rm = TRUE),
    mean_temp = mean(temperature_c, na.rm = TRUE)
  )

env_summary

#save table 
write.csv(env_summary, "/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/ENV/Range_supp/temp_summary_supp.csv", row.names = FALSE)

```

#Combining plots 
```{r}

#Combine the two plots vertically
combined_plot <- temp / light / current

#Display the combined plot
print(combined_plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/environmental_supps/combined_env.jpg", plot = combined_plot, width = 9, height = 15, dpi = 300)

```

