---
title: "Density_Analysis"
author: "Tessa Rehill"
date: "2024-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(RColorBrewer)
library(grid) 
library(lubridate)
library(car)
library(dunn.test)
library(stringr)
library(ggrepel)
library(lme4)
library(corrplot)
library(glmmTMB)
library(lmerTest)
library(onewaytests)
library(patchwork)
library(ggpattern)

```

#load data 
```{r}
transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/cleaned_monitoring_transects.csv")

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/cleaned_green_gravel_monitoring.csv")

```

#Depth across sites 
```{r}
#mean depth across sites 
mean(transects$mean_depth_m) #6.4m 
max(transects$mean_depth_m) #11.0 m 
min(transects$mean_depth_m) #4.95

#determining shallowest and deepest site 
#group by site and calculate the mean depth per site
mean_depth_summary <- transects %>%
  group_by(site) %>%
  summarise(
    mean_depth_m = mean(mean_depth_m, na.rm = TRUE)  # Mean depth for each site
  )

#find the shallowest site (minimum mean depth)
shallowest_site <- mean_depth_summary %>%
  filter(mean_depth_m == min(mean_depth_m, na.rm = TRUE))


#find the deepest site (maximum mean depth)
deepest_site <- mean_depth_summary %>%
  filter(mean_depth_m == max(mean_depth_m, na.rm = TRUE))


shallowest_site #Ellis N 5.2m 
deepest_site #Taylor 7.8m 
```

#Urchin densities by species (across sites) ###
```{r}
#reshape the data to long format for stacking
transects_long <- transects %>%
  pivot_longer(cols = c(red_density, green_density, purple_density), 
               names_to = "species", 
               values_to = "density")

#explicitly set the factor levels for 'species' if necessary
transects_long$species <- factor(transects_long$species, 
                                          levels = c("red_density", "green_density", "purple_density"))

#create the stacked bar plot
plot <- ggplot(transects_long, aes(x = site, y = density, fill = species)) + 
  geom_bar(stat = "identity") + 
  labs(
    title = "Urchin Densities by Species", 
    x = "Site", 
    y = "Urchin Density (m²)"
  ) + 
  scale_fill_manual(values = c("red_density" = "red", "green_density" = "green", "purple_density" = "purple")) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #rotate x-axis labels for readability

plot 

#mostly red urchins, probably don't need to visualize this.
unique(transects$total_n_gu) #one occurence of green urchin
unique(transects$total_n_pu) #three occurences of purple urchin

```


#Only baseline urchin densities (April)
```{r}
#create bar chart for total baseline urchin densities (April), across sites 
april_data <- transects %>%
  filter(month == "April") # Ensure the month column matches your dataset's format

#set the desired order for the sites
april_data$site <- factor(april_data$site, 
                                 levels = c("Taylor", "Dixon", "Ed King", "Ellis N", "Ellis S", "Nanat"))

#create the bar plot (black & white)
plot <- ggplot(april_data, aes(x = site, y = urchin_site_density, fill = site)) +
  geom_bar(stat = "identity", color = "black", fill = "grey") +  # Black outlines, grey fill
  
  # Labels and title
  labs(
    #title = "Baseline Urchin Densities (April)",
    x = "Site",
    y = "Urchin Density (m²)"
  ) +
  
  # Theme adjustments
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),  # Larger title, centered
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),  # Larger x-axis labels
    axis.text.y = element_text(size = 14),  # Larger y-axis labels
    axis.title.x = element_text(size = 16, face = "bold"),  # Larger x-axis title
    axis.title.y = element_text(size = 16, face = "bold"),  # Larger y-axis title
    legend.position = "none"  # Remove legend since color is not site-specific
  )

#display the plot
plot


ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/urchin_densities/baseline_densities.png", plot = plot, width = 10, height = 8, dpi = 300)
```

#Density analysis (same as above, do urchin densities differ by site, but remove March and August data)
  #its reasonable to keep month of March even if TR and EK were not surveyed 
```{r}

transects_new <- transects %>%
  filter(!month %in% c("March", "August"))

#testing to see if urchin densities are statistically different from each other 
#check normality of urchin densities per site, if greater than p>0.05 the data is normal 
shapiro.test(transects_new$urchin_site_density) #data is normal

#homogeneity of variances, if less than 0.05 variances are significantly different than each other
leveneTest(urchin_site_density ~ site, data = transects_new) #NOT significantly different from each other

#run a GLM to determine if there are differences in site-level urchin densities among sites
#make sure site is a factor 
transects_new$site <- as.factor(transects_new$site)

glm1 <- glm(urchin_site_density ~ site, data = transects_new, family = gaussian())
summary(glm1) #all significantly different from each other 

#set the desired order for the sites
transects_new$site <- factor(transects_new$site, 
                                 levels = c("Taylor", "Dixon", "Ed King", "Ellis N", "Ellis S", "Nanat"))

plot <- ggplot(transects_new, aes(x = site, y = urchin_site_density, fill = site)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c(
    "Nanat" = "#4da607", 
    "Dixon" = "#f9cd44", 
    "Ellis S" = "#4da607", 
    "Ellis N" = "#f9cd44", 
    "Ed King" = "#f9cd44", 
    "Taylor" = "#cd3737"
  )) + 
  labs(
    title = "Urchin Density Across Sites", 
    x = "Site", 
    y = "Urchin Density (m²)"
  ) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 20), 
        axis.text.y = element_text(size = 20), 
        axis.title.x = element_text(size = 24),  # Increase x-axis title size
      axis.title.y = element_text(size = 24)) + # Increase y-axis title size)  # Increase y-axis label size) +  # Rotate x-axis labels for readability
  guides(fill = "none")  # Remove the legend

plot 

# Save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/urchin_densities/col_box_densities_by_site.png", 
       plot = plot, width = 10, height = 8, dpi = 300)


###adding shapes (to create a legend)
#create a sample dataset
data_legend <- data.frame(
  Site = c("Nanat", "Dixon", "Ellis S", "Ellis N", "Ed King", "Taylor"),
  x = c(1, 2, 3, 4, 5, 6),  # Example x-coordinates
  y = c(2, 3, 4, 5, 6, 7)   # Example y-coordinates
)

# Reorder the Site factor to ensure Taylor appears first
data_legend$Site <- factor(data_legend$Site, levels = c("Taylor", "Dixon", "Ed King","Ellis N",  "Ellis S", "Nanat"))

# Define colors and shapes
colors <- c("Nanat" = "#4da607", 
            "Dixon" = "#f9cd44", 
            "Ellis S" = "#4da607", 
            "Ellis N" = "#f9cd44", 
            "Ed King" = "#f9cd44", 
            "Taylor" = "#cd3737")

shapes <- c("Nanat" = 15,  # Square
            "Ellis S" = 15,  # Square
            "Dixon" = 16,  # Circle
            "Ellis N" = 16,  # Circle
            "Ed King" = 16,  # Circle
            "Taylor" = 17)  # Triangle

# Plot
plot <- ggplot(data_legend, aes(x = x, y = y, color = Site, shape = Site)) +
  geom_point(size = 5) +
  scale_color_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_minimal() +
  labs(color = "Site", shape = "Site") +
  theme(
    legend.text = element_text(size = 24),        # Change to desired size
    legend.title = element_text(size = 20)        # Optional: change legend title size
  )

plot 

# Save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/urchin_densities/legend_densities.png", 
       plot = plot, width = 10, height = 8, dpi = 300)

#define categories
low_density_sites <- c("Nanat", "Ellis S")
medium_density_sites <- c("Ed King", "Dixon", "Ellis N")
high_density_sites <- c("Taylor")

#group data by category and calculate min/max urchin density
urchin_density_summary <- transects_new %>%
  mutate(category = case_when(
    site %in% low_density_sites ~ "Low",
    site %in% medium_density_sites ~ "Medium",
    site %in% high_density_sites ~ "High",
    TRUE ~ "Other"
  )) %>%
  group_by(category) %>%
  summarise(
    min_density = min(urchin_site_density, na.rm = TRUE),
    max_density = max(urchin_site_density, na.rm = TRUE)
  )

#print results
print(urchin_density_summary)

```

#Do urchin densities change across months? 
```{r}

shapiro.test(transects$urchin_site_density)  # Normality4
bartlett.test(urchin_site_density ~ month, data = transects)  # Homogeneity

#run one-way ANOVA
anova_result <- aov(urchin_site_density ~ month, data = transects)
summary(anova_result) #p = 0.596, not significantly different 

#no statistically significant differences in urchin densities between months

#determine mean urchin densities per site
#reorder months in chronological order
transects$month <- factor(transects$month, levels = c("March", "April", "May", "June"))

#calculate mean urchin density per site
mean_urchin_density <- transects %>% 
  group_by(site) %>% 
  summarise(mean_density = mean(urchin_site_density, na.rm = TRUE))

#calculate overall mean urchin density across all sites and months
overall_mean_density <- mean(mean_urchin_density$mean_density, na.rm = TRUE)
overall_mean_density #1.50

overall_sd <- sd(mean_urchin_density$mean_density, na.rm = TRUE)
overall_n <- nrow(mean_urchin_density)  #total number of transects
overall_se <- overall_sd / sqrt(overall_n)

overall_se
overall_n

# Define custom color palette
site_colors <- c("Taylor" = "#cd3737",  
                 "Ed King" = "#f9cd44",  
                 "Dixon" = "#f9cd44",  
                 "Ellis N" = "#f9cd44",  
                 "Ellis S" = "#4da607",  
                 "Nanat" = "#4da607")

# Define custom shape mapping
site_shapes <- c("Nanat" = 15,   # Green square
                 "Ellis S" = 15, # Green square
                 "Ed King" = 16, # Yellow circle
                 "Ellis N" = 16, # Yellow circle
                 "Dixon" = 16,   # Yellow circle
                 "Taylor" = 17)  # Red triangle

# Reorder site factor levels to ensure Taylor appears first in the legend
transects$site <- factor(transects$site, levels = c("Taylor", "Nanat", "Ellis S", "Dixon", "Ellis N", "Ed King"))
mean_urchin_density$site <- factor(mean_urchin_density$site, levels = c("Taylor", "Nanat", "Ellis S", "Dixon", "Ellis N", "Ed King"))

# Create the plot
plot <- ggplot(transects, aes(x = month, y = urchin_site_density, color = site, group = site, shape = site)) +
  geom_line(size = 1) +   # Line plot for each site
  geom_point(size = 3) +  # Add points for each month/site combination
  
  # Plot site mean values as points
  geom_point(data = mean_urchin_density,
             aes(x = "mean", y = mean_density, color = site, shape = site),
             size = 4, show.legend = TRUE) +  

  # Add labels for site mean values with better positioning
  geom_text_repel(data = mean_urchin_density,
                  aes(x = "mean", y = mean_density, label = round(mean_density, 2), color = site),
                  vjust = -1, hjust = -0.2, size = 5, show.legend = FALSE,  
                  box.padding = 0.5, max.overlaps = 10) +  

  # Add the overall mean line and text
  geom_hline(aes(yintercept = overall_mean_density), linetype = "dashed", color = "black") +  
  geom_text(aes(x = 1, y = overall_mean_density, 
                label = paste("Overall Mean:", round(overall_mean_density, 2))),
            color = "black", vjust = 2.5, hjust = -0.1, size = 5) +  

  # Labels and custom colors/shapes
  labs(
    #title = "Urchin Density by Month Across Sites",
    x = "Month",
    y = "Urchin Density (m²)"
  ) +
  scale_x_discrete(labels = c("March", "April", "May", "June", "mean")) +  
  scale_color_manual(values = site_colors) +  
  scale_shape_manual(values = site_shapes) +  # Apply custom shapes
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold"),  
    axis.title = element_text(size = 16, face = "bold"),  
    axis.text = element_text(size = 14),  
    legend.text = element_text(size = 14)  
  )

# Display the plot
print(plot)

#save the plot as a PNG file
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/urchin_densities/densities_month_site_col.png", 
       plot = plot, width = 8, height = 6, dpi = 300)



```

#Comparing urchin densities within caged and uncaged plots 
```{r}

#check distribution of urchin density data - histogram
ggplot(plots, aes(x = urchin_density)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~treatment) +
  labs(title = "Distribution of urchin density by treatment", x = "Urchin Density", y = "Count")


#calculate the mean, standard deviation, sample_size (n) and standard error of the mean 
#where the mean 

#calculate values for reporting 
#calculate the mean and standard error for each treatment
treatment_site <- plots %>%
  group_by(treatment, site) %>%
  summarise(
    site_mean_density = mean(urchin_density, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(treatment) %>%
  summarise(
    mean_density = mean(site_mean_density, na.rm = TRUE),  # Mean of urchin density
    sd_density = sd(site_mean_density, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    sem = sd_density / sqrt(n)                         # Standard error of the mean
  )

#print summary statistics to check the results
print(summary_stats)

#get values for open and caged treatments
open_stats <- summary_stats %>% filter(treatment == "open")
caged_stats <- summary_stats %>% filter(treatment == "caged")

#calculate and format the report output
open_report <- paste(open_stats$mean_density, "±", round(open_stats$sem, 2), "ind/m²")
caged_report <- paste(caged_stats$mean_density, "±", round(caged_stats$sem, 2), "ind/m²")

open_report
caged_report
#Black and white
# Create the bar chart with mean and standard error
plot <- ggplot(summary_stats, aes(x = treatment, y = mean_density, pattern = treatment)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_density - sem, 0),  
            ymax = mean_density + sem
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
  
      geom_text(
        aes(label = paste0("n = ", n), y = mean_density + sem + 0.1),  # Adjust position
        size = 10, 
        color = "black"
    ) +
  
    labs(
        x = "Treatment",
        y = "Urchin Density (m²)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 28, face = "bold"),  
        axis.text.x = element_text(size = 30),  
        axis.text.y = element_text(size = 30),  
        axis.title.x = element_text(size = 30, face = "bold"),  
        axis.title.y = element_text(size = 30, face = "bold")  
    )

# Display the plot
print(plot)

plot <- ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/urchin_densities/bwbar_treatment.png", 
       plot = plot, 
       width = 10, height = 10, dpi = 300)


#sumamrize urchin density by site and month and treatment
urchin_summary <- plots %>%
  group_by(treatment, site, month) %>%
  summarise(
    mean_urchin_density = mean(urchin_density, na.rm = TRUE),
    sd_urchin_density = sd(urchin_density, na.rm = TRUE),
    n = n(),
    se_urchin_density = sd_urchin_density / sqrt(n),
    .groups = "drop"
  ) %>%
  mutate(
    mean_urchin_density = ifelse(site == "Nanat" & treatment == "caged" & month == "August", 0, mean_urchin_density),
    se_urchin_density = ifelse(site == "Nanat" & treatment == "caged" & month == "August", 0, se_urchin_density)
  )


## Reorder factors for plotting
urchin_summary$site <- factor(urchin_summary$site, levels = c("Nanat", "Ellis S", "Ellis N", "Dixon", "Ed King", "Taylor"))
urchin_summary$month <- factor(urchin_summary$month, levels = c("April", "May", "June", "August"))

# Bar plot with patterns and error bars
plot <- ggplot(urchin_summary, aes(x = site, y = mean_urchin_density, pattern = treatment)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_urchin_density - se_urchin_density,
        ymax = mean_urchin_density + se_urchin_density),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "fixed", ncol = 1) +
  labs(
    x = "Site",
    y = "Urchin Density (ind/m²)",
    fill = "Treatment"
  ) +
  scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )

plot 
#save
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/urchin_densities/bwdensity_site_month.png", plot = plot,  width = 15, height = 12, dpi = 300)

```

#Site grazer density 
```{r}
#Assuming 'transects' is your dataframe
#Create a plot of urchin site density by site

# Select relevant columns and reshape the data
grazers_summary <- transects %>%
  dplyr::select(site, total_n_ru, total_n_gu, total_n_pu, total_n_sn, total_n_pyc, total_n_ab) %>%
  pivot_longer(cols = -site, names_to = "species", values_to = "count") %>%
  group_by(site) %>%
  mutate(count = count,  # Keep raw count
         proportion = count / sum(count))  # Calculate proportions per site

grazers_summary$site <- factor(grazers_summary$site, levels = c("Ellis N", "Ellis S", "Taylor", "Dixon", "Ed King", "Nanat" ))
# Define custom colors
grazer_colors <- c(
  "total_n_pu" = "#8a54b3",  # Purple Urchin
  "total_n_gu" = "#4da607",  # Green Urchin
  "total_n_ru" = "#cd3737",  # Red Urchin
  "total_n_pyc" = "#f9cd44", # Pycnopodia
  "total_n_ab" = "#076fa6",  # Abalone
  "total_n_sn" = "#62a4ea"   # Snail
)


# Create the grazer counts stacked bar chart
plot <- ggplot(grazers_summary, aes(x = site, y = count, fill = species)) +  
  geom_bar(stat = "identity") +  
  scale_fill_manual(values = grazer_colors, labels = c("Abalone", "Green Urchin", "Purple Urchin", "Pycnopodia", "Red Urchin", "Turban Snail")) +  
  labs(x = "Site", y = "Counts of Species", fill = "Species") +  
  theme_minimal() +  
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),  # Larger x-axis labels
    axis.text.y = element_text(size = 14),  # Larger y-axis labels
    axis.title.x = element_text(size = 16, face = "bold"),  # Larger x-axis title
    axis.title.y = element_text(size = 16, face = "bold"),  # Larger y-axis title
    legend.title = element_text(size = 16, face = "bold"),  # Larger legend title
    legend.text = element_text(size = 14)  # Larger legend labels
  )

plot
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/grazer_densities/grazer_counts.png", 
       plot = plot, 
       width = 15, height = 10, dpi = 300)
```

#Turban snails densities 
```{r}
#check distribution of urchin density data - histogram
ggplot(plots, aes(x = ts_density)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~treatment) +
  labs(title = "Distribution of turban snail density by treatment", x = "Snail Density", y = "Count")


#calculate values for reporting 
#calculate the mean and standard error for each treatment
treatment_site <- plots %>%
  group_by(treatment, site) %>%
  summarise(
    site_mean_density = mean(ts_density, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(treatment) %>%
  summarise(
    mean_density = mean(site_mean_density, na.rm = TRUE),  # Mean of urchin density
    sd_density = sd(site_mean_density, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    sem = sd_density / sqrt(n)                         # Standard error of the mean
  )

#print summary statistics to check the results
print(summary_stats)

#get values for open and caged treatments
open_stats <- summary_stats %>% filter(treatment == "open")
caged_stats <- summary_stats %>% filter(treatment == "caged")

#calculate and format the report output
open_report <- paste(open_stats$mean_density, "±", round(open_stats$sem, 2), "ind/m²")
caged_report <- paste(caged_stats$mean_density, "±", round(caged_stats$sem, 2), "ind/m²")

open_report
caged_report
# Create the bar chart with mean and standard error
plot <- ggplot(summary_stats, aes(x = treatment, y = mean_density, pattern = treatment)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_density - sem, 0),  
            ymax = mean_density + sem
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
     geom_text(
        aes(label = paste0("n = ", n), y = mean_density + sem + 0.1),  # Adjust position
        size = 10, 
        color = "black"
    ) +
    labs(
        x = "Treatment",
        y = "Red Turban Snail Density (m²)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 28, face = "bold"),  
        axis.text.x = element_text(size = 30),  
        axis.text.y = element_text(size = 30),  
        axis.title.x = element_text(size = 30, face = "bold"),  
        axis.title.y = element_text(size = 30, face = "bold")  
    )

# Display the plot
print(plot)

plot <- ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/turban_densities/bwbar_treatment.png", 
       plot = plot, 
       width = 10, height = 10, dpi = 300)


#sumamrize urchin density by site and month and treatment
turban_summary <- plots %>%
  group_by(treatment, site, month) %>%
  summarise(
    mean_density = mean(ts_density, na.rm = TRUE),
    sd_density = sd(ts_density, na.rm = TRUE),
    n = n(),
    se_density = sd_density / sqrt(n),
    .groups = "drop"
  )


#Ensure 'site' is ordered correctly
turban_summary$site <- factor(turban_summary$site, levels = c("Nanat", "Ellis S", "Ellis N", "Dixon", "Ed King", "Taylor"))
turban_summary$month <- factor(turban_summary$month, levels = c("April", "May", "June", "August"))

# Create the bar chart with pattern instead of color
plot <- ggplot(turban_summary, aes(x = site, y = mean_density, pattern = treatment)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_density - se_density,
        ymax = mean_density + se_density),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "fixed", ncol = 1) +
  labs(
    x = "Site",
    y = "Red Turban Snail Density (m²)",
    fill = "Treatment"
  ) +
  scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )



# Display the plot
print(plot)

#save
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/turban_densities/bwdensity_site_month.png", plot = plot,  width = 15, height = 10, dpi = 300)



```
#Turban snail densities across months 
```{r}

#reorder months in chronological order
transects$month <- factor(transects$month, levels = c("March", "April", "May", "June"))

#calculate mean ts density per site
mean_ts_density <- transects %>% 
  group_by(site) %>% 
  summarise(mean_density = mean(ts_site_density, na.rm = TRUE))

#calculate overall mean ts density across all sites and months
overall_mean_density <- mean(mean_ts_density$mean_density, na.rm = TRUE)
overall_mean_density #1.50

overall_sd <- sd(mean_ts_density$mean_density, na.rm = TRUE)
overall_n <- nrow(mean_ts_density)  #total number of transects
overall_se <- overall_sd / sqrt(overall_n)

overall_se
overall_n

# Define custom color palette
site_colors <- c("Taylor" = "#f9cd44",  
                 "Ed King" = "#f9cd44",  
                 "Dixon" = "#f9cd44",  
                 "Ellis N" = "#cd3737",  
                 "Ellis S" = "#cd3737",  
                 "Nanat" = "#4da607")

# Define custom shape mapping
site_shapes <- c("Nanat" = 15,   # Green square
                 "Ellis S" = 17, # Red triangle
                 "Ed King" = 16, # Yellow circle
                 "Ellis N" = 17, # Red triangle
                 "Dixon" = 16,   # Yellow circle
                 "Taylor" = 16)  # Yellow circle

# Reorder site factor levels to ensure Taylor appears first in the legend
transects$site <- factor(transects$site, levels = c("Taylor", "Nanat", "Ellis S", "Dixon", "Ellis N", "Ed King"))
mean_ts_density$site <- factor(mean_ts_density$site, levels = c("Taylor", "Nanat", "Ellis S", "Dixon", "Ellis N", "Ed King"))

# Create the plot
plot <- ggplot(transects, aes(x = month, y = ts_site_density, color = site, group = site, shape = site)) +
  geom_line(size = 1) +   # Line plot for each site
  geom_point(size = 3) +  # Add points for each month/site combination
  
  # Plot site mean values as points
  geom_point(data = mean_ts_density,
             aes(x = "mean", y = mean_density, color = site, shape = site),
             size = 4, show.legend = TRUE) +  

  # Add labels for site mean values with better positioning
  geom_text_repel(data = mean_ts_density,
                  aes(x = "mean", y = mean_density, label = round(mean_density, 2), color = site),
                  vjust = -1, hjust = -0.2, size = 5, show.legend = FALSE,  
                  box.padding = 0.5, max.overlaps = 10) +  

  # Add the overall mean line and text
  geom_hline(aes(yintercept = overall_mean_density), linetype = "dashed", color = "black") +  
  geom_text(aes(x = 1, y = overall_mean_density, 
                label = paste("Overall Mean:", round(overall_mean_density, 2))),
            color = "black", vjust = 2.5, hjust = -0.1, size = 5) +  

  # Labels and custom colors/shapes
  labs(
    #title = "ts Density by Month Across Sites",
    x = "Month",
    y = "Red Turban Snail Density (m²)"
  ) +
  scale_x_discrete(labels = c("March", "April", "May", "June", "mean")) +  
  scale_color_manual(values = site_colors) +  
  scale_shape_manual(values = site_shapes) +  # Apply custom shapes
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 18, face = "bold"),  
    axis.title = element_text(size = 16, face = "bold"),  
    axis.text = element_text(size = 14),  
    legend.text = element_text(size = 14)  
  )

# Display the plot
print(plot)

#save the plot as a PNG file
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/turban_densities/densities_month_site_col.png", 
       plot = plot, width = 8, height = 6, dpi = 300)

```


