---
title: "Kelp_Survival"
author: "Tessa Rehill"
date: "2024-12-07"
output: html_document
---
#what is driving kelp growth and survival? Or kelp loss... 
#make some visuals of survival and growth 
#test some stats 

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

#Set up libraries 
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(car)
library(multcomp)
library(corrplot)
library(dunn.test)
library(lme4)
library(cowplot)
```

#Read in files 
```{r}

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/cleaned_green_gravel_monitoring.csv")

growth_plots <-  plots[!is.na(plots$kelp_growth), ]


```

#Mean survival + growth at each site at each month, per treatment 
```{r}

#order sites on x-axis 
plots$site <- factor(plots$site, levels = c("Nanat", "Dixon", "Ellis S", "Ellis N", "Ed King", "Taylor"))

#KELP SURVIVAL ###########
#sumamrize urchin density by site and month and treatment
survival_summary <- plots %>%
  group_by(treatment, site, month) %>%
  summarise(
    mean_survival = mean(kelp_survival, na.rm = TRUE),
    sd_survival = sd(kelp_survival, na.rm = TRUE),
    n = n(),
    se_survival = sd_survival / sqrt(n),
    .groups = "drop"
  )

survival_summary$site <- factor(survival_summary$site, levels = c("Nanat", "Ellis S", "Ellis N", "Dixon", "Ed King", "Taylor"))
survival_summary$month <- factor(survival_summary$month, levels = c("April", "May", "June", "August"))

plot <- ggplot(survival_summary, aes(x = site, y = mean_survival, pattern = treatment)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_survival - se_survival,
        ymax = mean_survival + se_survival),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "fixed", ncol = 1) +
  labs(
    x = "Site",
    y = "Kelp Survival (%)",
    fill = "Treatment"
  ) +
  scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )


# Print plot
print(plot)

#save
ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_survival/bwsurvival_site_month.png", plot = plot,  width = 15, height = 10, dpi = 300)

#KELP GROWTH ####
growth_summary <- growth_plots %>%
  group_by(treatment, site, month) %>%
  summarise(
    mean_growth = mean(kelp_growth, na.rm = TRUE),
    sd_growth = sd(kelp_growth, na.rm = TRUE),
    n = n(),
    se_growth = sd_growth / sqrt(n),
    .groups = "drop"
  )

growth_summary$site <- factor(growth_summary$site, levels = c("Nanat", "Ellis S", "Ellis N", "Dixon", "Ed King", "Taylor"))
growth_summary$month <- factor(growth_summary$month, levels = c("April", "May", "June", "August"))

plot <- ggplot(growth_summary, aes(x = site, y = mean_growth, pattern = treatment)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_growth - se_growth,
        ymax = mean_growth + se_growth),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "free_y", ncol = 1) +
  labs(
    x = "Site",
    y = "Kelp Length (cm)",
    fill = "Treatment"
  ) +
  scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )

# Display the plot
print(plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_growth/bwgrowth_site_month.png", plot = plot, width = 15, height = 10, dpi = 300)


legend <- get_legend(plot)
ggdraw(legend)
```

#Mean survival + growth at each site at each month, per gravel size 
```{r}
#SURVIVAL #####################################################################
plots$month <- factor(plots$month, levels = c("April", "May", "June", "August"))

#KELP SURVIVAL ###########
#sumamrize urchin density by site and month and treatment
survival_summary <- plots %>%
  group_by(gravel_size, site, month) %>%
  summarise(
    mean_survival = mean(kelp_survival, na.rm = TRUE),
    sd_survival = sd(kelp_survival, na.rm = TRUE),
    n = n(),
    se_survival = sd_survival / sqrt(n),
    .groups = "drop"
  )

#Ensure 'site' is ordered correctly
survival_summary$site <- factor(survival_summary$site, levels = c("Nanat", "Ellis S", "Ellis N", "Dixon", "Ed King", "Taylor"))
survival_summary$month <- factor(survival_summary$month, levels = c("April", "May", "June", "August"))

# Create the bar chart with pattern instead of color
plot <- ggplot(survival_summary, aes(x = site, y = mean_survival, pattern = gravel_size)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_survival - se_survival,
        ymax = mean_survival + se_survival),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "fixed", ncol = 1) +
  labs(
    x = "Site",
    y = "Kelp Survival (%)",
    fill = "Gravel Size"
  ) +
  scale_pattern_manual(values = c("large" = "stripe", "small" = "none")) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )
# Display the plot
print(plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/gravel_size/bwsurv_gravel_site_month.png", plot = plot, width = 15, height = 10, dpi = 300)


#GROWTH  ##########################################################################
plots$month <- factor(plots$month, levels = c("April", "May", "June", "August"))

#B+W
growth_summary <- growth_plots %>%
  group_by(gravel_size, site, month) %>%
  summarise(
    mean_growth = mean(kelp_growth, na.rm = TRUE),
    sd_growth = sd(kelp_growth, na.rm = TRUE),
    n = n(),
    se_growth = sd_growth / sqrt(n),
    .groups = "drop"
  )

#Ensure 'site' is ordered correctly
growth_summary$site <- factor(growth_summary$site, levels = c("Nanat", "Ellis S", "Ellis N", "Dixon", "Ed King", "Taylor"))
growth_summary$month <- factor(growth_summary$month, levels = c("April", "May", "June", "August"))


plot <- ggplot(growth_summary, aes(x = site, y = mean_growth, pattern = gravel_size)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_growth - se_growth,
        ymax = mean_growth + se_growth),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "free_y", ncol = 1) +
  labs(
    x = "Site",
    y = "Kelp Length (cm)",
    fill = "Gravel Size"
  ) +
  scale_pattern_manual(values = c("small" = "none", "large" = "stripe")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )

# Display the plot
print(plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/gravel_size/bwgrwt_gravel_site_month.png", plot = plot, width = 15, height = 10, dpi = 300)

```

#Barplot of survival and growth, in open v cage
```{r}

#SURVIVAL #####################################################################
#bar chart 
#calculate mean and standard deviation for each treatment
treatment_site <- plots %>%
  group_by(treatment, site) %>%
  summarise(
    site_mean_surv = mean(kelp_survival, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(treatment) %>%
  summarise(
    mean_surv = mean(site_mean_surv, na.rm = TRUE),  # Mean of urchin density
    sd_surv= sd(site_mean_surv, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    se_surv = sd_surv / sqrt(n)                         # Standard error of the mean
  )

summary_stats

#b+w 
plot <- ggplot(summary_stats, aes(x = treatment, y = mean_surv, pattern = treatment)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_surv - se_surv, 0),  
            ymax = mean_surv + se_surv
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
   geom_text(
        aes(label = paste0("n = ", n), y = mean_surv + se_surv + 1),  # Adjust position
        size = 15, 
        color = "black"
    ) +
    labs(
        title = "Kelp Survival by Treatment",
        x = "Treatment",
        y = "Kelp Survival (%)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 38, face = "bold"),  
        axis.text.x = element_text(size = 40),  
        axis.text.y = element_text(size = 40),  
        axis.title.x = element_text(size = 40, face = "bold"),  
        axis.title.y = element_text(size = 40, face = "bold")  
    )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_survival/bwbar_surv_open_v_caged.png", plot = plot, width = 10, height = 10, dpi = 300)


#GROWTH ############################################################

treatment_site <- plots %>%
  group_by(treatment, site) %>%
  summarise(
    site_mean_grw = mean(kelp_growth, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(treatment) %>%
  summarise(
    mean_growth = mean(site_mean_grw, na.rm = TRUE),  # Mean of urchin density
    sd_growth= sd(site_mean_grw, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    se_growth = sd_growth / sqrt(n)                         # Standard error of the mean
  )


summary_stats


plot <- ggplot(summary_stats, aes(x = treatment, y = mean_growth, pattern = treatment)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_growth - se_growth, 0),  
            ymax = mean_growth + se_growth
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
   geom_text(
        aes(label = paste0("n = ", n), y = mean_growth + se_growth + 10),  # Adjust position
        size = 15, 
        color = "black"
    ) +
    labs(
        title = "Kelp Growth by Treatment",
        x = "Treatment",
        y = "Kelp Length (cm)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 38, face = "bold"),  
        axis.text.x = element_text(size = 40),  
        axis.text.y = element_text(size = 40),  
        axis.title.x = element_text(size = 40, face = "bold"),  
        axis.title.y = element_text(size = 40, face = "bold")  
    )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_growth/bwbar_grwt_open_v_caged.png", plot = plot, width = 10, height = 10, dpi = 300)


```

#Looking at GRAVEL SIZE effect on growth and survival
```{r}
#SURVIVAL #####################################################################

treatment_site <- plots %>%
  group_by(gravel_size, site) %>%
  summarise(
    site_mean_surv = mean(kelp_survival, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(gravel_size) %>%
  summarise(
    mean_surv = mean(site_mean_surv, na.rm = TRUE),  # Mean of urchin density
    sd_surv= sd(site_mean_surv, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    se_surv = sd_surv / sqrt(n)                         # Standard error of the mean
  )



summary_stats

#b+w 
plot <- ggplot(summary_stats, aes(x = gravel_size, y = mean_surv, pattern = gravel_size)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_surv - se_surv, 0),  
            ymax = mean_surv + se_surv
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
     geom_text(
        aes(label = paste0("n = ", n), y = mean_surv + se_surv + 3),  # Adjust position
        size = 15, 
        color = "black"
    ) +
    labs(
        title = "Kelp Survival by Gravel Size",
        x = "Gravel Size",
        y = "Kelp Survival (%)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("small" = "none", "large" = "stripe")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 28, face = "bold"),  
        axis.text.x = element_text(size = 30),  
        axis.text.y = element_text(size = 30),  
        axis.title.x = element_text(size = 40, face = "bold"),  
        axis.title.y = element_text(size = 40, face = "bold")  
    )

plot 


ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/gravel_size/bwbar_surv_smallvlarge.png", plot = plot, width = 10, height = 10, dpi = 300)

#GROWTH ############################################################
treatment_site <- plots %>%
  group_by(gravel_size, site) %>%
  summarise(
    site_mean_grw = mean(kelp_growth, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(gravel_size) %>%
  summarise(
    mean_growth = mean(site_mean_grw, na.rm = TRUE),  # Mean of urchin density
    sd_growth= sd(site_mean_grw, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    se_growth = sd_growth / sqrt(n)                         # Standard error of the mean
  )

summary_stats

#Plot

plot <- ggplot(summary_stats, aes(x = gravel_size, y = mean_growth, pattern = gravel_size)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_growth - se_growth, 0),  
            ymax = mean_growth + se_growth
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
       geom_text(
        aes(label = paste0("n = ", n), y = mean_growth + se_growth + 10),  # Adjust position
        size = 15, 
        color = "black"
    ) +
    labs(
        title = "Kelp Growth by Gravel Size",
        x = "Treatment",
        y = "Kelp Length (cm)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("small" = "none", "large" = "stripe")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 28, face = "bold"),  
        axis.text.x = element_text(size = 30),  
        axis.text.y = element_text(size = 30),  
        axis.title.x = element_text(size = 40, face = "bold"),  
        axis.title.y = element_text(size = 40, face = "bold")  
    )
plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/gravel_size/bwbar_growth_smallvlarge.png", plot = plot, width = 10, height = 10, dpi = 300)


plot_with_legend <- plot + theme(legend.position = "right")
legend <- get_legend(plot_with_legend)
ggdraw(legend)


```

#Kelp survival at the end of the study! 
```{r}

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/cleaned_green_gravel_monitoring_Aug.csv")

plots_august <- plots %>%
  filter(month == "August")

#SURVIVAL #####################################################################
#bar chart 
#calculate mean and standard deviation for each treatment
treatment_site <- plots_august %>%
  group_by(treatment, site) %>%
  summarise(
    site_mean_surv = mean(kelp_survival, na.rm = TRUE),  # Mean of urchin density
  )


summary_stats <- treatment_site %>%
  group_by(treatment) %>%
  summarise(
    mean_surv = mean(site_mean_surv, na.rm = TRUE),  # Mean of urchin density
    sd_surv= sd(site_mean_surv, na.rm = TRUE),      # Standard deviation of urchin density
    n = n(),                                           # Sample size
    se_surv = sd_surv / sqrt(n)                         # Standard error of the mean
  )

summary_stats
#b+w 
plot <- ggplot(summary_stats, aes(x = treatment, y = mean_surv, pattern = treatment)) +
    geom_bar_pattern(
        stat = "identity", 
        position = position_dodge(), 
        width = 0.7,
        fill = "white",  # No fill color
        color = "black",  # Add black border to all bars
        pattern_density = 0.05,       
        pattern_spacing = 0.05,       
        pattern_fill = "black",       
        pattern_color = "black",      
        pattern_angle = 45            
    ) +
    geom_errorbar(
        aes(
            ymin = pmax(mean_surv - se_surv, 0),  
            ymax = mean_surv + se_surv
        ),
        width = 0.2,  
        color = "black",
        position = position_dodge(0.7)
    ) +
   geom_text(
        aes(label = paste0("n = ", n), y = mean_surv + se_surv + 1),  # Adjust position
        size = 15, 
        color = "black"
    ) +
    labs(
        #title = "Kelp Survival by Treatment",
        x = "Treatment",
        y = "Kelp Survival (%)"
    ) +
    theme_minimal() +
    scale_pattern_manual(values = c("caged" = "crosshatch", "open" = "none")) +  
    theme(
        legend.position = "none",
        plot.title = element_text(size = 38, face = "bold"),  
        axis.text.x = element_text(size = 40),  
        axis.text.y = element_text(size = 40),  
        axis.title.x = element_text(size = 40, face = "bold"),  
        axis.title.y = element_text(size = 40, face = "bold")  
    )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_survival/bwbar_surv_open_v_caged_august.png", plot = plot, width = 10, height = 10, dpi = 300)


```
#Summary figures

```{r}
plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/processed_data/cleaned_green_gravel_monitoring.csv")

#SURVIVAL ####

#determine site means 
site_means <- plots %>%
  group_by(site, treatment, gravel_size, month) %>%
  summarise(
    mean_site_surv = mean(kelp_survival, na.rm = TRUE),
    .groups = "drop"
  )

#summarize across treatments
summary_stats <- site_means %>%
  group_by(treatment, gravel_size, month) %>%
  summarise(
    mean_survival = mean(mean_site_surv, na.rm = TRUE),
    sd_surv   = sd(mean_site_surv, na.rm = TRUE),
    n         = n(),  # should be 6 sites per group
    se_survival   = sd_surv / sqrt(n),
    .groups   = "drop"
  )

survival_summary <- summary_stats

survival_summary$month <- factor(survival_summary$month, levels = c("April", "May", "June", "August"))

# Create the bar chart with pattern instead of color
plot <- ggplot(survival_summary, aes(x = treatment, y = mean_survival, pattern = gravel_size)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_survival - se_survival,
        ymax = mean_survival + se_survival),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "fixed", ncol = 1) +
  labs(
    x = "Treatment",
    y = "Kelp Survival (%)",
    fill = "Gravel Size"
  ) +
  scale_pattern_manual(values = c("large" = "stripe", "small" = "none")) +
  scale_y_continuous(limits = c(0, 100)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )
# Display the plot
print(plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_survival/summary_plot.png", plot = plot, width = 10, height = 10, dpi = 300)

#GROWTH ####

#determine site means 
site_means <- plots %>%
  group_by(site, treatment, gravel_size, month) %>%
  summarise(
    mean_site_grw = mean(kelp_growth, na.rm = TRUE),
    .groups = "drop"
  )

#summarize across treatments
summary_stats <- site_means %>%
  group_by(treatment, gravel_size, month) %>%
  summarise(
    mean_growth = mean(mean_site_grw, na.rm = TRUE),
    sd_surv   = sd(mean_site_grw, na.rm = TRUE),
    n         = n(),  # should be 6 sites per group
    se_growth   = sd_surv / sqrt(n),
    .groups   = "drop"
  )

growth_summary <- summary_stats


growth_summary$month <- factor(growth_summary$month, levels = c("April", "May", "June", "August"))


plot <- ggplot(growth_summary, aes(x = treatment, y = mean_growth, pattern = gravel_size)) +
  geom_bar_pattern(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.7,
    fill = "white",
    color = "black",
    pattern_density = 0.05,
    pattern_spacing = 0.05,
    pattern_fill = "black",
    pattern_color = "black",
    pattern_angle = 45,
    pattern_alpha = 1
  ) +
  geom_errorbar(
    aes(ymin = mean_growth - se_growth,
        ymax = mean_growth + se_growth),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"
  ) +
  facet_wrap(~ month, scales = "free_y", ncol = 1) +
  labs(
    x = "Treatment",
    y = "Kelp Length (cm)",
    fill = "Gravel Size"
  ) +
  scale_pattern_manual(values = c("small" = "none", "large" = "stripe")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 30),
    axis.text.y = element_text(size = 40),
    axis.title.x = element_text(size = 40, face = 'bold'),
    axis.title.y = element_text(size = 40, face = 'bold'),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(size = 40)
  )

# Display the plot
print(plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Grazer_exclusion/plots/kelp_growth/summary_plot.png", plot = plot, width = 10, height = 10, dpi = 300)
```

