---
title: "GLMMs"
author: "Tessa Rehill"
date: "2025-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 

```{r}
library(here)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)
library(broom)
library(lme4)
library(lmerTest)
```

#Load data 
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")

#remove growth = 0 from dataset 
growth_data <- data %>% drop_na(kelp_growth)

```

#Correlation among ABIOTIC predictors
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "max_temperature_c",  "mean_current", "max_current", "bottom_salinity",  "bottom_do" )], use = "complete.obs")

#rename variables
# Rename row and column names
colnames(cor_matrix) <- rownames(cor_matrix) <- c("Light (lux)", "Temp (mean)", "Temp (max)", 
                                                  "Current (mean)", "Current (max)", "Bottom salinity", "Bottom DO")

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/abiotic_correlation_plot.png", plot = plot, width = 5, height = 5, dpi = 300)

```
#Standardize and scale predictors 
```{r}
#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$max_temperature_c <- scale(data$max_temperature_c)
data$mean_light_lux <- scale(data$mean_light_lux)
data$bottom_do <- scale(data$bottom_do)
data$bottom_salinity <- scale(data$bottom_salinity)
data$urchin_density <- scale(data$urchin_density)
data$urchin_site_density <- scale(data$urchin_site_density)
data$mean_current <- scale(data$mean_current)
data$ts_density <- scale(data$ts_density)
data$ts_site_density <- scale(data$ts_site_density)
```


#Investigating the abiotic variables for kelp survival and growth 

##Kelp survival
###base model + predictor 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment  + (1 | site), 
                      family = binomial, 
                      data = data)

#predictors - makes sense, 
#predictor_list<- c("grazers_site_density",  "grazer_density","mean_temperature_c", "max_temperature_c", "mean_current","gravel_size", "mean_light_lux","bottom_do", "bottom_salinity")

###
predictor_list<- c( "urchin_site_density", "urchin_density", "ts_density", "ts_site_density",  "gravel_size", "mean_temperature_c", "max_temperature_c", "mean_light_lux", "mean_current",  "bottom_do","bottom_salinity", "month")


```

###forward + backward predictor selection based on AIC 
```{r}
#initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

#stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment +", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment +", 
            paste(remaining_predictors, collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

#print results
print(selected_predictors)
summary(current_model)

exp(cbind(OR = fixef(current_model)$cond, confint(current_model, parm = "beta_", method = "Wald")))
``` 

  #final_model 
```{r}
survival_model <- current_model
```

  #look at model selection via dredging 
```{r}
#scaled_data <-  data %>%
#  mutate(across(c(urchin_site_density, urchin_density, ts_density, ts_site_density, 
#                  mean_temperature_c, max_temperature_c, mean_light_lux, 
#                 mean_current, max_current, bottom_salinity, bottom_do, bottom_ph), 
#                scale)) %>%
#  na.omit(data)



#global_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment  + urchin_site_density + urchin_density + ts_site_density + ts_density +  gravel_size  + mean_temperature_c + max_temperature_c +  mean_light_lux + mean_current + bottom_do + bottom_salinity + (1 | site), 
                      #family = binomial(link="logit"), 
                      #data = scaled_data, 
                      #na.action = "na.fail")  # Ensure missing data handling is explicit)

#summary(global_model)


#b <- dredge(global_model, beta = c("none", "sd", "partial.sd"), evaluate = TRUE,
 # rank = "AICc", fixed = "cond(treatment)")


#get.models(b, subset = 1)[[1]] 
```
  
  #visualize (+ save)
```{r}

#final model
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/survival_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/survival_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/suvival_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()



```

  #collinearity 
```{r}
lm_model <- lm(n_gravel ~ treatment +ts_site_density + gravel_size + mean_current + mean_light_lux, data = data)

vif(lm_model, type="predictor") 


```

##Growth models 
  #base model + predictors 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(kelp_growth ~ treatment + (1 | site), 
                      family = Gamma(link="log"), 
                      data = growth_data)

#predictors 
#predictor_list<- c("grazers_site_density",  "grazer_density","mean_temperature_c", "max_temperature_c", "mean_current","gravel_size", "mean_light_lux","bottom_do", "bottom_salinity")


predictor_list<- c("urchin_site_density", "urchin_density", "ts_density", "ts_site_density",  "gravel_size", "mean_temperature_c", "max_temperature_c", "mean_light_lux", "mean_current",  "bottom_do","bottom_salinity")


```

  #chose model based on  forward and backward feature seleciton
```{r}
# Initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

# Stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("kelp_growth ~ treatment +", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = Gamma(link="log"), data = growth_data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("kelp_growth ~ treatment +", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = Gamma(link="log"), data = growth_data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

# Print results
print(selected_predictors)
summary(current_model)

```
  #model selection via dredging 
```{r}
#predictor_list<- c( "urchin_site_density", "urchin_density", "ts_density", "ts_site_density",  "gravel_size", "mean_temperature_c", "max_temperature_c", "mean_light_lux", "mean_current",  "bottom_do","bottom_salinity")

#remove growth = 0 from dataset 
#growth_data <- data %>% drop_na(kelp_growth)

#scaled_growth_data <-  growth_data %>%
#  mutate(across(c(urchin_site_density, urchin_density, ts_density, ts_site_density, 
#                  mean_temperature_c, max_temperature_c, mean_light_lux, 
#                  mean_current, bottom_salinity, max_current, bottom_do, bottom_ph), 
#               scale))%>%
#  na.omit(scaled_growth_data)

#global_model <- glmmTMB(kelp_growth ~ treatment  +  urchin_site_density + urchin_density + ts_site_density + ts_density + gravel_size  + mean_temperature_c + max_temperature_c +  mean_light_lux + mean_current + bottom_do + bottom_salinity + (1 | site), 
#                      family = Gamma(link="log"), 
#                     data = scaled_growth_data, 
#                     na.action = "na.fail")  # Ensure missing data handling is explicit)

#a <- dredge(global_model, beta = c("none", "sd", "partial.sd"), evaluate = TRUE,
#  rank = "AICc", fixed = "cond(treatment)")

#get.models(a, subset = 1)[[1]] 


```


```{r}
lm_model <- lm(kelp_growth ~ treatment + ts_density + mean_temperature_c + mean_light_lux + mean_current + urchin_site_density + urchin_density + gravel_size + bottom_do, data = growth_data)

vif(lm_model, type="predictor") 

vif(lm_model) 
```
 
 #visualize 
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput


#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()


#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```

  #final model 
```{r}
growth_model <- current_model
```

#Combine growth and survival model (co-efficient plot)

#PLOT ALL COEFFICIENT FIGURES 
```{r}
# Extract coefficients and confidence intervals for each model
kelp_model_tidy <- tidy(survival_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Survival GLMM")

growth_model_tidy <- tidy(growth_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Growth GLMM")


# Combine all models into one data frame
all_models <- bind_rows(kelp_model_tidy, growth_model_tidy)

# Filter out random effects (if you only want fixed effects)
all_models <- all_models %>%
  filter(effect == "fixed" | is.na(effect))  # GLM doesn't have an 'effect' column, so allow NA

# Define the color scheme for each model
model_colors <- c(
  "Kelp Survival GLMM" = "#eacf62",
  "Kelp Growth GLMM" = "#FF5733"
)

# Define the desired order of predictors in reverse
predictor_order <- rev(c('treatmentopen', 'gravel_sizesmall', 'urchin_site_density','urchin_density', 'ts_site_density', 'ts_density', 
                         "max_temperature_c", "mean_temperature_c", "mean_light_lux",'mean_current', 'bottom_do', 'bottom_salinity','(Intercept)'))


# Modify the ggplot code to reorder the terms and increase font sizes
plot <- ggplot(all_models, aes(x = factor(term, levels = predictor_order),  
                               y = estimate, ymin = conf.low, ymax = conf.high,  
                               color = model)) +    
  geom_pointrange(position = position_dodge(width = 0.5)) +    
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +    
  coord_flip() +    
  scale_color_manual(values = model_colors) +    
  labs(
    title = "Coefficient and Confidence Interval Plot",
    x = "Predictors",
    y = "Estimate",
    color = "Model"
  ) +    
  theme_minimal() +
  theme(
    text = element_text(size = 18),         # Base text size
    axis.text = element_text(size = 16),    # Axis tick labels
    axis.title = element_text(size = 18),   # Axis titles
    legend.text = element_text(size = 16),  # Legend text
    legend.title = element_text(size = 18), # Legend title
    plot.title = element_text(size = 20, face = "bold") # Plot title
  )

plot

plot <- ggplot(
  all_models %>% filter(!term %in% c("mean_current", "(Intercept)")),  # Remove 'current' and 'intercept'
  aes(
    x = factor(term, levels = predictor_order[!predictor_order %in% c("mean_current", "(Intercept)")]),  # Remove from ordering
    y = estimate, ymin = conf.low, ymax = conf.high,
    color = model
  )) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  coord_flip() +
  scale_color_manual(values = model_colors) +
  labs(
    title = "Coefficient and Confidence Interval Plot",
    x = "Predictors",
    y = "Estimate",
    color = "Model"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    plot.title = element_text(size = 20, face = "bold")
  )

plot



ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/combined_dot_and_whisker_plot.png", 
       plot = plot, width = 15, height = 10, dpi = 300)

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
