---
title: "Regressions"
author: "Tessa Rehill"
date: "2025-01-19"
output: html_document
---

Run linear regressions on response variables (urchin density, kelp survival, kelp growth) to environmental and other predictors.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)
library(GGally)
```


#Read in data 
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")

```

#Supplemental Pairwise Correlation Plot (ENV variables)
```{r}

#variables of interest (ONLY env variables) - mean_temperature_c, mean_light_lux, bottom_do, bottom_salinity, bottom_ph, mean_depth_m, max_temperature_c, max_current, mean_current

#select only the environmental variables
env_data <- data %>%
  dplyr::select(mean_temperature_c, mean_light_lux, bottom_do, bottom_salinity, 
         bottom_ph, mean_depth_m, max_temperature_c, max_current, mean_current) %>%
  rename(
    `temp (mean)` = mean_temperature_c,
    `temp (max)` = max_temperature_c,
    `light (lux)` = mean_light_lux,
    `bottom DO` = bottom_do,
    `bottom pH` = bottom_ph,
    `bottom salinity` = bottom_salinity,
    `depth (mean)` = mean_depth_m,
    `current (max)` = max_current,
    `current (mean)` = mean_current
  )


#define a custom function for the lower panel with red trend lines
custom_smooth <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_point(alpha = 0.6) +  # Keeps points semi-transparent
    geom_smooth(method = "lm", color = "red", size = 0.5, se = FALSE)  # Red, thin regression line
}

#create the scatterplot matrix
p <- ggpairs(env_data, 
             lower = list(continuous = custom_smooth)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        axis.text.y = element_text(angle = 45, hjust = 1))  # Rotate y-axis labels
p

# Save the plot as a PNG file
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/environmental_supps/environmental_scatterplot_matrix.png", plot = p, width = 10, height = 10, dpi = 300)


```

#Regressions - Growth
```{r}

#temperature 
ggplot(data, aes(x = mean_temperature_c, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and Mean Temperature",
    x = "Mean Temperature (°C)",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()


#site density 
ggplot(data, aes(x = urchin_site_density, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and Urchin Density ",
    x = "Urchin Density",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()

#pH
ggplot(data, aes(x = bottom_ph, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and pH",
    x = "pH",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()

#DO
ggplot(data, aes(x = bottom_do, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and DO",
    x = "DO",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()


#salinity
ggplot(data, aes(x = bottom_salinity, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and Salinity",
    x = "Bottom salinity",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()

#co-occuring algae
ggplot(data, aes(x = percent_occuring_algae, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and Algae",
    x = "Co-occuring Algae (%)",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()

#grazer density
ggplot(data, aes(x = grazer_density, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and Grazer Density",
    x = "Grazer Density",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()

#urchin density
ggplot(data, aes(x = urchin_density, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationship Between Kelp Growth and Urchin Density",
    x = "Urchin Density",
    y = "Log - Kelp Growth (mm)"
  ) +
  theme_minimal()


```

#Kelp Growth
```{r}

#reshape the data to a long format
long_data <- data %>%
  pivot_longer(
    cols = c(mean_temperature_c, urchin_site_density, bottom_ph, bottom_do, 
             bottom_salinity, percent_occuring_algae, grazer_density, urchin_density, mean_light_lux, grazers_site_density, mean_current, max_temperature_c),
    names_to = "variable",
    values_to = "value"
  )

#create the faceted plot
ggplot(long_data, aes(x = value, y = kelp_growth)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationships Between Kelp Growth and Environmental Variables",
    x = "Environmental Variable",
    y = "Kelp Growth (mm)"
  ) +
  theme_minimal() +
  facet_wrap(~variable, scales = "free_x", ncol = 3)  # Facet by variable, adjust columns as needed

```

#Kelp survival 
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")

#reshape the data to a long format
long_data <- data %>%
  pivot_longer(
    cols = c(mean_temperature_c, urchin_site_density, bottom_ph, bottom_do, 
             bottom_salinity, percent_occuring_algae, grazer_density, urchin_density, mean_light_lux),
    names_to = "variable",
    values_to = "value"
  )

#create the faceted plot
ggplot(long_data, aes(x = value, y = kelp_survival)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationships Between Kelp Survival and Environmental Variables",
    x = "Environmental Variable",
    y = "Survival (%)"
  ) +
  theme_minimal() +
  facet_wrap(~variable, scales = "free_x", ncol = 3)  # Facet by variable, adjust columns as needed
```

#Urchin density 
```{r}
#reshape the data to a long format
long_data <- data %>%
  pivot_longer(
    cols = c(mean_temperature_c, bottom_ph, bottom_do, 
             bottom_salinity, percent_occuring_algae, grazer_density,  mean_light_lux),
    names_to = "variable",
    values_to = "value"
  )

#create the faceted plot
ggplot(long_data, aes(x = value, y = urchin_density)) +
  geom_point(color = "blue", alpha = 0.6) +  # Points
  geom_smooth(method = "lm", color = "red", linetype = "dashed") +  # Linear regression line
  labs(
    title = "Relationships Between Urchin Densities and Environmental Variables",
    x = "Environmental Variable",
    y = "Urchin Density"
  ) +
  theme_minimal() +
  facet_wrap(~variable, scales = "free_x", ncol = 3)  # Facet by variable, adjust columns as needed
```

#distinguishing between open + caged 
```{r}
data <- data %>%
  dplyr::select(
    percent_occuring_algae, kelp_survival, urchin_density, kelp_growth,
    mean_temperature_c, mean_light_lux, bottom_temp, bottom_do,
    bottom_salinity, bottom_ph, mean_depth_m, urchin_site_density,
    species_richness, shannon_div, simp_div, grazer_density, site, month, treatment
  ) 

# Filter data for caged and open treatments
caged_data <- data %>% 
  filter(treatment == "caged") %>%
  group_by(site, month) %>%
  summarise(across(
    c(percent_occuring_algae:grazer_density), 
    \(x) mean(x, na.rm = TRUE)  # Updated to use an anonymous function
  ), .groups = "drop")

open_data <- data %>% 
  filter(treatment == "open") %>%
  group_by(site, month) %>%
  summarise(across(
    c(percent_occuring_algae:grazer_density), 
    \(x) mean(x, na.rm = TRUE)
  ), .groups = "drop")


```

#run regressions 
```{r}

# Define the response variables to analyze
response_vars <- c("kelp_survival", "kelp_growth", "urchin_density")

#reshape data for plotting (CAGED treatment example)
caged_long <- caged_data %>%
  pivot_longer(cols = all_of(response_vars), 
               names_to = "response_variable", 
               values_to = "value")

open_long <- open_data %>%
  pivot_longer(cols = all_of(response_vars), 
               names_to = "response_variable", 
               values_to = "value")

# Create regression plots with facets

#TEMPERATURE ################################################################
plot1 <- ggplot(caged_long, aes(x = mean_temperature_c, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Mean Temperature (°C)", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs Temperature") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = mean_temperature_c, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Mean Temperature (°C)", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs Temperature") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)

#pH ####################################################################
plot1 <- ggplot(caged_long, aes(x = bottom_ph, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Bottom pH", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs pH") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = bottom_ph, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Bottom pH", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs pH") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)

#Grazer density ######################################################
plot1 <- ggplot(caged_long, aes(x = grazer_density, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Grazer Density (m^2)", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs Grazer Density") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = grazer_density, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Grazer Density (m^2)", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs Grazer Density") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)

#Algae Comp.  ######################################################
plot1 <- ggplot(caged_long, aes(x = percent_occuring_algae, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Occuring Algae (%)", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs Occuring Algae (%)") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = percent_occuring_algae, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Occuring Algae (%)", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs Occuring Algae (%)") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)


#DO 
plot1 <- ggplot(caged_long, aes(x = bottom_do, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "DO", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs DO") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = bottom_do, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "DO", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs DO") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)

#Salinity 
plot1 <- ggplot(caged_long, aes(x = bottom_salinity, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Salinity", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs Salinity") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = bottom_salinity, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Salinity", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs Salinity") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)

#Light Lux 
plot1 <- ggplot(caged_long, aes(x = mean_light_lux, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Mean Light Lux", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs Mean Light Lux") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = mean_light_lux, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Mean Light Lux", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs Mean Light Lux") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)

#Site density 
plot1 <- ggplot(caged_long, aes(x = urchin_site_density, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Site Density", y = "Response Value", 
       title = "Caged Plots - Linear Regression of Response Variables vs Site Density") +
  theme_minimal()

plot2 <- ggplot(open_long, aes(x = urchin_site_density, y = value)) + 
  geom_point(alpha = 0.6) +  # Add scatter plot
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +  # Linear regression line
  facet_wrap(~response_variable, scales = "free_y") +  # Facet by response variable
  labs(x = "Site Density", y = "Response Value", 
       title = "Open Plots - Linear Regression of Response Variables vs Site Density") +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2)


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
