---
title: "Density_Analysis"
author: "Tessa Rehill"
date: "2024-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(RColorBrewer)
library(grid) 
library(lubridate)
library(car)
library(dunn.test)
library(stringr)
library(ggrepel)
library(lme4)
library(corrplot)
library(glmmTMB)
library(lmerTest)
library(onewaytests)
library(patchwork)

```

#load data 
```{r}
transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_monitoring_transects.csv")

transect_summary <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/density_urchin_transects.csv")

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

env <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_ysi_measurement.csv")

hobo <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/ENV/HOBO/summary_hobo.csv")
```

#Depth across sites 
```{r}
#mean depth across sites 
mean(transects$depth_m) #6.4m 
max(transects$depth_m) #11.0 m 
min(transects$depth_m) #0.7m 

#determining shallowest and deepest site 
#group by site and calculate the mean depth per site
mean_depth_summary <- transects %>%
  group_by(site) %>%
  summarise(
    mean_depth_m = mean(depth_m, na.rm = TRUE)  # Mean depth for each site
  )

#find the shallowest site (minimum mean depth)
shallowest_site <- mean_depth_summary %>%
  filter(mean_depth_m == min(mean_depth_m, na.rm = TRUE))


#find the deepest site (maximum mean depth)
deepest_site <- mean_depth_summary %>%
  filter(mean_depth_m == max(mean_depth_m, na.rm = TRUE))


shallowest_site
deepest_site 
```

#Urchin densities by species (across sites) ###
```{r}
#reshape the data to long format for stacking
transects_summary_long <- transects_summary %>%
  pivot_longer(cols = c(red_density, green_density, purple_density), 
               names_to = "species", 
               values_to = "density")

#explicitly set the factor levels for 'species' if necessary
transects_summary_long$species <- factor(transects_summary_long$species, 
                                          levels = c("red_density", "green_density", "purple_density"))

#create the stacked bar plot
plot <- ggplot(transects_summary_long, aes(x = site, y = density, fill = species)) + 
  geom_bar(stat = "identity") + 
  labs(
    title = "Urchin Densities by Species", 
    x = "Site", 
    y = "Urchin Density (m^2)"
  ) + 
  scale_fill_manual(values = c("red_density" = "red", "green_density" = "green", "purple_density" = "purple")) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #rotate x-axis labels for readability

plot 

#save
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/densities_by_sp.png", plot = plot, width = 10, height = 8, dpi = 300)

#mostly red urchins, probably don't need to visualize this.
unique(transects_summary$total_n_gu) #one occurence of green urchin
unique(transects_summary$total_n_pu) #three occurences of purple urchin

```
#Are urchin densities significantly different across sites? 
```{r}
#testing to see if urchin densities are statistically different from each other 
#check normality of urchin densities per site, if greater than p>0.05 the data is normal 
shapiro.test(transects_summary$urchin_density) #p = 0.1757, normal

#homogeneity of variances, if less than 0.05 variances are significantly different than each other
leveneTest(urchin_density ~ site, data = transects_summary) #p = 0.0513, marginal 

#conduct another test to test variances 
#run the Brown-Forsythe test
bf_test <- bf.test(urchin_density ~ site, data = transects_summary)

#display the results, 
bf_test #variances are NOT equal across groups

#unequal variances, so we'll use a kruschal test, if p<0.05 there is significance 
kruskal.test(urchin_density ~ site, data = transects_summary) #p = 0.003271

#determine which sites are significantly different from each other, use Dunn's test 
#Dunn's test for pairwise comparisons
#this is how we can determine threshold for 'low', 'med', 'high' urchin densities 
dunn.test(transects_summary$urchin_density, transects_summary$site, kw = TRUE, label = TRUE) 

#significant differences; 
  #Ellis S v Nanat
  #Nanat v Taylor 
  #Taylor v Ellis S

#non-significant 
  #Ed king v Dixon 
  #Ellis S v Ellis N 
  #Taylor v Ellis N
  #Taylor v Ed King 

#set the desired order for the sites
transects_summary$site <- factor(transects_summary$site, 
                                 levels = c("Taylor", "Ed King", "Dixon", "Ellis N", "Ellis S", "Nanat"))
#visualize via boxplot 
ggplot(transects_summary, aes(x = site, y = urchin_density)) +
  geom_boxplot() +
  labs(
    title = "Urchin Density Across Sites",
    x = "Site",
    y = "Urchin Density (m^2)"
  ) +
  theme_minimal()

#color based on urchin density 
ggplot(transects_summary, aes(x = site, y = urchin_density, fill = site)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c(
    "Nanat" = "green", 
    "Dixon" = "orange", 
    "Ellis S" = "orange", 
    "Ellis N" = "orange", 
    "Ed King" = "orange", 
    "Taylor" = "red"
  )) + 
  labs(
    title = "Urchin Density Across Sites", 
    x = "Site", 
    y = "Urchin Density (m^2)"
  ) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  guides(fill = "none")  # Remove the legend

#save
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/box_densities_by_site.png", plot = plot, width = 10, height = 8, dpi = 300)


#add category to transect summary
transects_summary <- transects_summary %>%
  mutate(density_cat = case_when(
    site %in% c("Nanat") ~ "low",
    site %in% c("Ed King", "Ellis N", "Dixon", "Ellis S") ~ "medium",
    site == "Taylor" ~ "high",
    TRUE ~ "unknown"  # This line is for any site that doesn't match the above, if necessary
  ))

```

#Only baseline urchin densities (April)
```{r}
#create bar chart for total baseline urchin densities (April), across sites 
april_data <- transects_summary %>%
  filter(month == "April") # Ensure the month column matches your dataset's format

#set the desired order for the sites
april_data$site <- factor(april_data$site, 
                                 levels = c("Taylor", "Dixon", "Ed King", "Ellis N", "Ellis S", "Nanat"))
# Create the bar plot
ggplot(april_data, aes(x = site, y = urchin_density)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Baseline Urchin Densities (April)",
    x = "Site",
    y = "Urchin Density (m^2)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  #rotate x-axis labels for readability

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/baseline_densities.png", plot = plot, width = 10, height = 8, dpi = 300)
```

#Density analysis (same as above, do urchin densities differ by site, but remove March and August data)
  #its reasonable to keep month of March even if TR and EK were not surveyed 
```{r}

transects_summary_new <- transects_summary %>%
  filter(!month %in% c("March", "August"))

#testing to see if urchin densities are statistically different from each other 
#check normality of urchin densities per site, if greater than p>0.05 the data is normal 
shapiro.test(transects_summary_new$urchin_density) #data is normal

#homogeneity of variances, if less than 0.05 variances are significantly different than each other
leveneTest(urchin_density ~ site, data = transects_summary_new) #NOT significantly different from each other

#since conditions for a one-way ANOVA are met 
#run ANOVA to compare urchin densities across sites
anova_result <- aov(urchin_density ~ site, data = transects_summary_new)
summary(anova_result) #significant differences in urchin densities across sites 

#post-hoc pairwise comparisons
tukey_result <- TukeyHSD(anova_result)
tukey_result

#significant differences between 
  #taylor and all sites 
  #ellis S and Nanat
  #ellis S and Ed King/Dixon

#set the desired order for the sites
transects_summary_new$site <- factor(transects_summary_new$site, 
                                 levels = c("Taylor", "Ed King", "Dixon", "Ellis N", "Ellis S", "Nanat"))
#visualize via boxplot 
ggplot(transects_summary_new, aes(x = site, y = urchin_density)) +
  geom_boxplot() +
  labs(
    title = "Urchin Density Across Sites",
    x = "Site",
    y = "Urchin Density (m^2)"
  ) +
  theme_minimal()

```

#Do urchin densities change across months? 
```{r}

shapiro.test(transects_summary$urchin_density)  # Normality
bartlett.test(urchin_density ~ month, data = transects_summary)  # Homogeneity

#run one-way ANOVA
anova_result <- aov(urchin_density ~ month, data = transects_summary)
summary(anova_result) #p = 0.596, not significantly different 

#no statistically significant differences in urchin densities between months

#visualize 
plot <- ggplot(transects_summary, aes(x = factor(month), y = urchin_density)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.6) +
  theme_minimal() +
  labs(x = "Month", y = "Urchin Density")

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/densities_by_month.png", plot = plot, width = 10, height = 8, dpi = 300)

#determine mean urchin densities 
# Calculate mean urchin density per site
mean_density_per_site <- transects_summary %>%
  group_by(site) %>%
  summarize(
    mean_urchin_density = mean(urchin_density, na.rm = TRUE),
    .groups = "drop"
  )

#plot 
#reorder months in chronological order
transects_summary$month <- factor(transects_summary$month, levels = c("March", "April", "May", "June"))

#calculate mean urchin density per site
mean_urchin_density <- transects_summary %>% 
  group_by(site) %>% 
  summarise(mean_density = mean(urchin_density, na.rm = TRUE))

#calculate overall mean urchin density across all sites and months
overall_mean_density <- mean(transects_summary$urchin_density, na.rm = TRUE)
overall_mean_density #1.42

#create the plot
plot <- ggplot(transects_summary, aes(x = month, y = urchin_density, color = site, group = site)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  # Plot site mean values as points, change x to "mean"
  geom_point(data = mean_urchin_density, 
             aes(x = "mean", y = mean_density, color = site), 
             shape = 17, size = 4, show.legend = FALSE) +  # Hide legend for mean points
  # Add labels for site mean values with better positioning to avoid overlap
  geom_text_repel(data = mean_urchin_density, 
                  aes(x = "mean", y = mean_density, label = round(mean_density, 2), color = site), 
                  vjust = -1, hjust = -0.2, size = 3, show.legend = FALSE, 
                  box.padding = 0.5, max.overlaps = 10) +  # Prevent text overlap
  # Add the overall mean line and text
  geom_hline(aes(yintercept = overall_mean_density), linetype = "dashed", color = "black") +  # Add overall mean line
  # Add overall mean text with better positioning
  geom_text(aes(x = 1, y = overall_mean_density, label = paste("Overall Mean:", round(overall_mean_density, 2))),
            color = "black", vjust = +2.5, hjust = -0.1, size = 4) +  # Adjust position for the text label
  labs(
    title = "Urchin Density by Month Across Sites",
    x = "Month",
    y = "Urchin Density (m^2)"
  ) +
  scale_x_discrete(labels = c("March", "April", "May", "June", "mean")) +  # Add "mean" to x-axis labels
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/densities_month_site.png", plot = plot, width = 10, height = 8, dpi = 300)
```
#Visualizing urchin counts over transects 
```{r}
transects$n_ur <- transects$n_ru + transects$n_gu + transects$n_pu

#calculate mean and SE of urchin density for each distance, site, and month
agg_data <- transects %>%
  group_by(site, month, distance_m) %>%
  summarise(
    mean_urchin_density = mean(n_ur, na.rm = TRUE),
    se_urchin_density = sd(n_ur, na.rm = TRUE) / sqrt(n())
  )

agg_data <- agg_data %>%
  filter(!month %in% c("August"))

agg_data$se_urchin_density[is.na(agg_data$se_urchin_density)] <- 0

#create a list to store plots for each month
plot_list <- list()

#create the plots for each month and store them in the list
for (m in unique(agg_data$month)) {
  plot <- ggplot(subset(agg_data, month == m), aes(x = distance_m, y = mean_urchin_density)) +
    geom_bar(stat = "identity", position = "dodge", width = 0.8, fill = "skyblue") +  # Bars for mean density
    geom_errorbar(
      aes(ymin = mean_urchin_density - se_urchin_density, 
          ymax = mean_urchin_density + se_urchin_density),
      width = 0.25, position = position_dodge(0.8)  # Error bars for SE
    ) +
    facet_wrap(~ site, scales = "fixed", ncol = 1) +  # Fixed y-axis across all sites
    labs(
      x = "Distance (m)",  # Label for the x-axis
      y = "Mean urchin density",  # Label for the y-axis
      title = paste("Mean urchin densities - Month", m)  # Title for the month
    ) +
    theme_minimal() +  # Clean theme
    theme(
      strip.text = element_text(size = 12),  # Adjust font size for site labels
      axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
    ) +
    # Ensure y-axis starts at 0 for all sites
    expand_limits(y = 0)  # Explicitly set minimum y-axis value to 0
  # Add the plot to the list
  plot_list[[as.character(m)]] <- plot
}

#combine the plots vertically
combined_plot <- wrap_plots(plot_list, ncol = 1)

# Print the combined plot
print(combined_plot)

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/summary_transect_counts.png", 
       plot = combined_plot, 
       width = 10, height = 20, dpi = 300)


```

#Were cages effective at reducing urchins?
#Did urchin densities vary significantly between caged and open plots?
```{r}
#Were cages effective at reducing grazers? 
#Did urchin densities vary significantly between caged and uncaged plots, irrelavant of site or month 

#check distribution of urchin density data - histogram
ggplot(plots, aes(x = urchin_density)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~treatment) +
  labs(title = "Distribution of urchin density by treatment", x = "Urchin Density", y = "Count")

#shapiro-Wilk test for normality, p less than 0.05, normally distributed
shapiro.test(plots$urchin_density[plots$treatment == "caged"]) #normally distributed 
shapiro.test(plots$urchin_density[plots$treatment == "open"]) #normally distributed 

#t-test to compare urchin densities between treatments
t_test_result  <- t.test(urchin_density ~ treatment, data = plots) #p = 0.0001039
t_test_result

p_value <- formatC(t_test_result$p.value, format = "e", digits = 2)

#make box plot to visualize 
ggplot(plots, aes(x = treatment, y = urchin_density, fill = treatment)) +
  geom_boxplot() +
  labs(
    title = "Urchin Density by Treatment",
    x = "Treatment",
    y = "Urchin Density (m²)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("caged" = "lightblue", "open" = "lightgreen")) +
  theme(legend.position = "none") +
  annotate(
    "text",
    x = 1.5, y = max(plots$urchin_density) + 0.5, # Adjust y position if needed
    label = paste("p =", p_value),
    size = 5,
    hjust = 0.5
  )

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/box_treatment.png", 
       plot = combined_plot, 
       width = 10, height = 20, dpi = 300)
```

#CODE GRAVEYARD ##############################################################

#Generalised linear mixed model (GLMM) (only densities - no survival/growth)
#To determine factors affecting urchin densities 

```{r}

#create env summary, for bottom temperature, do, ph, salinity
bottom_data <- env %>%
 group_by(site, month) %>%
  summarize(
    bottom_temp = temp[which.max(depth_m)],
    bottom_do = do[which.max(depth_m)],
    bottom_salinity = salinity[which.max(depth_m)],
    bottom_ph = ph[which.max(depth_m)],
    .groups = "drop" # Ungroups after summarization
  )

#make month a character
bottom_data <- bottom_data %>%
  mutate(month = as.character(month))

#merge environmental data and depth data by site/month to plot data 
plots_glm <- plots %>%
  left_join(hobo, by = c("site", "month")) %>%
  left_join(mean_depth_summary, by = c("site")) %>%
  left_join(bottom_data, by = c("site", "month"))


#log transform urchin density data (need to add a small constant because of zero values)
plots_glm$log_urchin_density <- log(plots_glm$urchin_density + 1)


#run a generalized linear mixed effect model 

#inspecting urchin distributions (to verify assumptions of fitting a GLMM)
hist(plots_glm$log_urchin_density, main = "Histogram of (log) Urchin Density", xlab = "Urchin Density")

#shapiro-Wilk test for normality, p less than 0.05, normally distributed
shapiro.test(plots_glm$log_urchin_density) #normally distributed 

mean_density <- mean(plots_glm$log_urchin_density, na.rm = TRUE)
var_density <- var(plots_glm$log_urchin_density, na.rm = TRUE)
overdispersion <- var_density / mean_density #0.653 (oof)
print(overdispersion) # Should ideally be near 1 for Poisson, 

#check for over zero inflation 
sum(plots_glm$log_urchin_density == 0, na.rm = TRUE) / nrow(plots) #a lot of zeros (40%), use Zero-Inflated Guassian

plots_glm$month <- as.factor(plots_glm$month)
plots_glm$month <- as.numeric(plots_glm$month)
#correlation matrix of environmental variables 
cor_matrix <- cor(plots_glm[, c("mean_light_lux", "mean_temperature_c", 'bottom_salinity', 'bottom_ph', 'mean_depth_m', 'bottom_temp', 'bottom_do', 'month')], use = "complete.obs")
print(cor_matrix)

#visualize with a heatmap 
corrplot(cor_matrix, method = "circle")

#VIF factor - VIF quantifies how much the variance of a predictor is inflated due to multicollinearity. 
  #High VIF values (>5 or 10) indicate problematic collinearity. 
lm_model <- lm(log_urchin_density ~ bottom_temp + bottom_do + bottom_ph + bottom_salinity + mean_temperature_c + mean_light_lux, data = plots_glm) 

vif(lm_model) 

lm_model <- lm(log_urchin_density ~   mean_temperature_c + mean_light_lux, data = plots_glm)

vif(lm_model) #only HOBO measurements

lm_model <- lm(log_urchin_density ~   mean_temperature_c + mean_light_lux + bottom_do,  data = plots_glm)

vif(lm_model) 

lm_model <- lm(log_urchin_density ~   mean_temperature_c + mean_light_lux + bottom_do + bottom_salinity,  data = plots_glm)

vif(lm_model) #salinity = 5.75

lm_model <- lm(log_urchin_density ~   mean_temperature_c + mean_light_lux + bottom_do + month,  data = plots_glm)

vif(lm_model) 

#check for NA 
selected_plots <- plots_glm %>%
  dplyr::select(log_urchin_density, site, mean_temperature_c, treatment, month, mean_light_lux, bottom_do)

summary(selected_plots) #No NA values

#model! 
#standardizing predictors
plots_glm$mean_temperature_c <- scale(plots_glm$mean_temperature_c)
plots_glm$bottom_do <- scale(plots_glm$bottom_do)
plots_glm$mean_light_lux <- scale(plots_glm$mean_light_lux)
plots_glm$mean_depth_m <- scale(plots_glm$mean_depth_m)
plots_glm$month <- scale(plots_glm$month)

#fit the linear mixed model
model <- lmer(log_urchin_density ~ mean_temperature_c + bottom_do + mean_light_lux + month + 
              treatment + mean_depth_m + month + (1 | site),
              data = plots_glm)
summary(model)

model
AIC(model)

#visualize
# Create a dataframe for plotting
plot_data <- data.frame(Fitted = fitted(model),
                        Residuals = resid(model))

# Plot residuals vs. fitted values
plot <- ggplot(plot_data, aes(x = Fitted, y = Residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/residualvfitted.png", 
       plot = combined_plot, 
       width = 10, height = 20, dpi = 300)

#QQ plot
# Extract residuals from the model
residuals <- residuals(model)

# Create a QQ plot
plot <- qqnorm(residuals, main = "QQ Plot of Residuals")
plot <- qqline(residuals, col = "red", lwd = 2)

plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/residualvfitted.png", 
       plot = combined_plot, 
       width = 10, height = 20, dpi = 300)


```
#Testing simpler models + comparing AIC
```{r}
#testing AIC of simpler models - including only one ENV predictor = always lower AIC, including treatment = always lowers AIC
model_1 <- lmer(log_urchin_density ~ mean_temperature_c  +treatment + (1 | site) + (1 | month),
              data = plots_glm)
summary(model_1)

model_1
AIC(model_1) #253

model_2 <- lmer(log_urchin_density ~ mean_temperature_c  + bottom_do + treatment + (1 | month) + (1 | site),
              data = plots_glm)
summary(model_2)

model_2
AIC(model_2) #258

model_3 <- lmer(log_urchin_density ~ bottom_do + treatment +  (1 | month) + (1 | site),
              data = plots_glm)
summary(model_3)

model_3
AIC(model_3) #252

model_4 <- lmer(log_urchin_density ~ mean_light_lux + treatment + (1 | month) + (1 | site),
              data = plots_glm)
summary(model_4)

model_4
AIC(model_4) #250

model_5 <- lmer(log_urchin_density ~ mean_light_lux + mean_temperature_c + treatment + (1 | month) + (1 | site),
              data = plots_glm)
summary(model_5)

model_5
AIC(model_5) #257

model_6 <- lmer(log_urchin_density ~ mean_light_lux + bottom_do +  treatment +(1 | month) + (1 | site),
              data = plots_glm)
summary(model_6)

model_6
AIC(model_6) #256

###
temp <- lmer(log_urchin_density ~ mean_temperature_c +  treatment +(1 | month) + (1 | site),
              data = plots_glm)
summary(temp)

temp
AIC(temp) #253.11


do <- lmer(log_urchin_density ~ bottom_do +  treatment +(1 | month) + (1 | site),
              data = plots_glm)
summary(do)

do
AIC(do) #252.938

sali <- lmer(log_urchin_density ~ bottom_salinity +  treatment +(1 | month) + (1 | site),
              data = plots_glm)
summary(sali)

sali
AIC(sali) #255

light <- lmer(log_urchin_density ~ mean_light_lux +  treatment +(1 | month) + (1 | site),
              data = plots_glm)
summary(light)

light
AIC(light) #250

```


#addressing heteroscedastic linear model - compare to other model types
#Generalized Least Squares 
```{r}

library(nlme)
# Fit a GLS model with a variance structure
gls_model <- gls(log_urchin_density ~ mean_temperature_c + bottom_do + 
                 mean_light_lux + month + treatment + mean_depth_m, 
                 data = plots, 
                 weights = varIdent(form = ~ 1 | treatment))


# Summary of GLS model
summary(gls_model)

# Plot residuals to assess heteroscedasticity
plot(gls_model, which = 1)


#compare to previous linear model
AIC(model)  # for mixed-effects model
AIC(gls_model)  # for GLS model
BIC(model)
BIC(gls_model)

```
#Weighted Least Squares 
```{r}
# Fit weighted linear model using weights based on fitted values
model_weighted <- lm(log_urchin_density ~ mean_temperature_c + bottom_do + 
                     mean_light_lux + month + treatment + mean_depth_m, 
                     data = plots, 
                     weights = 1 / fitted(lm_model)^2)

# Summary of the weighted model
summary(model_weighted)

plot(model_weighted)

# Compare AIC/BIC with the original model
AIC(model)  # for mixed-effects model
AIC(gls_model)  # for GLS model
AIC(model_weighted)
BIC(model)
BIC(gls_model)
BIC(model_weighted)


```


