---
title: "Density Models"
author: "Tessa Rehill"
date: "2024-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)
library(MASS)

```

#Run GLMM (Tweedie distribution, density)

  #load data
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")

```

  #format response variables 
```{r}

typeof(data$urchin_density) #double
typeof(data$site_density) #double 

```


#URCHIN PLOT DENSITY
  #considering overdispersion and zero inflation 
```{r}
#check for overdispersion 
mean_density <- mean(data$urchin_density, na.rm = TRUE)
var_density <- var(data$urchin_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #2.35, high dispersion

#check for zero inflation 
sum(data$urchin_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (39%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

```

  #checking correlation of numerical predictors 
```{r}
data$month <- as.factor(data$month)
data$month <- as.numeric(data$month)

cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "bottom_ph", "percent_occuring_algae", "shannon_div", "simp_div", "species_richness", "month")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot


#compute VIF factor 
#with all predictors 
lm_model <- lm(urchin_density ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m  + bottom_do + species_richness + month + 
               shannon_div + simp_div, data = data)

vif(lm_model) #so many >5



```

  #scaling predictors 
```{r}

#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$simp_div <- scale(data$simp_div)
data$shannon_div <- scale(data$shannon_div)
data$species_richness <- scale(data$species_richness)
data$percent_occuring_algae <- scale(data$percent_occuring_algae)

#turn into factors 
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)

#turn into numeric
data$treatment <- as.numeric(data$treatment)
data$month <- as.numeric(data$month)
data$site <- as.numeric(data$site)

#scale 
data$treatment <- scale(data$treatment)
data$month <- scale(data$month)
data$site <- scale(data$site)

```

 
#Stepwise selection based on AIC and (considering collinearity)
  #only one diversity indices selected (based on best AIC)
```{r}


#starting with a minimal model (only random effects)
base_model <- glmmTMB(urchin_density ~ 1  + (1 | plot_id) +(1 | site)  +(1 | month) , 
                      family = tweedie(link = "log"), 
                      data = data)

#predictors 
predictor_list <- c("percent_occuring_algae", "treatment",
                    "gravel_size", "mean_temperature_c", "mean_light_lux", "bottom_ph",
                    "bottom_do", "bottom_salinity", "mean_depth_m", 
                    "simp_div", "species_richness", "shannon_div")

#initialize variables to store the results
selected_predictors <- c()  # To keep track of selected predictors
current_model <- base_model  # Start with the base model
current_aic <- AIC(base_model)  # AIC of the base model
aic_results <- data.frame(Model = "Base Model", AIC = current_aic)

#forward feature selection loop
for (predictor in predictor_list) {
  # Try adding each predictor to the current model
  formula <- as.formula(
    paste("urchin_density ~", paste(c(selected_predictors, predictor), collapse = " + "), 
          "+ (1 | plot_id) + (1 | site) + (1 | month)")
  )
  
  #fit the model with the new predictor
  new_model <- glmmTMB(
    formula,
    family = tweedie(),
    data = data
  )
  
  #compute AIC
  new_aic <- AIC(new_model)
  
  #store results
  aic_results <- rbind(aic_results, data.frame(Model = paste("Add", predictor), AIC = new_aic))
  
  #check if the AIC improves
  if (new_aic < current_aic) {
    # Update the current model and AIC
    current_model <- new_model
    current_aic <- new_aic
    selected_predictors <- c(selected_predictors, predictor)  # Add predictor to the list
  }
}

#print results
print(aic_results)

current_model 

#final selected model
summary(current_model)

#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```

#Final model 
```{r}
current_model
```

#Visualize 
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
#co-efficient plot 
#create coefficient plot 
```{r}
#extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

#combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

#add the 'model' column
coef_tidy$model <- "Model 1"

#filter out 'cond.sd(Intercept)' from the data
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")

#create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(
    title = "Plot-Level- Urchin Density GLMM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(legend.position = "none") + # Remove the legend
  scale_color_manual(values = "#076fa6")   # Set line color


#print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)

```
#SITE LEVEL DENSITY ##############################################################
#Now look at site-level densities (from transect data)
#NOT a GLMM - just a regular linear model  (no nested, site variability)

```{r}

#select predictors at site level
data <- data %>%
  dplyr::select(month, site, bottom_do, site_density, 
                mean_temperature_c, mean_light_lux, bottom_salinity, 
                bottom_ph, mean_depth_m, species_richness, 
                shannon_div, simp_div)

data <- distinct(data)

#remove `.V1` suffix from column names
colnames(data) <- gsub("\\.V1$", "", colnames(data))

```


```{r}
#check for overdispersion 
mean_density <- mean(data$site_density, na.rm = TRUE)
var_density <- var(data$site_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #0.58 

#check for zero inflation 
sum(data$site_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (19%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

```

  #checking correlation of predictors 
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "bottom_ph", "shannon_div", "simp_div", "species_richness")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#compute VIF factor 
#with all predictors 
lm_model <- lm(site_density ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m  + bottom_do + species_richness + 
               shannon_div + simp_div, data = data)

vif(lm_model) #so many >5



```
 
  #scaling predictors 
```{r}

#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$simp_div <- scale(data$simp_div)
data$shannon_div <- scale(data$shannon_div)
data$species_richness <- scale(data$species_richness)

#turn into factors 
data$month <- factor(data$month)
#turn into numeric
data$month <- as.numeric(data$month)

#scale 
data$month <- scale(data$month)

```


  #fit GLM using stepwise AIC selection 
```{r}

glm_model <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                        bottom_salinity + mean_depth_m + bottom_ph + month,
                      data = data, family = gaussian())

stepwise <- step(glm_model, direction = "both", trace = 1)


#view the best model
summary(stepwise)

stepwise_model <- stepwise #based on lowest AIC

```
  #stepwise model 
```{r}
stepwise_model
summary(stepwise_model)
```
  #check colinearity 
```{r}
#Compute VIF factor 
#with all predictors 
lm_model <- lm(site_density ~ mean_light_lux + bottom_salinity + bottom_ph,
                data = data)

vif(lm_model) #all less than 5

```
  
 
  #visualize
```{r}
#simple model 

fittedModel <- stepwise_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
 
 #Coeffiencts 
```{r}
# Extract coefficients in tidy format, including confidence intervals
coef_tidy <- broom.mixed::tidy(stepwise_model, conf.int = TRUE)

# Combine 'term' with additional info (if necessary) and remove 'component'
# Here, we'll only use the 'term' column, assuming 'term' provides all the information you need
coef_tidy <- coef_tidy %>%
  mutate(term = term)  # If you want a custom naming format, modify this step accordingly

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")

# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +  # Apply a minimal theme
  labs(
    title = "Site-Level: Urchin GLM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(
    legend.position = "none",  # Remove the legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(values = "#076fa6")  # Set line color

print(dw)
#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)

```
 #GRAVEYARD###############################
 
```{r}

#Stepwise selection based on AIC and (considering collinearity)
  #only one diversity indices selected (based on best AIC)


#base model
base_model <- glmmTMB(urchin_density ~ 1 + (1 | plot_id),
                      family = tweedie(),
                      data = data)

#predictor groups (collinear variables grouped together)
predictor_groups <- list(
  group1 = c("shannon_div", "ssimp_div", "sspecies_richness"), 
  group2 = c("bottom_salinity", "bottom_do", "bottom_ph"),    # Bottom water properties
  group3 = c("mean_light_lux", "mean_temperature_c", "mean_depth_m"), # Environmental variables
  group4 = c("treatment", "percent_occuring_algae", "month", "site")  # Less correlated predictors
)

#initialize variables
selected_predictors <- c()
selected_diversity_index <- NULL  # Track chosen diversity index
current_model <- base_model
current_aic <- AIC(base_model)
aic_results <- data.frame(Model = "Base Model", AIC = current_aic)

#forward selection loop
for (group in predictor_groups) {
  for (predictor in group) {
    # Skip predictors already in the model
    if (predictor %in% selected_predictors) next
    
    # Ensure only one diversity index is selected
    if (predictor %in% predictor_groups$group1) {
      if (!is.null(selected_diversity_index)) next  # Skip if diversity index already chosen
    }
    
    # Check for collinearity with selected predictors
    is_collinear <- FALSE
    for (existing in selected_predictors) {
      correlation <- cor(data[[predictor]], data[[existing]], use = "pairwise.complete.obs")
      if (abs(correlation) > 0.7) {  # Collinearity threshold
        is_collinear <- TRUE
        break
      }
    }
    if (is_collinear) next  # Skip if collinear
    
    # Create model formula
    formula <- as.formula(
      paste("urchin_density ~", paste(c(selected_predictors, predictor), collapse = " + "),
            "+ (1 | plot_id)")
    )
    
    # Fit the model
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    # Update if AIC improves
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      aic_results <- rbind(aic_results, data.frame(Model = paste("Add", predictor), AIC = new_aic))
      
      # If the selected predictor is a diversity index, track it
      if (predictor %in% predictor_groups$group1) {
        selected_diversity_index <- predictor
      }
    }
  }
}

# Final model summary
print(aic_results)
summary(current_model)
```

 
 
#Stepwise selection based on AIC and (considering collinearity)
  #only one diversity indices selected (based on best AIC)
```{r}


#starting with a minimal model (only random effects)
base_model <- glmmTMB(urchin_density ~ 1  + (1 | plot_id) +(1 | site)  +(1 | month) , 
                      family = tweedie(link = "log"), 
                      data = data)

#predictors 
predictor_list <- c("percent_occuring_algae", "treatment",
                    "gravel_size", "mean_temperature_c", "mean_light_lux", "bottom_ph",
                    "bottom_do", "bottom_salinity", "mean_depth_m", 
                    "simp_div", "species_richness", "shannon_div")

#initialize variables to store the results
selected_predictors <- c()  # To keep track of selected predictors
current_model <- base_model  # Start with the base model
current_aic <- AIC(base_model)  # AIC of the base model
aic_results <- data.frame(Model = "Base Model", AIC = current_aic)

#forward feature selection loop
for (predictor in predictor_list) {
  # Try adding each predictor to the current model
  formula <- as.formula(
    paste("urchin_density ~", paste(c(selected_predictors, predictor), collapse = " + "), 
          "+ (1 | plot_id) + (1 | site) + (1 | month)")
  )
  
  #fit the model with the new predictor
  new_model <- glmmTMB(
    formula,
    family = tweedie(),
    data = data
  )
  
  #compute AIC
  new_aic <- AIC(new_model)
  
  #store results
  aic_results <- rbind(aic_results, data.frame(Model = paste("Add", predictor), AIC = new_aic))
  
  #check if the AIC improves
  if (new_aic < current_aic) {
    # Update the current model and AIC
    current_model <- new_model
    current_aic <- new_aic
    selected_predictors <- c(selected_predictors, predictor)  # Add predictor to the list
  }
}

#print results
print(aic_results)

current_model 

#final selected model
summary(current_model)

#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```
 
```{r}
#ft three separate models for each diversity index
glm_model_richness <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                            bottom_salinity + mean_depth_m + species_richness + 
                            month, 
                          data = data, family = gaussian())

glm_model_shannon <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                           bottom_salinity + mean_depth_m + shannon_div + 
                           month, 
                         data = data, family = gaussian())

glm_model_simp <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                        bottom_salinity + mean_depth_m + simp_div + 
                        month, 
                      data = data, family = gaussian())

#run stepwise selection on each model
stepwise_richness <- step(glm_model_richness, direction = "both", trace = 1)
stepwise_shannon <- step(glm_model_shannon, direction = "both", trace = 1)
stepwise_simp <- step(glm_model_simp, direction = "both", trace = 1)
```
 
Best model for urchin that fits 
```{r}
site_density ~ mean_temperature_c + mean_light_lux + 
    bottom_salinity + mean_depth_m, family = gaussian())

```

 
 