---
title: "Density Models"
author: "Tessa Rehill"
date: "2024-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)
library(MASS)

```

#Run GLMM (Tweedie distribution, density)

  #load data
```{r}
#transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_monitoring_transects.csv")

#plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```

  #format response variables 
```{r}

typeof(data$urchin_density) #double
typeof(data$site_density) #double 

```


#URCHIN PLOT DENSITY
  #considering overdispersion and zero inflation 
```{r}
#check for overdispersion 
mean_density <- mean(data$urchin_density, na.rm = TRUE)
var_density <- var(data$urchin_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #2.35, high dispersion

#check for zero inflation 
sum(data$urchin_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (39%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

```

  #checking correlation of numerical predictors 
```{r}
data$month <- as.factor(data$month)
data$month <- as.numeric(data$month)

cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "bottom_ph", "percent_occuring_algae", "shannon_div", "simp_div", "species_richness", "month")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/env/correlation_plot.png", plot = plot, width = 10, height = 8, dpi = 300)

#Compute VIF factor 
#with all predictors 
lm_model <- lm(urchin_density ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m  + bottom_do + species_richness + month + 
               shannon_div + simp_div, data = data)

vif(lm_model) #so many >5



```

  #scaling predictors 
```{r}

#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$simp_div <- scale(data$simp_div)
data$shannon_div <- scale(data$shannon_div)
data$species_richness <- scale(data$species_richness)
data$percent_occuring_algae <- scale(data$percent_occuring_algae)

#turn into factors 
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)

#turn into numeric
data$treatment <- as.numeric(data$treatment)
data$month <- as.numeric(data$month)
data$site <- as.numeric(data$site)

#scale 
data$treatment <- scale(data$treatment)
data$month <- scale(data$month)
data$site <- scale(data$site)


```

  #simplest model
```{r}
#fit the simplest GLMM model
simple_model <- glmmTMB(
  urchin_density ~ 1 +  treatment +  (1 |plot_id) +  (1 |site),
  data = data,
  family = tweedie()
)

summary(simple_model)
AIC(simple_model) #619.2


```
  
  #visualize
```{r}
#simple model 
fittedModel <- simple_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```

#Stepwise selection based on AIC and (considering collinearity)
  #only one diversity indices selected (based on best AIC)
```{r}

#base model
base_model <- glmmTMB(urchin_density ~ 1 + (1 | plot_id),
                      family = tweedie(),
                      data = data)

#predictor groups (collinear variables grouped together)
predictor_groups <- list(
  group1 = c("shannon_div", "species_richness", "simp_div"),  # Diversity indices
  group2 = c("bottom_salinity", "bottom_do", "bottom_ph"),    # Bottom water properties
  group3 = c("mean_light_lux", "mean_temperature_c", "mean_depth_m"), # Environmental variables
  group4 = c("treatment", "percent_occuring_algae","month", "site")  # Less correlated predictors
)

#initialize variables
selected_predictors <- c()
selected_diversity_index <- NULL  # Track chosen diversity index
current_model <- base_model
current_aic <- AIC(base_model)
aic_results <- data.frame(Model = "Base Model", AIC = current_aic)

# Forward selection loop
for (group in predictor_groups) {
  for (predictor in group) {
    # Skip predictors already in the model
    if (predictor %in% selected_predictors) next
    
    # Ensure only one diversity index is selected
    if (predictor %in% predictor_groups$group1) {
      if (!is.null(selected_diversity_index)) next  # Skip if diversity index already chosen
    }
    
    # Check for collinearity with selected predictors
    is_collinear <- FALSE
    for (existing in selected_predictors) {
      correlation <- cor(data[[predictor]], data[[existing]], use = "pairwise.complete.obs")
      if (abs(correlation) > 0.7) {  # Collinearity threshold
        is_collinear <- TRUE
        break
      }
    }
    if (is_collinear) next  # Skip if collinear
    
    # Create model formula
    formula <- as.formula(
      paste("urchin_density ~", paste(c(selected_predictors, predictor), collapse = " + "),
            "+ (1 | plot_id)")
    )
    
    # Fit the model
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    # Update if AIC improves
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      aic_results <- rbind(aic_results, data.frame(Model = paste("Add", predictor), AIC = new_aic))
      
      # If the selected predictor is a diversity index, track it
      if (predictor %in% predictor_groups$group1) {
        selected_diversity_index <- predictor
      }
    }
  }
}

# Final model summary
print(aic_results)
summary(current_model)

```

#checking collinearity among predictors  
```{r}
lm_model <- lm(urchin_density ~  bottom_ph + bottom_do + mean_depth_m +  percent_occuring_algae +  
               shannon_div + simp_div, data = data)

vif(lm_model) #VIF over 5 can not be included together (exists for diversity indices)
```

#Final model 
```{r}
current_model
```

#Visualize 
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

# Save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

# Save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

# Save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
#co-efficient plot 
#create coefficient plot 
```{r}
# Extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

# Combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) + 
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(title = "Model Coefficients with Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Predictor",
       color = "Component")

# Print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 

# Save the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)

```
#SITE LEVEL DENSITY ##############################################################
#Now look at site-level densities (from transect data)
#NOT a GLMM - just a regular linear model  (no nested, site variability)

```{r}


data <- data %>% 
  select(month, site, bottom_do, site_density, 
         mean_temperature_c, mean_light_lux, bottom_salinity, 
         bottom_ph, mean_depth_m, species_richness, 
         shannon_div, simp_div)

data <- distinct(data)

# Remove `.V1` suffix from column names
colnames(data) <- gsub("\\.V1$", "", colnames(data))



```


```{r}
#check for overdispersion 
mean_density <- mean(data$site_density, na.rm = TRUE)
var_density <- var(data$site_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #0.58 

#check for zero inflation 
sum(data$site_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (19%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

```

  #checking correlation of predictors 
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "bottom_ph", "shannon_div", "simp_div", "species_richness")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#Compute VIF factor 
#with all predictors 
lm_model <- lm(site_density ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m  + bottom_do + species_richness + 
               shannon_div + simp_div, data = data)

vif(lm_model) #so many >5



```
 
  #scaling predictors 
```{r}

#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$simp_div <- scale(data$simp_div)
data$shannon_div <- scale(data$shannon_div)
data$species_richness <- scale(data$species_richness)

#turn into factors 
data$month <- factor(data$month)



#turn into numeric
data$month <- as.numeric(data$month)

#scale 
data$month <- scale(data$month)



```


```{r}

# Fit three separate models for each diversity index
glm_model_richness <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                            bottom_salinity + mean_depth_m + species_richness + 
                            month, 
                          data = data, family = gaussian())

glm_model_shannon <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                           bottom_salinity + mean_depth_m + shannon_div + 
                           month, 
                         data = data, family = gaussian())

glm_model_simp <- glm(site_density ~ bottom_do + mean_temperature_c + mean_light_lux + 
                        bottom_salinity + mean_depth_m + simp_div + 
                        month, 
                      data = data, family = gaussian())

# Run stepwise selection on each model
stepwise_richness <- step(glm_model_richness, direction = "both", trace = 1)
stepwise_shannon <- step(glm_model_shannon, direction = "both", trace = 1)
stepwise_simp <- step(glm_model_simp, direction = "both", trace = 1)

# Compare AIC of the models
AIC(stepwise_richness, stepwise_shannon, stepwise_simp)

# View the best model
summary(stepwise_shannon)  # or whichever model is best based on AIC

stepwise_model <- stepwise_shannon #based on lowest AIC

```
  #stepwise model 
```{r}
stepwise_model
```

 
  #visualize
```{r}
#simple model 
fittedModel <- stepwise_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

# Save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

# Save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

# Save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_models/site_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
 
 