---
title: "GLMM"
author: "Tessa Rehill"
date: "2024-12-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)
library(broom)

```

#load libraries 

```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```

#format predictors 
```{r}

#format survival 
#kelp survival as proportion
data$kelp_prop <- data$kelp_survival/100

#transform response variable with small constant (beta regression can not handle zeros)
data$kelp_prop_adj <- (data$kelp_prop * (nrow(data) - 1) + 0.5) / nrow(data)

```

#consider overdispersion and overinflation 
```{r}
#URCHIN MODELS 
#check for overdispersion 
mean_density <- mean(data$urchin_density, na.rm = TRUE)
var_density <- var(data$urchin_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #2.35, high dispersion

#check for zero inflation 
sum(data$urchin_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (39%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

#KELP SURVIVAL 
#check for overdispersion (in kelp proportion adjusted with a constant)
mean_survival <- mean(data$kelp_prop_adj, na.rm = TRUE)
var_survival <- var(data$kelp_prop_adj, na.rm = TRUE)
overdispersion <- var_survival / mean_survival 
print(overdispersion)  #37 - high dispersion

sum(data$kelp_prop_adj == 0, na.rm = TRUE) / nrow(data) #no zeros 

#KELP GROWTH 
#check for overdispersion 
mean_growth <- mean(data$kelp_growth, na.rm = TRUE)
var_growth <- var(data$kelp_growth, na.rm = TRUE)
overdispersion <- var_growth / mean_growth
print(overdispersion)  #661- high dispersion

#check for zero inflation 
sum(data$kelp_growth == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (48%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

#URCHIN SITE DENSITY 
#check for overdispersion 
mean_density <- mean(data$site_density, na.rm = TRUE)
var_density <- var(data$site_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #0.58 

#check for zero inflation 
sum(data$site_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (19%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power


```

#correlation plot (all env variables)
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity",  "bottom_do", "bottom_ph", "urchin_density", "percent_occuring_algae", "site_density",  "simp_div", "species_richness", "grazer_density")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/correlation_plot.png", plot = plot, width = 10, height = 8, dpi = 300)

```

#scale predictors
```{r}
#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$mean_light_lux <- scale(data$mean_light_lux)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_do <- scale(data$bottom_do)
data$bottom_salinity <- scale(data$bottom_salinity)
data$simp_div <- scale(data$simp_div)
data$species_richness <- scale(data$species_richness)
data$percent_occuring_algae <- scale(data$percent_occuring_algae)

#month needs to be turned back into a factor
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)

#turn into numeric + scale
#data$treatment <- as.numeric(data$treatment) #fits urchin density better???
#data$treatment <- scale(data$treatment)

#data$gravel_size <- as.numeric(data$gravel_size) #ruins urchin density fit???
#data$gravel_size <- scale(data$gravel_size)

```

#GLMM URCHIN DENSITY
  #base model + predictors 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(urchin_density ~ 1  +(1 | site), 
                      family = tweedie(link = "log"), 
                      data = data)

#predictors 
predictor_list <- c("percent_occuring_algae", "treatment",
                    "gravel_size", "mean_temperature_c", "mean_light_lux", "bottom_ph", "bottom_do", "bottom_salinity",
                    "simp_div", "species_richness")

```

  
#MODEL ALGORITHM - forward and backward stepwise selection
```{r}

#initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

#stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("urchin_density ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("urchin_density ~", 
            paste(remaining_predictors, collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

#print results
print(selected_predictors)
summary(current_model)

```


  #visualize (+ save)
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T, n=250)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()


#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
  
  #check collinearity 
```{r}
#compute VIF factor 
#with selected predictors 
lm_model <- lm(urchin_density ~ treatment + mean_temperature_c + bottom_ph, data = data)

vif(lm_model) #below 5
```

  #coefficient plot 
```{r}
#extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

#combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

#add the 'model' column
coef_tidy$model <- "Model 1"

#filter out 'cond.sd(Intercept)' from the data
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")

#create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(
    title = "Plot-Level- Urchin Density GLMM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(legend.position = "none") + # Remove the legend
  scale_color_manual(values = "#076fa6")   # Set line color


#print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```

  #final model 
```{r}
urchin_plot <- current_model
```

#GLM URCHIN SITE DENSITY 
```{r}
#select predictors at site level
site_data <- data %>%
  dplyr::select(month, site, bottom_do, site_density, 
                mean_temperature_c, mean_light_lux, bottom_salinity, 
                bottom_ph, mean_depth_m, species_richness, 
                shannon_div, simp_div)

site_data <- distinct(site_data)

#remove `.V1` suffix from column names
colnames(site_data) <- gsub("\\.V1$", "", colnames(site_data))
```

   #fit GLM using stepwise AIC selection
```{r}
glm_model <- glm(site_density ~ bottom_do + bottom_salinity + bottom_ph + mean_temperature_c, 
                      data = site_data, family = gaussian())

#removed light as predictor, ecologically irrelevant 

stepwise <- step(glm_model, direction = "both", trace = 1)

stepwise_model <- stepwise #based on lowest AIC
#view the best model
summary(stepwise_model)

```

  #check colinearity 
```{r}
#Compute VIF factor 
#with all predictors 
lm_model <- lm(site_density ~ mean_temperature_c + bottom_do + bottom_ph,
                data = data)

vif(lm_model) #all less than 5

```

  #visualize 
```{r}
#simple model 

fittedModel <- stepwise_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/site_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/site_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/site_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```

  #coefficients
```{r}
#extract coefficients in tidy format, including confidence intervals
coef_tidy <- broom.mixed::tidy(stepwise_model, conf.int = TRUE)

#combine 'term' with additional info (if necessary) and remove 'component'
#here, we'll only use the 'term' column, assuming 'term' provides all the information you need
coef_tidy <- coef_tidy %>%
  mutate(term = term)  # If you want a custom naming format, modify this step accordingly

#add the 'model' column
coef_tidy$model <- "Model 1"

#filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")

#create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +  # Apply a minimal theme
  labs(
    title = "Site-Level: Urchin GLM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(
    legend.position = "none",  # Remove the legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(values = "#076fa6")  # Set line color

print(dw)

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/site_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)

```

  #final model
```{r}
urchin_site <- stepwise_model
```

#KELP SURVIVORSHIP GLMM
  #scale predictors 
```{r}
data$urchin_density <- scale(data$urchin_density)
data$site_density <- scale(data$site_density)
data$grazer_density <- scale(data$grazer_density)
```
  
  #base model + predictor 
```{r}
#define predictors and initial formula
predictor_list <-  c( "percent_occuring_algae", "treatment", "site_density", "urchin_density",
                "gravel_size",  "mean_temperature_c", "mean_light_lux", 
                   "bottom_ph", "bottom_salinity", "simp_div", "species_richness", "grazer_density")


#initialize the empty model formula (with no predictors)
base_model <- glmmTMB(
  kelp_prop_adj ~ 1 + (1 | site), 
  data = data,
  family = beta_family(link = "logit")
)
```

#MODEL ALGORITHM
```{r}

# Initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

# Stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("kelp_prop_adj ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = beta_family(link="logit"), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("kelp_prop_adj ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = beta_family(link="logit"), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

# Print results
print(selected_predictors)
summary(current_model)

``` 

#visualize (+ save)
```{r}

#final model
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
  
  #collinearity 
```{r}
lm_model <- lm(kelp_prop_adj ~ site_density + species_richness + gravel_size + mean_temperature_c + simp_div + percent_occuring_algae , data = data)

vif(lm_model) #all below 5 VIF


```

  #coefficient plot 
```{r}
#extract coefficients in tidy format, including confidence intervals
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

#combine 'term' with additional info (if necessary) and remove 'component'
#here, we'll only use the 'term' column, assuming 'term' provides all the information you need

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "sd__(Intercept)")

# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +  # Apply a minimal theme
  labs(
    title = "Kelp Survival GLMM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(
    legend.position = "none",  # Remove the legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(values = "#076fa6")  # Set line color

print(dw)

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/surv_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```

  #final_model 
```{r}
kelp_model <- current_model
```

#KELP GROWTH MODEL 
  #base model + predictors 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(kelp_growth ~ 1 + (1 | site)  , 
                      family = tweedie(), 
                      data = data)

#predictors 
predictor_list <- c( "percent_occuring_algae", "treatment", "site_density", "urchin_density",
                "gravel_size",  "mean_temperature_c", "mean_light_lux", 
                   "bottom_ph", "bottom_salinity","grazer_density",  "simp_div")


```

 
  #NEW MODEL ALGORITHM - forward and backward 
```{r}


# Initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

# Stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("kelp_growth ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("kelp_growth ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

# Print results
print(selected_predictors)
summary(current_model)

```
  

  #coefficients 
```{r}
lm_model <- lm(kelp_growth ~  gravel_size + mean_temperature_c + 
                  simp_div + site_density + bottom_ph + grazer_density, data = data)

vif(lm_model) #all below 5

lm_model <- lm(kelp_growth ~  gravel_size + mean_temperature_c + 
                   site_density + bottom_ph + grazer_density, data = data)

vif(lm_model) #all below 5
```

  #visualize + save 
```{r}
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/grw_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/grw_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/grw_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()
```

  #coefficients
```{r}
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

# Combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")



# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) + 
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(title = "Model Coefficients with Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Predictor",
       color = "Component")

# Print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 


#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/grw_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```

  #final model 
```{r}
growth_model <- current_model
```

#PLOT ALL COEFFICIENT FIGURES 
```{r}
# Extract coefficients and confidence intervals for each model
kelp_model_tidy <- tidy(kelp_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Survival GLMM")

urchin_plots_tidy <- tidy(urchin_plot, conf.int = TRUE) %>%
  mutate(model = "Urchin Plot-Level Density GLMM")

growth_model_tidy <- tidy(growth_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Growth GLMM")

# Correctly tidy the GLM (urchin_sites)
urchin_sites_tidy <- broom::tidy(urchin_site, conf.int = TRUE) %>%
  mutate(model = "Urchin Site-Level Density GLM")

# Combine all models into one data frame
all_models <- bind_rows(kelp_model_tidy, urchin_plots_tidy, urchin_sites_tidy, growth_model_tidy)

# Filter out random effects (if you only want fixed effects)
all_models <- all_models %>%
  filter(effect == "fixed" | is.na(effect))  # GLM doesn't have an 'effect' column, so allow NA

# Define the color scheme for each model
model_colors <- c(
  "Kelp Survival GLMM" = "#eacf62",
  "Urchin Plot-Level Density GLMM" = "#62a4ea",
  "Urchin Site-Level Density GLM" = "#076fa6",
  "Kelp Growth GLMM" = "#FF5733"
)

# Create the coefficient plot with custom colors
plot <- ggplot(all_models, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high, color = model)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  coord_flip() +
  scale_color_manual(values = model_colors) +
  labs(
    title = "Coefficient and Confidence Interval Plot",
    x = "Predictors",
    y = "Estimate",
    color = "Model"
  ) +
  theme_minimal()

plot

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/combined_dot_and_whisker_plot.png", plot = plot, width = 8, height = 6, dpi = 300)


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
