---
title: "GLMM"
author: "Tessa Rehill"
date: "2024-12-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)
library(broom)
library(lme4)
library(lmerTest)


```

#load libraries 

```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```


#consider overdispersion and overinflation 
```{r}
#URCHIN MODELS 
#check for overdispersion 
mean_density <- mean(data$urchin_density, na.rm = TRUE)
var_density <- var(data$urchin_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #2.35, high dispersion

#check for zero inflation 
sum(data$urchin_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (39%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

#KELP SURVIVAL 
#check for overdispersion (in kelp proportion adjusted with a constant)
mean_survival <- mean(data$kelp_survival, na.rm = TRUE)
var_survival <- var(data$kelp_survival, na.rm = TRUE)
overdispersion <- var_survival / mean_survival 
print(overdispersion)  #37 - high dispersion

sum(data$kelp_survival == 0, na.rm = TRUE) / nrow(data) #no zeros 

#KELP GROWTH 
#check for overdispersion 
mean_growth <- mean(data$kelp_growth, na.rm = TRUE)
var_growth <- var(data$kelp_growth, na.rm = TRUE)
overdispersion <- var_growth / mean_growth
print(overdispersion)  #661- high dispersion

#check for zero inflation 
sum(data$kelp_growth == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (48%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

#URCHIN SITE DENSITY 
#check for overdispersion 
mean_density <- mean(data$urchin_site_density, na.rm = TRUE)
var_density <- var(data$urchin_site_density, na.rm = TRUE)
overdispersion <- var_density / mean_density
print(overdispersion)  #0.58 

#check for zero inflation 
sum(data$urchin_site_density == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (19%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power


```

#correlation plot (all env variables)
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "percent_occuring_algae", "mean_temperature_c", "max_temperature_c",  "mean_current", "bottom_salinity",  "bottom_do", "bottom_ph", "urchin_density", "urchin_site_density", "grazer_density", "grazers_site_density")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

high_corr_pairs <- which(abs(cor_matrix) > 0.7, arr.ind = TRUE)

#Print the pairs with their correlation values
high_corr_pairs_df <- data.frame(
  Var1 = rownames(cor_matrix)[high_corr_pairs[, 1]],
  Var2 = colnames(cor_matrix)[high_corr_pairs[, 2]],
  Correlation = cor_matrix[high_corr_pairs]
)

#Filter out pairs with correlation equal to 1
high_corr_pairs_df <- high_corr_pairs_df[high_corr_pairs_df$Correlation != 1, ]

# Create a unique pair identifier by sorting the variables
high_corr_pairs_df$pair <- apply(high_corr_pairs_df[, c("Var1", "Var2")], 1, function(x) paste(sort(x), collapse = "_"))

# Remove duplicate pairs
high_corr_pairs_df <- high_corr_pairs_df[!duplicated(high_corr_pairs_df$pair), ]

# Remove the 'pair' column now that we've filtered duplicates
high_corr_pairs_df <- high_corr_pairs_df[, -ncol(high_corr_pairs_df)]

# Print the updated dataframe
print(high_corr_pairs_df)

#remove max_current because associated with a lot of variables 

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/correlation_plot.png", plot = plot, width = 10, height = 8, dpi = 300)

```

#scale predictors
```{r}
#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$mean_light_lux <- scale(data$mean_light_lux)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_do <- scale(data$bottom_do)
data$bottom_salinity <- scale(data$bottom_salinity)


data$percent_occuring_algae <- scale(data$percent_occuring_algae)
data$max_temperature_c <- scale(data$max_temperature_c)
data$mean_current <- scale(data$mean_current)
data$grazers_site_density <- scale(data$grazers_site_density)

#month needs to be turned back into a factor
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)

#turn into numeric + scale
#data$treatment <- as.numeric(data$treatment) #fits urchin density better???
#data$treatment <- scale(data$treatment)

#data$gravel_size <- as.numeric(data$gravel_size) #ruins urchin density fit???
#data$gravel_size <- scale(data$gravel_size)

```

#GLMM URCHIN DENSITY
  #base model + predictors 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(urchin_density ~ 1  + (1 | site), 
                      family = tweedie(link = "log"), 
                      data = data)

#predictors 
predictor_list <- c( "treatment","gravel_size", "mean_temperature_c", "max_temperature_c", "mean_current", "bottom_do", "bottom_salinity")



#ph removed because highly correlated 
#mean light and gravel removed because ecologically irrelevant 

```

  
#MODEL ALGORITHM - forward and backward stepwise selection
```{r}

#initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

#stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("urchin_density ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("urchin_density ~", 
            paste(remaining_predictors, collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

#print results
print(selected_predictors)
summary(current_model)

```


  #visualize (+ save)
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T, n=100, refit = T)
simulationOutput


#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()


#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
  
  #check collinearity 
```{r}
#compute VIF factor 
#with selected predictors 
lm_model <- lm(urchin_density ~ treatment + mean_temperature_c + mean_current + bottom_salinity, data = data)
vif(lm_model)

```

  #coefficient plot 
```{r}
#extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

#combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

#add the 'model' column
coef_tidy$model <- "Model 1"

#filter out 'cond.sd(Intercept)' from the data
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")

#create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(
    title = "Plot-Level- Urchin Density GLMM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(legend.position = "none") + # Remove the legend
  scale_color_manual(values = "#076fa6")   # Set line color


#print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/urchin_plot_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```

  #final model 
```{r}
urchin_plot <- current_model
```


#KELP SURVIVORSHIP GLMM
  #scale predictors 
```{r}
data$urchin_density <- scale(data$urchin_density)
data$urchin_site_density <- scale(data$urchin_site_density)
data$grazer_density <- scale(data$grazer_density)
```
  
  #base model + predictor 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ 1  +(1 | site), 
                      family = binomial, 
                      data = data)

#predictors 
predictor_list <- c("percent_occuring_algae", "treatment", "urchin_site_density", "urchin_density", "grazer_density", "grazers_site_density",
                    "max_temperature_c", "mean_current", "gravel_size", "mean_temperature_c", "mean_light_lux", "bottom_ph", "bottom_do", "bottom_salinity")
```

#MODEL ALGORITHM
```{r}
#initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

#stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~", 
            paste(remaining_predictors, collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

#print results
print(selected_predictors)
summary(current_model)


``` 

#visualize (+ save)
```{r}

#final model
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
  
  #collinearity 
```{r}
lm_model <- lm(n_gravel ~ urchin_site_density + gravel_size + max_temperature_c + mean_current + grazer_density, data = data)

vif(lm_model) #all below 5

lm_model <- lm(n_kelp_gravel ~ urchin_site_density + gravel_size + max_temperature_c + mean_current + grazer_density, data = data)

vif(lm_model) #all below 5

lm_model <- lm(kelp_survival ~ urchin_site_density + gravel_size + max_temperature_c + mean_current + grazer_density, data = data)

vif(lm_model) #all below 5



```

  #coefficient plot 
```{r}
#extract coefficients in tidy format, including confidence intervals
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

#combine 'term' with additional info (if necessary) and remove 'component'
#here, we'll only use the 'term' column, assuming 'term' provides all the information you need

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "sd__(Intercept)")

# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +  # Apply a minimal theme
  labs(
    title = "Kelp Survival GLMM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(
    legend.position = "none",  # Remove the legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(values = "#076fa6")  # Set line color

print(dw)

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/surv_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```

  #final_model 
```{r}
kelp_model <- current_model
```

#KELP GROWTH MODEL 
  #base model + predictors 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(kelp_growth ~ 1 + (1 | site)  , 
                      family = tweedie(), 
                      data = data)

#predictors 
predictor_list <- c("percent_occuring_algae", "treatment", "urchin_site_density", "urchin_density", "grazer_density", "grazers_site_density",
                    "max_temperature_c", "mean_current", "gravel_size", "mean_temperature_c", "mean_light_lux", "bottom_do", "bottom_salinity")

#remove ph due to high collinearity (and low ecological importance and big data gaps)

```

 
  #NEW MODEL ALGORITHM - forward and backward 
```{r}
# Initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

# Stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("kelp_growth ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("kelp_growth ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = tweedie(), data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

# Print results
print(selected_predictors)
summary(current_model)

```
#check collinearity
```{r}
lm_model <- lm(kelp_growth ~  max_temperature_c + 
                  mean_current + urchin_site_density + percent_occuring_algae + grazer_density, data = data)

vif(lm_model) #all below 5

```

  #visualize 
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput


#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()


#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```

  #coefficients
```{r}

coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

# Combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")



# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) + 
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(title = "Model Coefficients with Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Predictor",
       color = "Component")

# Print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 


#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/grw_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```

  #final model 
```{r}
growth_model <- current_model
```

#PLOT ALL COEFFICIENT FIGURES 
```{r}
# Extract coefficients and confidence intervals for each model
kelp_model_tidy <- tidy(kelp_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Survival GLMM")

urchin_plots_tidy <- tidy(urchin_plot, conf.int = TRUE) %>%
  mutate(model = "Urchin Plot-Level Density GLMM")

growth_model_tidy <- tidy(growth_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Growth GLMM")



# Combine all models into one data frame
all_models <- bind_rows(kelp_model_tidy, urchin_plots_tidy, growth_model_tidy)

# Filter out random effects (if you only want fixed effects)
all_models <- all_models %>%
  filter(effect == "fixed" | is.na(effect))  # GLM doesn't have an 'effect' column, so allow NA

# Define the color scheme for each model
model_colors <- c(
  "Kelp Survival GLMM" = "#eacf62",
  "Urchin Plot-Level Density GLMM" = "#62a4ea",
  "Kelp Growth GLMM" = "#FF5733"
)

# Define the desired order of predictors in reverse
predictor_order <- rev(c('treatmentopen', 'gravel_sizesmall', 'urchin_site_density',  
                         'grazer_density', 'mean_current', 'mean_temperature_c',  
                         'max_temperature_c', 'bottom_salinity', 'percent_occuring_algae', '(Intercept)'))

# Modify the ggplot code to reorder the terms and increase font sizes
plot <- ggplot(all_models, aes(x = factor(term, levels = predictor_order),  
                               y = estimate, ymin = conf.low, ymax = conf.high,  
                               color = model)) +    
  geom_pointrange(position = position_dodge(width = 0.5)) +    
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +    
  coord_flip() +    
  scale_color_manual(values = model_colors) +    
  labs(
    title = "Coefficient and Confidence Interval Plot",
    x = "Predictors",
    y = "Estimate",
    color = "Model"
  ) +    
  theme_minimal() +
  theme(
    text = element_text(size = 18),         # Base text size
    axis.text = element_text(size = 16),    # Axis tick labels
    axis.title = element_text(size = 18),   # Axis titles
    legend.text = element_text(size = 16),  # Legend text
    legend.title = element_text(size = 18), # Legend title
    plot.title = element_text(size = 20, face = "bold") # Plot title
  )

plot

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/combined_dot_and_whisker_plot.png", 
       plot = plot, width = 15, height = 10, dpi = 300)

```

#
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
density_categories  <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/density_categories.csv")

merged_data <- merge(data, density_categories, by = "site")

# Convert 'treatment' and 'urchin_density' to factors
merged_data$treatment <- as.factor(merged_data$treatment)
merged_data$urchin_density <- as.factor(merged_data$Dens_Cat)


# Mixed-effects model for kelp survival
growth_model <- lm(kelp_growth ~ Dens_Cat, data = merged_data)

# Check the model summary
summary(growth_model)


# Survival 
survival_model <- lm(kelp_survival ~ Dens_Cat, data = merged_data)

# Check the model summary
summary(survival_model)

# Mixed-effects model for kelp survival
growth_model <- lm(kelp_growth ~ gravel_size * mean_current, data = merged_data)

# Check the model summary
summary(growth_model)


# Survival 
survival_model <- lm(kelp_survival ~ gravel_size * mean_current, data = merged_data)

# Check the model summary
summary(survival_model)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
