---
title: "Kelp_Survival"
author: "Tessa Rehill"
date: "2024-12-07"
output: html_document
---
#what is driving kelp growth and survival? Or kelp loss... 
#


#make some visuals of survival and growth 
#test some stats 

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

#Set up libraries 
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(car)
library(multcomp)
library(corrplot)
library(dunn.test)
library(lme4)
```

#Read in files 
```{r}
transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_monitoring_transects.csv")

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

env <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_ysi_measurement.csv")

hobo <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/ENV/HOBO/summary_hobo.csv")

dens <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/density_urchin_transects.csv")
```

#Mean survival + growth at each site at each month, per treatment 
```{r}

#order sites on x-axis 
plots$site <- factor(plots$site, levels = c("Nanat", "Dixon", "Ellis S", "Ellis N", "Ed King", "Taylor"))

#SURVIVAL #####################################################################
plots$month <- factor(plots$month, levels = c("April", "May", "June", "August"))

plot <- ggplot(plots, aes(x = site, y = kelp_survival, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  facet_wrap(~ month, scales = "fixed", ncol = 1) +         # Stack facets vertically
  labs(
    x = "Site",
    y = "Kelp Survival (%)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis range
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot

#save
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_survival/survival_site_month.png", plot = plot, width = 10, height = 15, dpi = 300)


#GROWTH  ##########################################################################
plot <- ggplot(plots, aes(x = site, y = kelp_growth, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  facet_wrap(~ month, scales = "free_y", ncol = 1) +        # Stack facets vertically with free y-scales
  labs(
    x = "Site",
    y = "Kelp Growth (mm)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_growth/growth_site_month.png", plot = plot, width = 15, height = 10, dpi = 300)

###now with log growth
plots <- plots %>%
  mutate(log_kelp_growth = log(kelp_growth + 1))  # Add 1 to avoid log(0) or negative values

plot <- ggplot(plots, aes(x = site, y = log_kelp_growth, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  facet_wrap(~ month, scales = "free_y", ncol = 1) +        # Stack facets vertically with free y-scales
  labs(
    x = "Site",
    y = "LOG Kelp Growth (mm)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_growth/log_growth_site_month.png", plot = plot, width = 10, height = 8, dpi = 300)

```
#Mean survival + growth at each site at each month, per gravel size 
```{r}
#SURVIVAL #####################################################################
plots$month <- factor(plots$month, levels = c("April", "May", "June", "August"))

plot <- ggplot(plots, aes(x = site, y = kelp_survival, fill = gravel_size)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  facet_wrap(~ month, scales = "fixed", ncol = 1) +        # Stack facets vertically with fixed y-axis
  labs(
    x = "Site",
    y = "Kelp Survival (%)",
    fill = "Gravel Size"
  ) +
  scale_fill_manual(values = c("large" = "#f3ac46", "small" = "#aae9ad")) +  # Custom colors
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis range
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot 

#save
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/gravel_size/gravel_site_month.png", plot = plot, width = 10, height = 15, dpi = 300)

#GROWTH  ##########################################################################
plot <- ggplot(plots, aes(x = site, y = kelp_growth, fill = gravel_size)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  facet_wrap(~ month, scales = "free_y", ncol = 1) +        # Stack facets vertically with free y-scales
  labs(
    x = "Site",
    y = "Kelp Growth (mm)",
    fill = "Gravel Size"
  ) +
  scale_fill_manual(values = c("large" = "#f3ac46", "small" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/gravel_size/gravel_site_month.png", plot = plot, width = 10, height = 15, dpi = 300)
```

#Boxplot of survival and growth, in open v cage, run t-test
```{r}

#SURVIVAL #####################################################################
#run t-test
#shapiro-Wilk test for normality, p less than 0.05, normally distributed
shapiro.test(plots$kelp_survival[plots$treatment == "caged"]) #normally distributed 
shapiro.test(plots$kelp_survival[plots$treatment == "open"]) #normally distributed 

#check variances, if the p-value from Levene’s test is greater than 0.05, you can assume that the variances are equal, and the assumption is met. 
leveneTest(kelp_survival ~ treatment, data = plots) #p = 0.7304

#t-test to compare kelp survivorship between treatments
t_test_result  <- t.test(kelp_survival ~ treatment, data = plots) #p = 0.7695
t_test_result
  #mean in caged = 13.96%, mean in open = 12.98%

p_value <- formatC(t_test_result$p.value, format = "e", digits = 2)

#visualize 
plot <- ggplot(plots, aes(x = treatment, y = kelp_survival, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  labs(
    x = "Treatment",
    y = "Kelp Survival (%)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  annotate(
    "text",
    x = 1.5,  # Midpoint between two treatment groups
    y = max(plots$kelp_survival, na.rm = TRUE) + 5,  # Adjusted Y position
    label = paste("p =", p_value),
    size = 5,
    hjust = 0.5
  )

plot

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_survival/surv_open_v_caged.png", plot = plot, width = 10, height = 8, dpi = 300)

#GROWTH ############################################################
#run t-test
#shapiro-Wilk test for normality, p less than 0.05, normally distributed
shapiro.test(plots$kelp_growth[plots$treatment == "caged"]) #normally distributed 
shapiro.test(plots$kelp_growth[plots$treatment == "open"]) #normally distributed 

#check variances, if the p-value from Levene’s test is greater than 0.05, you can assume that the variances are equal, and the assumption is met. 
leveneTest(kelp_growth ~ treatment, data = plots) #p = 0.4006 #normal 

#t-test to compare kelp growth between treatments
t_test_result  <- t.test(kelp_growth ~ treatment, data = plots) #p = 0.4022
t_test_result
  #mean in caged = 43.15mm, mean in open = 66.09mm

p_value <- formatC(t_test_result$p.value, format = "e", digits = 2)
#visualize 
plot <- ggplot(plots, aes(x = treatment, y = kelp_growth, fill = treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  labs(
    x = "Treatment",
    y = "Kelp Growht (mm)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  annotate(
    "text",
    x = 1.5,  # Midpoint between two treatment groups
    y = max(plots$kelp_survival, na.rm = TRUE) + 5,  # Adjusted Y position
    label = paste("p =", p_value),
    size = 5,
    hjust = 0.5
  )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_growth/grwt_open_v_caged.png", plot = plot, width = 10, height = 8, dpi = 300)

#LOG GROWTH ############################################################
#let's log growth and perform the same t-test
shapiro.test(plots$log_kelp_growth[plots$treatment == "caged"]) #normally distributed 
shapiro.test(plots$log_kelp_growth[plots$treatment == "open"]) #normally distributed 

#check variances, if the p-value from Levene’s test is greater than 0.05, you can assume that the variances are equal, and the assumption is met. 
leveneTest(log_kelp_growth ~ treatment, data = plots) #p = 0.7587 #normal 

#run t-test for log-transformed kelp growth by treatment
t_test_result <- t.test(log_kelp_growth ~ treatment, data = plots, var.equal = TRUE)

# Print the result
t_test_result #no difference, slightly MORE growth in open... #p = 0.9639

#visualize 
plot <- ggplot(plots, aes(x = treatment, y = log_kelp_growth, fill=treatment)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +  # Side-by-side boxplots
  labs(
    x = "Site",
    y = "Log - Kelp Growth (mm)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_growth/loggrwt_open_v_caged.png", plot = plot, width = 10, height = 8, dpi = 300)

```
#Difference in survivorship by site 
```{r}

#summarize survivorship by site for open and caged plots 
surv_summary <- plots %>%
  group_by(site, treatment) %>%
  summarise(
    mean_survival = mean(kelp_survival, na.rm = TRUE),
    sd_survival = sd(kelp_survival, na.rm = TRUE),
    min_survival = min(kelp_survival, na.rm = TRUE),
    max_survival = max(kelp_survival, na.rm = TRUE),
    n = n()  # Number of observations
  )

#plotting the summarized data
plot <- ggplot(surv_summary, aes(x = site, y = mean_survival, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), width = 0.7) +  # Bar plot
  geom_errorbar(aes(ymin = pmax(mean_survival - sd_survival, 0), 
                    ymax = mean_survival + sd_survival), 
                position = position_dodge(width = 0.75), width = 0.25) +  # Error bars
  labs(
    x = "Site",
    y = "Mean Kelp Survival (%)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot 

#plotting with standard error (rather than standard deviation)
kelp_summary <- plots %>%
  group_by(site, treatment) %>%
  summarise(
    mean_survival = mean(kelp_survival, na.rm = TRUE),
    sd_survival = sd(kelp_survival, na.rm = TRUE),
    n = n(),
    se_survival = sd_survival / sqrt(n)
  )

plot <- ggplot(kelp_summary, aes(x = site, y = mean_survival, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), width = 0.7) +  # Bar plot
  geom_errorbar(aes(ymin = pmax(mean_survival - se_survival, 0), 
                    ymax = mean_survival + se_survival), 
                position = position_dodge(width = 0.75), width = 0.25) +  # Error bars
  labs(
    x = "Site",
    y = "Mean Kelp Survival (%)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_survival/survival_by_month_se.png", plot = plot, width = 10, height = 8, dpi = 300)

#Is there a difference in kelp survivorship across sites (irrelevant of treatment)?

#assess normality - histogram 
ggplot(plots, aes(x = kelp_survival, fill = site)) +
  geom_histogram(binwidth = 5, alpha = 0.7, position = "dodge") +
  facet_wrap(~ site) +
  labs(title = "Kelp Survival Distribution by Site") +
  theme_minimal()

#Statistically test for differences 
#check for normal data using Shapiro test 
plots %>%
  group_by(site) %>%
  summarise(shapiro_p_value = shapiro.test(kelp_survival)$p.value) #all normal

#run Levene's test for homogeneity of variance, if P < 0.05, significantly different variances
levene_result <- leveneTest(kelp_survival ~ site, data = plots)
levene_result #significantly different variances

#Run Welch's ANOVA for kelp survival across sites
welch_anova_result <- oneway.test(kelp_survival ~ site, data = plots, var.equal = FALSE)
welch_anova_result #significant differences between sites, p = 0.0007222

#which sites are most different from each other (Post Hoc pairwise comparison)
#post-hoc pairwise t-tests (with Welch's correction) after significant ANOVA
pairwise_result <- pairwise.t.test(plots$kelp_survival, plots$site, p.adjust.method = "bonferroni")
pairwise_result #no significances between sites

#no significance?? post-hoc pairwise comparisons show that, despite the overall significance from the ANOVA, the individual differences between sites are not significant when the Bonferroni correction is applied, which accounts for multiple testing.

#Kruskal test
kruskal.test(kelp_survival ~ site, data = plots) #p = 0.01

#Dunn test 
dunn.test(plots$kelp_survival, plots$site, kw = TRUE, label = TRUE) 



```
#Differences in growth among sites 
```{r}
#summarize survivorship by site for open and caged plots 
grwt_summary <- plots %>%
  group_by(site, treatment) %>%
  summarise(
    mean_growth = mean(kelp_growth, na.rm = TRUE),
    sd_growth = sd(kelp_growth, na.rm = TRUE),
    min_growth = min(kelp_growth, na.rm = TRUE),
    max_growth = max(kelp_growth, na.rm = TRUE),
    n = n()  # Number of observations
  )

# Plotting the summarized data
ggplot(grwt_summary, aes(x = site, y = mean_growth, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), width = 0.7) +  # Bar plot
  geom_errorbar(aes(ymin = pmax(mean_growth - sd_growth, 0), 
                    ymax = mean_growth + sd_growth), 
                position = position_dodge(width = 0.75), width = 0.25) +  # Error bars
  labs(
    x = "Site",
    y = "Mean Growth (mm)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

#plotting with standard error (rather than standard deviation)
kelp_summary <- plots %>%
  group_by(site, treatment) %>%
  summarise(
    mean_growth = mean(kelp_growth, na.rm = TRUE),
    sd_growth = sd(kelp_growth, na.rm = TRUE),
    n = n(),
    se_growth = sd_growth / sqrt(n)
  )

plot <- ggplot(kelp_summary, aes(x = site, y = mean_growth, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), width = 0.7) +  # Bar plot
  geom_errorbar(aes(ymin = pmax(mean_growth - se_growth, 0), 
                    ymax = mean_growth + se_growth), 
                position = position_dodge(width = 0.75), width = 0.25) +  # Error bars
  labs(
    x = "Site",
    y = "Mean Kelp Growth (mm)",
    fill = "Treatment"
  ) +
  scale_fill_manual(values = c("open" = "#f3ac46", "caged" = "#aae9ad")) +  # Custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

plot 

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/kelp_growth/growth_by_month_se.png", plot = plot, width = 10, height = 8, dpi = 300)


#Is there a difference in kelp growth across sites (irrelevant of treatment)?

# Histogram for each site to assess normality
ggplot(plots, aes(x = kelp_growth, fill = site)) +
  geom_histogram(binwidth = 5, alpha = 0.7, position = "dodge") +
  facet_wrap(~ site) +
  labs(title = "Kelp Growth Distribution by Site") +
  theme_minimal()

#check for normal data using Shapiro test 
plots %>%
  group_by(site) %>%
  summarise(shapiro_p_value = shapiro.test(kelp_growth)$p.value) #all normal

# Run Levene's test for homogeneity of variance, if P < 0.05, significantly different variances
levene_result <- leveneTest(kelp_growth ~ site, data = plots)
levene_result #significantly different variances

#run Welch's ANOVA for kelp growth across sites
welch_anova_result <- oneway.test(kelp_growth ~ site, data = plots, var.equal = FALSE)
welch_anova_result #significant differences between sites, p = 0.002392

#which sites are most different from each other (Post Hoc pairwise comparison)
# Post-hoc pairwise t-tests (with Welch's correction) after significant ANOVA
pairwise_result <- pairwise.t.test(plots$kelp_growth, plots$site, p.adjust.method = "bonferroni")
pairwise_result #only significant difference between site = Nanat (this is an important finding to me)

#Kruskal test
kruskal.test(kelp_growth ~ site, data = plots) #p = 0.0008481

#Dunn test 
dunn.test(plots$kelp_growth, plots$site, kw = TRUE, label = TRUE) 
  #Ellis N + Ed King (p = 0.0140)
  #Ellis S + Ellis N (p = 0.0025)
  #Nanat + Ed King (p= 0.0038)
  #Nanat + Ellis S (p=0.0004)
  #Taylor + Dixon (p = 0.0083)
  #Taylor Ellis N ( p = 0.0005)
  #Taylor + Nanat (p = 0.0001)



```

#Does kelp growth and survivorship differ across months? 
```{r}

#SURVIVORSHIP ##################################################################
# Test for kelp growth across months
# Check normality
shapiro_results_surv <- plots %>%
  group_by(month) %>%
  summarize(p_value = shapiro.test(kelp_survival)$p.value)

shapiro_results_surv #normal

# Check homogeneity of variance
levene_surv <- leveneTest(kelp_survival ~ month, data = plots)
levene_surv # p = 2.987e-078

#Perform Kruskal test
kw_surv <- kruskal.test(kelp_survival ~ month, data = plots)
kw_surv #p = 4.156e-10

#Dunn test 
dunn_surv <- dunn.test(plots$kelp_survival, plots$month, kw = TRUE, label = TRUE)
dunn_surv

#survivorship differs across all months 

#GROWTH #########################################################################
# Test for kelp growth across months
# Check normality
shapiro_results_growth <- plots %>%
  group_by(month) %>%
  summarize(p_value = shapiro.test(kelp_growth)$p.value)
shapiro_results_growth #normal

# Check homogeneity of variance
levene_growth <- leveneTest(kelp_growth ~ month, data = plots)
levene_growth # p = 0.0371

#Perform Kruskal test
kw_growth <- kruskal.test(kelp_growth ~ month, data = plots)
kw_growth #p = 0.02397

#Dunn test 
dunn_growth <- dunn.test(plots$kelp_growth, plots$month, kw = TRUE, label = TRUE)
dunn_growth


```





#CODE GRAVEYARD ###############################################################
#Add site density to plots dataframe (to be used as a predictor in our models)
#And environment 
```{r}
#change column name urchin_density in dens to site_density
dens <- dens %>% rename(site_density = urchin_density)

# Merge dens into plots by site and month
plots_glm <- merge(plots, dens[, c("site", "month", "mean_depth_m", "site_density")], 
               by = c("site", "month"), 
               all.x = TRUE)

#create env summary, for bottom temperature, do, ph, salinity
bottom_data <- env %>%
 group_by(site, month) %>%
  summarize(
    bottom_temp = temp[which.max(depth_m)],
    bottom_do = do[which.max(depth_m)],
    bottom_salinity = salinity[which.max(depth_m)],
    bottom_ph = ph[which.max(depth_m)],
    .groups = "drop" # Ungroups after summarization
  )

#make month a character
bottom_data <- bottom_data %>%
  mutate(month = as.character(month))

#merge environmental data and depth data by site/month to plot data 
plots_glm <- plots_glm %>%
  left_join(hobo, by = c("site", "month")) %>%
  left_join(bottom_data, by = c("site", "month"))

selected_columns <- c("site", "month", "plot_id", "date", "treatment", "gravel_size", "kelp_survival", "urchin_density", "kelp_growth", "mean_depth_m", "site_density", "mean_temperature_c", "mean_light_lux", "bottom_do", "bottom_salinity", "percent_occuring_algae", "plot_id")

model_df <- plots_glm[, selected_columns]

```

#Zero-inflated generalized linear mixed-effects models
```{r}
summary(model_df)

model_df$log_kelp_survival <- log(model_df$kelp_survival + 1)

#create proportion (between 0 > x > 1) for kelp survival 
model_df$kelp_proportion <- model_df$kelp_survival/100

#model_df$log_kelp_proportion<- log(model_df$kelp_proportion)
model_df$log_kelp_proportion <- log(model_df$kelp_proportion + 1)

model_df$kelp_proportion_transformed <- pmax(pmin(model_df$log_kelp_proportion, 1 - 1e-5), 1e-5)

model_df$log_kelp_growth <- log(model_df$kelp_growth + 1)

model_df$kelp_growth_transformed <- pmax(pmin(model_df$log_kelp_growth, 1 - 1e-5), 1e-5)


#change month to integer 
  #convert 'month' to a factor
    model_df$month <- as.factor(model_df$month)

  #convert the factor levels to numeric
    model_df$month <- as.numeric(model_df$month)

#CHECK ENVIRONMENTAL PREDICTORS #################################################
#correlation matrix of predictors 
cor_matrix <- cor(model_df[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", 
                               "bottom_do", "urchin_density", "percent_occuring_algae", "site_density")], 
                  use = "complete.obs")

print(cor_matrix)

#check VIF factor 
lm_model <- lm(kelp_proportion_transformed ~ mean_light_lux + percent_occuring_algae + 
               mean_temperature_c + bottom_salinity + mean_depth_m + 
               bottom_do + month + urchin_density + site_density, data = model_df)

vif(lm_model) #salinity, bottom_do, mean_temp, is too high 

lm_model <- lm(kelp_proportion_transformed ~ mean_light_lux + percent_occuring_algae + mean_temperature_c + 
                mean_depth_m + bottom_do + month + urchin_density + site_density, 
                data = model_df)

vif(lm_model) #all below 5 

#standardizing predictors
model_df$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df$bottom_do <- scale(model_df$bottom_do)
model_df$mean_light_lux <- scale(model_df$mean_light_lux)
model_df$mean_depth_m <- scale(model_df$mean_depth_m)
model_df$urchin_density <- scale(model_df$urchin_density)
model_df$site_density <- scale(model_df$site_density)

#month needs to be turned back into a factor
model_df$month <- factor(model_df$month)
model_df$treatment <- as.factor(model_df$treatment)
model_df$gravel_size <- as.factor(model_df$gravel_size)

#MODEL FITTING ##################################################################
#model_bin <- glmmTMB(log_kelp_proportion ~ treatment + gravel_size + 
#                      mean_temperature_c + mean_depth_m + bottom_do + 
#                      urchin_density + site_density + (1 | site),
#                      ziformula = ~ 1,
#                      family = binomial(), #try poisson
#                      data = model_df)


#summary(model_bin)

#GAUSSIAN 
#Gaussian 
model_gau_1 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      mean_temperature_c + mean_depth_m + bottom_do + 
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = gaussian(), 
                      data = model_df)
summary(model_gau_1)
AIC(model_gau_1)

model_gau_2 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      mean_temperature_c + mean_depth_m + bottom_do + 
                      urchin_density + site_density + (1 | site) + (1 | month),
                      ziformula = ~ 1,
                      family = gaussian(), 
                      data = model_df)
summary(model_gau_2)
AIC(model_gau_2)

model_gau_3 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = gaussian(), 
                      data = model_df)
summary(model_gau_3)
AIC(model_gau_3)

model_gau_4 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + mean_temperature_c + mean_depth_m + (1 | site),
                      ziformula = ~ 1,
                      family = gaussian(), 
                      data = model_df)
summary(model_gau_4)
AIC(model_gau_4)

model_gau_5<- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + mean_temperature_c + (1 | site),
                      ziformula = ~ 1,
                      family = gaussian(), 
                      data = model_df)
summary(model_gau_5)
AIC(model_gau_5)

model_gau_6<- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + mean_depth_m + (1 | site),
                      ziformula = ~ 1,
                      family = gaussian(), 
                      data = model_df)
summary(model_gau_6)
AIC(model_gau_6)

AIC(model_gau_1, model_gau_2, model_gau_3, model_gau_4, model_gau_5, model_gau_6)

#POISSON

model_df$log_kelp_survival <- as.integer(model_df$log_kelp_survival)
model_df$kelp_survival <- as.integer(model_df$kelp_survival)


model_poi_1 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      mean_temperature_c + mean_depth_m + bottom_do + 
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

summary(model_poi_1)
AIC(model_poi_1) #lowest AIC 

model_poi_1 <- glmmTMB(kelp_survival ~ treatment + gravel_size + 
                      mean_temperature_c + mean_depth_m + bottom_do + 
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

summary(model_poi_1)
AIC(model_poi_1) #lowest AIC 




##
model_poi_2 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      mean_temperature_c +
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

summary(model_poi_2)

##
model_poi_3 <- glmmTMB(log_kelp_survival ~ treatment +
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

summary(model_poi_3)

##
model_poi_4 <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

summary(model_poi_4)


AIC(model_poi_1, model_poi_2, model_poi_3, model_poi_4)





#
model_poi <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + (1 | site) + (1 | month),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

AIC(model_poi)

#
model_poi <- glmmTMB(log_kelp_survival ~ treatment + gravel_size + 
                      urchin_density + site_density + (1 | site) + (1 | month),
                      ziformula = ~ 1,
                      family = poisson(), 
                      data = model_df)

AIC(model_poi)



```

#Validating model - use DHARMa package
```{r}
library(DHARMa)

fittedModel <- model_poi_1

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

```

```{r}






model<- model_gau_4
response <- model_df$log_kelp_survival

#visualize
plot_data <- data.frame(Fitted = fitted(model),
                        Residuals = resid(model))

#plot residuals vs. fitted values
plot <- ggplot(plot_data, aes(x = Fitted, y = Residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

plot 

#QQ plot

residuals <- residuals(model)

plot <- qqnorm(residuals, main = "QQ Plot of Residuals")
plot <- qqline(residuals, col = "red", lwd = 2)

plot

# Predicted vs Observed
predicted_values <- predict(model, type = "response")
plot(predicted_values, model_df$response, main = "Predicted vs Observed")
abline(0, 1, col = "red")  # Line representing perfect prediction

```



#CODE GRAVEYARD ################################################################

#SURVIVAL - MODEL GLMM 
```{r}
#check missing values
summary(model_df)

#log predictor variable - 
model_df$kelp_survival_log <- log(model_df$kelp_survival + 1)  
  #If kelp_survival is 0 or 100%, add 1 before log

summary(model_df)

#change month to integer 
  #convert 'month' to a factor
    model_df$month <- as.factor(model_df$month)

  #convert the factor levels to numeric
    model_df$month <- as.numeric(model_df$month)

#CHECK ENVIRONMENTAL PREDICTORS #################################################
#correlation matrix of predictors 
cor_matrix <- cor(model_df[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", 
                               "bottom_do", "urchin_density", "percent_occuring_algae", "site_density")], 
                  use = "complete.obs")

print(cor_matrix)

#check VIF factor 
lm_model <- lm(kelp_survival_log ~ mean_light_lux + percent_occuring_algae + 
               mean_temperature_c + bottom_salinity + mean_depth_m + 
               bottom_do + month + urchin_density + site_density, data = model_df)

vif(lm_model) #salinity, bottom_do, mean_temp, is too high 

lm_model <- lm(kelp_survival_log ~ mean_light_lux + percent_occuring_algae + mean_temperature_c + 
                mean_depth_m + bottom_do + month + urchin_density + site_density, 
                data = model_df)

vif(lm_model) #all below 5 

#standardizing predictors
model_df$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df$bottom_do <- scale(model_df$bottom_do)
model_df$mean_light_lux <- scale(model_df$mean_light_lux)
model_df$mean_depth_m <- scale(model_df$mean_depth_m)
model_df$urchin_density <- scale(model_df$urchin_density)
model_df$site_density <- scale(model_df$site_density)

#month needs to be turned back into a character
model_df$month <- factor(model_df$month)

#FITTING MODEL ##################################################################
#zero inflated 

#Fit mixed-effects model for survival 
model <- lmer(kelp_survival_log ~ treatment +  mean_light_lux + mean_temperature_c + 
                   bottom_do  + urchin_density + site_density + 
                   (1|site) + (1|month) + (1|plot_id) + (1|mean_depth_m), data = model_df)


summary(model)

AIC(model)

model <- lmer(kelp_survival_log ~ treatment + urchin_density + 
                   (1|site) + (1|month) + (1|plot_id) + (1|mean_depth_m), data = model_df)


summary(model)

AIC(model)


##
model <- lmer(kelp_survival_log ~ treatment +  mean_light_lux + mean_temperature_c + 
                   bottom_do  + urchin_density + site_density + 
                   (1|site) + (1|month) + (1|plot_id) + (1|mean_depth_m), data = model_df)


summary(model)

AIC(model)

model <- final 

#visualize
plot_data <- data.frame(Fitted = fitted(model),
                        Residuals = resid(model))

#plot residuals vs. fitted values
plot <- ggplot(plot_data, aes(x = Fitted, y = Residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

plot 

#ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/residualvfitted.png", 
#       plot = combined_plot, 
#       width = 10, height = 20, dpi = 300)

#QQ plot

residuals <- residuals(model)

plot <- qqnorm(residuals, main = "QQ Plot of Residuals")
plot <- qqline(residuals, col = "red", lwd = 2)

plot
#ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/residualvfitted.png", 
#       plot = combined_plot, 
#       width = 10, height = 20, dpi = 300)

#ANOVA 
#perform ANOVA on the linear model
anova_result <- anova(model)

#print the ANOVA table
print(anova_result)


```
```{r}
#zero inflated beta regression model 

#standardizing predictors
model_df$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df$bottom_do <- scale(model_df$bottom_do)
model_df$mean_light_lux <- scale(model_df$mean_light_lux)
model_df$mean_depth_m <- scale(model_df$mean_depth_m)
model_df$urchin_density <- scale(model_df$urchin_density)
model_df$site_density <- scale(model_df$site_density)

model_df$kelp_proportion <- model_df$kelp_survival/100

model_df$kelp_proportion_transformed <- pmax(pmin(model_df$kelp_proportion, 1 - 1e-5), 1e-5)


model_zoib <- glmmTMB(kelp_proportion_transformed ~ treatment + gravel_size + 
                      mean_temperature_c  + mean_depth_m + bottom_do + 
                      urchin_density + site_density + (1 | site),
                      ziformula = ~ 1,
                      family = beta_family(),
                      data = model_df,
                      control = glmmTMBControl(optimizer = "bobyqa"))



summary(model_zoib)
```


```{r}
model_df$kelp_proportion <- model_df$kelp_survival/100

#rescale the variables
model_df_scaled <- model_df
model_df_scaled$mean_light_lux <- scale(model_df$mean_light_lux)
model_df_scaled$percent_occuring_algae <- scale(model_df$percent_occuring_algae)
model_df_scaled$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df_scaled$bottom_salinity <- scale(model_df$bottom_salinity)
model_df_scaled$mean_depth_m <- scale(model_df$mean_depth_m)
model_df_scaled$bottom_do <- scale(model_df$bottom_do)
model_df_scaled$urchin_density <- scale(model_df$urchin_density)
model_df_scaled$site_density <- scale(model_df$site_density)


mixed_model <- glmer(kelp_proportion ~ mean_light_lux + percent_occuring_algae + 
                     mean_temperature_c + bottom_salinity + mean_depth_m + 
                     bottom_do + urchin_density + site_density + 
                     (1 | site) + (1 | month), family = binomial(link = "logit"), data = model_df_scaled, control = glmerControl(optimizer = "bobyqa"))


summary(mixed_model)


#

mixed_model_convergence <- glmer(kelp_proportion ~ mean_light_lux + percent_occuring_algae + 
                                 mean_temperature_c + bottom_salinity + month + 
                                 bottom_do + urchin_density + site_density + 
                                 (1 | site) + (1 | mean_depth_m) , 
                                 family = binomial(link = "logit"), 
                                 data = model_df_scaled, 
                                 control = glmerControl(optimizer = "bobyqa"))
summary(mixed_model_convergence)
```
#Beta regression - specific for values between 0-1
```{r}
library(betareg)
model_df$kelp_proportion <- model_df$kelp_survival / 100

# Apply logit transformation to kelp_proportion (values strictly between 0 and 1)
model_df$kelp_proportion_logit <- qlogis(model_df$kelp_proportion)

# Fit the model with random effects for site and month
model_glmm <- glmmTMB(kelp_proportion_logit ~ mean_light_lux + mean_temperature_c + 
                      mean_depth_m + bottom_do + urchin_density + site_density + 
                      (1|site) + (1|month) + (1|plot_id), 
                      data = model_df, 
                      family = beta_family(link = "logit"))

# View the summary of the model
summary(model_glmm)


```

PCA for survivorship? 

#MODEL - GROWTH 
```{r}
##
model_df$kelp_growth_log <- log(model_df$kelp_growth + 1)  # If kelp growth is 0 or 100%, add 1 before log

predictor <- model_df$kelp_growth_log

model <- lmer(predictor~ treatment + gravel_size + mean_light_lux + mean_temperature_c + 
                   mean_depth_m + bottom_do  + urchin_density + site_density + 
                   (1|site) + (1|month), data = model_df)


summary(model)

#visualize
# Create a dataframe for plotting
plot_data <- data.frame(Fitted = fitted(model),
                        Residuals = resid(model))

# Plot residuals vs. fitted values
plot <- ggplot(plot_data, aes(x = Fitted, y = Residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

plot 

#ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/residualvfitted.png", 
#       plot = combined_plot, 
#       width = 10, height = 20, dpi = 300)

#QQ plot
# Extract residuals from the model
residuals <- residuals(model)

# Create a QQ plot
plot <- qqnorm(residuals, main = "QQ Plot of Residuals")
plot <- qqline(residuals, col = "red", lwd = 2)

plot
#ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/urchin_densities/residualvfitted.png", 
#       plot = combined_plot, 
#       width = 10, height = 20, dpi = 300)

# Perform ANOVA on the linear model
anova_result <- anova(model)

# Print the ANOVA table
print(anova_result)



```

```{r}
#turn survival into a proportion (between 0-1)
model_df$kelp_survival <- model_df$kelp_survival/100



mixed_model <- glmer(kelp_survival_log ~ mean_light_lux + percent_occuring_algae + 
                     mean_temperature_c + bottom_salinity + mean_depth_m + 
                     bottom_do + month + urchin_density + site_density + 
                     (1 | site), family = binomial(link = "logit"), data = model_df)
summary(mixed_model)
```

#zero inflated beta regression model 
```{r}
#zero inflated beta regression model 
model_df$kelp_proportion_transformed <- pmax(pmin(model_df$kelp_proportion, 1 - 1e-5), 1e-5)


model_zoib <- glmmTMB(kelp_proportion_transformed ~ mean_light_lux + percent_occuring_algae + 
                      mean_temperature_c + bottom_salinity + mean_depth_m + 
                      bottom_do + urchin_density + site_density + 
                      (1 | site), # Random intercept for 'site'
                      ziformula = ~ 1,  # Zero-inflation part (only intercept)
                      family = beta_family(),  # Beta regression
                      data = model_df)

summary(model_zoib)
```

