---
title: "Survivor_Models_WORKFLOW"
author: "Tessa Rehill"
date: "2024-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)


```

#Run BETA regression GLMM with forward feature selection #######################
  #load data
```{r}

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```

  #create kelp survival as a proportion, and add number kelp and gravel to dataframe 
```{r}

selected_columns <- plots[, c("site", "month", "plot_id", "n_kelp_gravel", "n_gravel")]

#merge the selected columns to 'data' by 'site' and 'month'
data <- merge(data, selected_columns, by = c("site", "month", "plot_id"), all.x = TRUE)

#kelp survival as proportion
data$kelp_prop <- data$kelp_survival/100

#transform response variable with small constant (beta regression can not handle zeros)
data$kelp_prop_adj <- (data$kelp_prop * (nrow(data) - 1) + 0.5) / nrow(data)

#create new columns for the success (kelp) and failure (non-kelp)
data$kelp_success <- data$n_kelp_gravel
data$kelp_failure <- data$n_gravel - data$n_kelp_gravel

#kelp survival as integer
data$kelp_survival <- as.integer(data$kelp_survival)

summary(data)

```

  #considering overdispersion and zero inflation 
```{r}
#check for overdispersion  (in kelp proportion without constant )
mean_survival <- mean(data$kelp_prop, na.rm = TRUE)
var_survival <- var(data$kelp_prop, na.rm = TRUE)
overdispersion <- var_survival / mean_survival 
print(overdispersion)  #38 - high dispersion

#check for overdispersion (in kelp proportion adjusted with a constant)
mean_survival <- mean(data$kelp_prop_adj, na.rm = TRUE)
var_survival <- var(data$kelp_prop_adj, na.rm = TRUE)
overdispersion <- var_survival / mean_survival 
print(overdispersion)  #37 - high dispersion


#check for zero inflation 
sum(data$kelp_prop == 0, na.rm = TRUE) / nrow(plots) #a lot of zeros (55%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

sum(data$kelp_prop_adj == 0, na.rm = TRUE) / nrow(plots) #no zeros 

```

  #checking correlation of predictors 
```{r}


cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "bottom_ph", "urchin_density", "percent_occuring_algae", "site_density", "shannon_div", "simp_div", "species_richness")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/correlation_plot.png", plot = plot, width = 10, height = 8, dpi = 300)


#Compute VIF factor 
#with all predictors 
lm_model <- lm(kelp_prop_adj ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m + bottom_do + species_richness + 
               shannon_div + simp_div  + site_density + urchin_density, data = data)

vif(lm_model) #so many >5


```

  #scaling predictors 
```{r}

#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$urchin_density <- scale(data$urchin_density)
data$site_density <- scale(data$site_density)
data$simp_div <- scale(data$simp_div)
data$shannon_div <- scale(data$shannon_div)
data$species_richness <- scale(data$species_richness)
data$percent_occuring_algae <- scale(data$percent_occuring_algae)

#month needs to be turned back into a factor
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)


```

  #simplest model 
```{r}
simple_beta <- glmmTMB(
  kelp_prop_adj ~ month + gravel_size + treatment + urchin_density + (1 | site)  + (1 | plot_id), 
  data = data,
  family = beta_family(link = "logit")
)

summary(simple_beta)
AIC(simple_beta) 

#final model
fittedModel <- simple_beta

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

```

#Define predictors and base model structure 
```{r}
#define predictors and initial formula
predictors <-  c( "percent_occuring_algae", "treatment", "site_density", "urchin_density",
                "gravel_size",  "mean_temperature_c", "mean_light_lux", 
                "bottom_do", "bottom_temp", "bottom_salinity", "mean_depth_m", 
                "simp_div", "species_richness", "bottom_ph")

#initialize the empty model formula (with no predictors)
base_formula <- "kelp_prop_adj ~ 1 + (1 | site)  + (1 | plot_id)  + (1 | month)"  # Initial random effects
```

  #beta regression GLM - forward feature selection
```{r}

#set up a function to perform forward feature selection
forward_feature_selection <- function(predictors, base_formula, data) {
  
  #initialize the best model to the base model
  best_model <- glmmTMB(as.formula(base_formula), data = data, family = beta_family(link = "logit"))
  best_aic <- AIC(best_model)
  best_formula <- base_formula
  selected_predictors <- character(0)  # No predictors selected initially
  
  #loop through the remaining predictors and add one at a time
  while (length(predictors) > 0) {
    
    #try adding each remaining predictor to the model
    candidate_aic <- numeric(length(predictors))
    for (i in 1:length(predictors)) {
      
      #create the new formula with the current predictor
      new_formula <- paste(best_formula, "+", predictors[i])
      
      #fit the new model
      model <- glmmTMB(as.formula(new_formula), data = data, family = beta_family(link = "logit"))
      
      #calculate the AIC of the new model
      candidate_aic[i] <- AIC(model)
    }
    
    #identify the model with the lowest AIC
    best_new_predictor <- predictors[which.min(candidate_aic)]
    if (min(candidate_aic) < best_aic) {
      
      #update the best model and the formula
      best_aic <- min(candidate_aic)
      best_formula <- paste(best_formula, "+", best_new_predictor)
      
      #add the best predictor to the selected predictors
      selected_predictors <- c(selected_predictors, best_new_predictor)
      
      #remove the selected predictor from the list of remaining predictors
      predictors <- setdiff(predictors, best_new_predictor)
      
      #print the updated model and AIC
      print(paste("Selected predictor:", best_new_predictor, "Current AIC:", best_aic))
    } else {
      #if no improvement, stop the selection process
      break
    }
  }
  
  #return the final model and selected predictors
  final_model <- glmmTMB(as.formula(best_formula), data = data, family = beta_family(link = "logit"))
  return(list(final_model = final_model, selected_predictors = selected_predictors, best_aic = best_aic))
}

#run forward feature selection
result <- forward_feature_selection(predictors, base_formula, data)

#view the selected predictors and the final model's AIC
print(paste("Selected predictors:", paste(result$selected_predictors, collapse = ", ")))
print(paste("Final model AIC:", result$best_aic))

#view the summary of the final model
summary(result$final_model)


```
 
```{r}
lm_model <- lm(kelp_prop_adj ~ gravel_size + mean_temperature_c + simp_div + species_richness + percent_occuring_algae, data = data)

vif(lm_model) #all under 5
 
 
  #best model 
beta_model <- glmmTMB(
  kelp_prop_adj ~ gravel_size + mean_temperature_c + simp_div + species_richness + percent_occuring_algae + (1 | site) + (1 | plot_id) + (1 | month), 
  data = data,
  family = beta_family(link = "logit")
)

AIC(beta_model)
summary(beta_model)

```

  #coefficient plot 
```{r}

#extract coefficients in tidy format, including confidence intervals
coef_tidy <- broom.mixed::tidy(beta_model, conf.int = TRUE)

#combine 'term' with additional info (if necessary) and remove 'component'
#here, we'll only use the 'term' column, assuming 'term' provides all the information you need

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "sd__(Intercept)")

# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +  # Apply a minimal theme
  labs(
    title = "Kelp Survival GLMM Coefficients and Confidence Intervals",
    x = "Coefficient Estimate",
    y = "Predictor"
  ) +
  theme(
    legend.position = "none",  # Remove the legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  ) +
  scale_color_manual(values = "#076fa6")  # Set line color

print(dw)

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/surv_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)

```


  #visualize model fit
```{r}

#final model
fittedModel <- beta_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/beta_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/beta_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/beta_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()


```

#Zero Inflated Negative Binomial GLMM - manual stepwise model selection ########
  #Read in files 
```{r}
transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_monitoring_transects.csv")

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")

```


  #Formatting predictors and response variables for GLMs
```{r}
model_df <- data

#standardizing predictors
model_df$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df$bottom_do <- scale(model_df$bottom_do)
model_df$bottom_ph <- scale(model_df$bottom_ph)
model_df$mean_light_lux <- scale(model_df$mean_light_lux)
model_df$mean_depth_m <- scale(model_df$mean_depth_m)
model_df$urchin_density <- scale(model_df$urchin_density)
model_df$site_density <- scale(model_df$site_density)
model_df$simp_div <- scale(model_df$simp_div)
model_df$species_richness <- scale(model_df$species_richness)
model_df$percent_occuring_algae <- scale(model_df$percent_occuring_algae)

#month needs to be turned back into a factor
model_df$month <- factor(model_df$month)
model_df$treatment <- as.factor(model_df$treatment)
model_df$gravel_size <- as.factor(model_df$gravel_size)
model_df$plot_id <- as.factor(model_df$plot_id)

#turn kelp survival into a integer 
model_df$kelp_survival <- as.integer(model_df$kelp_survival)

```

  #run with forward feature selection (terrible model fit) 
```{r}

#starting with a minimal model (only random effects)
base_model <- glmmTMB(kelp_survival ~ 1 + (1 | site) + (1 | plot_id), 
                      zi = ~ 1, # Zero-inflation model
                      family = nbinom2, 
                      data = model_df)

#check summary and AIC of the base model
summary(base_model)
AIC(base_model)

#forward stepwise selection by adding predictors one by one
predictor_list <- c("percent_occuring_algae", "treatment", "site_density", "urchin_density",
                    "gravel_size", "mean_temperature_c", "mean_light_lux", 
                    "bottom_do", "bottom_salinity", "bottom_ph", "mean_depth_m", 
                    "simp_div", "species_richness", "month")

#function to add predictors iteratively
add_predictors <- function(base_model, predictor_list) {
  current_model <- base_model
  current_AIC <- AIC(current_model)
  
  # Loop through predictor_list
  for (predictor in predictor_list) {
    # Create new formula by adding one predictor at a time
    formula <- as.formula(paste("kelp_survival ~", paste(c("1", predictor), collapse = " + "), 
                               "+ (1 | site) + (1 | plot_id)  "))
    
    # Define the zero-inflation portion of the model with relevant predictors
    zi_formula <- as.formula(paste("~", paste(predictor_list, collapse = " + ")))
    
    # Fit the new model with added predictor and updated zero-inflation formula
    new_model <- glmmTMB(formula, zi = zi_formula, family = nbinom2, data = model_df)
    
    # Compare AIC
    new_AIC <- AIC(new_model)
    
    # If the new model has a lower AIC, update the current model
    if (new_AIC < current_AIC) {
      current_model <- new_model
      current_AIC <- new_AIC
    }
  }
  
  return(current_model)
}

#run the forward selection
final_model <- add_predictors(base_model, predictor_list)

#check final model summary
summary(final_model)
AIC(final_model)


```

  #visualize model fit
```{r}

#final model
fittedModel <- final_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```

  #Model selection (manual)
```{r}
#include ALL predictors in first model 
model1 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do  + mean_depth_m + percent_occuring_algae + simp_div +
                  (1 | site) + (1 | month) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do,
                  family = nbinom2, 
                  data = model_df)
summary(model1)
AIC(model1) #939

#after stepwise AIC --> see old scripts 
#final model, lowest AIC, with predictors we are interested in 
model <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do  + simp_div + month + 
                  (1 | site) + (1 | plot_id),
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do  + simp_div + month,
                  family = nbinom2, 
                  data = model_df)

summary(model)
AIC(model) #935


final <- model

```


  #validate - visually
```{r}
#final model
fittedModel <- final

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/zinb_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/zinb_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/zinb_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```

  #create coefficient plot 
```{r}

#extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(final, conf.int = TRUE)

#combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

# Add the 'model' column
coef_tidy$model <- "Model 1"

#create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) + 
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(title = "Model Coefficients with Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Predictor",
       color = "Component")

# Print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 

#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/survival_models/zinb_dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)




```


  #other validation
```{r}

# 1. Residual Diagnostics for both models
# Residuals vs Fitted plot for final model (Poisson)
plot(fitted(final), residuals(final), main = "Residuals vs Fitted (final)", 
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")


# 2. Zero-Inflation Check
# Predicted probabilities of zero inflation for the final model
pred_zeros_final <- predict(final, type = "response", se.fit = TRUE)
plot(pred_zeros_final$fit, type = "l", main = "Predicted Zero Probabilities (final)", 
     xlab = "Observation", ylab = "Predicted Probability of Zero")

# 3. Model Comparison (AIC) - see stepwise AIC model selection
cat("AIC for final2 model (Negative Binomial):", AIC(final), "\n")

# 4. Dispersion Check for Overdispersion
# For final model (Poisson)
simulationOutput_final <- simulateResiduals(fittedModel = final, plot = TRUE)
testDispersion(simulationOutput_final)  # Check for overdispersion

# 5. Check Convergence
# Check model summaries to see if they converged properly
summary(final)

# 6. Fitted vs Observed Plot
# Fitted vs. Observed plot for final model (Poisson)
fitted_values_final <- fitted(final)
plot(model_df$kelp_survival, fitted_values_final, 
     main = "Observed vs Fitted (final)", xlab = "Observed", ylab = "Fitted")
abline(a = 0, b = 1, col = "red")  # 45-degree line

# 7. Cross-validation (optional)
set.seed(123)

# Split data into training and testing sets
trainIndex <- createDataPartition(model_df$kelp_survival, p = 0.8, list = FALSE)
trainData <- model_df[trainIndex, ]
testData <- model_df[-trainIndex, ]

# Train both models on training data
model_final_train <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month  + simp_div + 
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month + simp_div,
                  family = nbinom2, 
                  data = model_df)

# Predict on test data
predictions_final <- predict(model_final_train, newdata = testData)


# Calculate RMSE (Root Mean Squared Error)
RMSE_final <- sqrt(mean((predictions_final - testData$kelp_survival)^2))

cat("RMSE for final2 model (Negative Binomial):", RMSE_final, "\n")

# 8. Model Comparison (Likelihood Ratio Test)
anova(final, simple)  # Compare models using likelihood ratio test


fittedModel <- final
summary(fittedModel)
AIC(fittedModel)
simulated_residuals <- simulateResiduals(fittedModel)
plotQQunif(simulated_residuals)
plot(simulated_residuals)



```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
