---
title: "PCA"
author: "Tessa Rehill"
date: "2024-12-12"
output: html_document
---

#notes 
  - need to make PCA predictors ALL the same (including non-numeric predictor -> factors -> numeric), include ALLLLL 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}

library(dplyr)
library(vegan)
library(tidyverse)
library(ggplot2)
library(tidyr)

```

#load data 
```{r}

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
sites <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/raw_data/metadata/site_coordinates.csv")

data <- merge(data, sites, by = "site", all.x = TRUE) #merge site data with data 
data$density <- as.factor(data$density)


```

#Change values to NUMERIC 
```{r}

data$site <- as.factor(data$site)
data$site <- as.numeric(data$site)

data$plot_id <- as.factor(data$plot_id)
data$plot_id <- as.numeric(data$plot_id)

data$gravel_size <- as.factor(data$gravel_size)
data$gravel_size <- as.numeric(data$gravel_size)

data$treatment <- as.factor(data$treatment)
data$treatment <- as.numeric(data$treatment)

data$month <- as.factor(data$month)
data$month <- as.numeric(data$month)

data$density <- as.factor(data$density)
data$density <- as.numeric(data$density)
```


#RUN PCA w/ site 
```{r}

#prepare the data: Remove rows with NA values in both the numeric columns and site
data_clean <- data %>%
  select(urchin_density, site_density, kelp_growth, mean_temperature_c, 
         mean_light_lux, bottom_do, bottom_salinity, bottom_ph, gravel_size, treatment, mean_depth_m, species_richness, 
         shannon_div, simp_div, site, percent_occuring_algae, kelp_survival, month, density) %>%
  drop_na()  #drop rows with NA values in any of the selected columns


#perform PCA
pca <- prcomp(data_clean %>% select(-site), scale. = TRUE)  # Remove 'site' column for PCA

#create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$site <- data_clean$site  # Add 'site' from cleaned data

#plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(site))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Site", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Site")

```
#RUN PCA - w/ treatment 
```{r}
#prepare the data: Remove rows with NA values in both the numeric columns and site
data_clean <- data %>%
  select(urchin_density, site_density, kelp_growth, mean_temperature_c, 
         mean_light_lux, bottom_do, bottom_salinity, bottom_ph, gravel_size, treatment, mean_depth_m, species_richness, 
         shannon_div, simp_div, site, percent_occuring_algae, kelp_survival, month, density) %>%
  drop_na()  #drop rows with NA values in any of the selected columns

#perform PCA
pca <- prcomp(data_clean %>% select(-treatment), scale. = TRUE)  # Remove 'treatment' column for PCA

#create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$treatment <- data_clean$treatment  # Add 'site' from cleaned data

#plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(treatment))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Treatment", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Treatment")



```

#RUN PCA - w/ density 
```{r}

data_clean <- data %>%
  select(urchin_density, site_density, kelp_growth, mean_temperature_c, 
         mean_light_lux, bottom_do, bottom_salinity, bottom_ph, gravel_size, treatment, mean_depth_m, species_richness, 
         shannon_div, simp_div, site, percent_occuring_algae, kelp_survival, month, density) %>%
  drop_na()  #drop rows with NA values in any of the selected columns

#perform PCA
pca <- prcomp(data_clean %>% select(-density), scale. = TRUE)  # Remove 'density' column for PCA

#create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$density <- data_clean$density  

#plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(density))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Density", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Density")


```
#RUN PCA - w/ plot_id
```{r}
data_clean <- data %>%
  select(urchin_density, site_density, kelp_growth, mean_temperature_c, 
         mean_light_lux, bottom_do, bottom_salinity, bottom_ph, gravel_size, treatment, mean_depth_m, species_richness, 
         shannon_div, simp_div, site, percent_occuring_algae, kelp_survival, month, density, plot_id) %>%
  drop_na()  #drop rows with NA values in any of the selected columns

#perform PCA
pca <- prcomp(data_clean %>% select(-plot_id), scale. = TRUE)  # Remove 'density' column for PCA

#create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$plot_id <- data_clean$plot_id  

#plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(plot_id))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Plot_ID", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "plot_id")


```

#RUN PCA on density with character labels 
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
sites <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/raw_data/metadata/site_coordinates.csv")

data <- merge(data, sites, by = "site", all.x = TRUE) #merge site data with data 
data$density <- as.factor(data$density)

data$site <- as.factor(data$site)
data$site <- as.numeric(data$site)

data$plot_id <- as.factor(data$plot_id)
data$plot_id <- as.numeric(data$plot_id)

data$gravel_size <- as.factor(data$gravel_size)
data$gravel_size <- as.numeric(data$gravel_size)

data$treatment <- as.factor(data$treatment)
data$treatment <- as.numeric(data$treatment)

data$month <- as.factor(data$month)
data$month <- as.numeric(data$month)

data_clean <- data %>%
  select(urchin_density, site_density, kelp_growth, mean_temperature_c, 
         mean_light_lux, bottom_do, bottom_salinity, bottom_ph, gravel_size, treatment, mean_depth_m, species_richness, 
         shannon_div, simp_div, site, percent_occuring_algae, kelp_survival, month, density, plot_id) %>%
  drop_na()  #drop rows with NA values in any of the selected columns

#perform PCA
pca <- prcomp(data_clean %>% select(-density), scale. = TRUE)  # Remove 'density' column for PCA

#create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$density <- data_clean$density  

#plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(density))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by DENSITY CATEGORIESd", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "density")



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
