---
title: "Growth Models"
author: "Tessa Rehill"
date: "2024-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)


```

#Run GLMM (Tweedie distribution)

  #load data
```{r}

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```

  #considering overdispersion and zero inflation 
```{r}
#check for overdispersion 
mean_growth <- mean(data$kelp_growth, na.rm = TRUE)
var_growth <- var(data$kelp_growth, na.rm = TRUE)
overdispersion <- var_growth / mean_growth
print(overdispersion)  #661- high dispersion

#check for zero inflation 
sum(data$kelp_growth == 0, na.rm = TRUE) / nrow(data) #a lot of zeros (48%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

```

  #checking correlation of predictos 
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "urchin_density", "percent_occuring_algae", "site_density", "shannon_div", "simp_div", "species_richness")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#Compute VIF factor 
#with all predictors 
lm_model <- lm(kelp_growth ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m + bottom_temp + bottom_do + species_richness + 
               shannon_div + simp_div + site_density + urchin_density, data = data)

vif(lm_model) #so many >5

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/env/kelp_correlation_plot.png", plot = plot, width = 10, height = 8, dpi = 300)



```

  #scaling predictors 
```{r}


#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$urchin_density <- scale(data$urchin_density)
data$site_density <- scale(data$site_density)
data$simp_div <- scale(data$simp_div)
data$percent_occuring_algae <- scale(data$percent_occuring_algae)

#month needs to be turned back into a factor
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)

```

  #simplest model 
```{r}
# Fit the GLMM model
simple_model <- glmmTMB(
  kelp_growth ~  gravel_size + urchin_density + treatment + (1 | site) + (1 |plot_id) + (1 |month) ,
  data = data,
  family = tweedie()
)

summary(simple_model)
```

#Visualize 
```{r}
#simple model 
fittedModel <- simple_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput



```
#Forward featyre selection based on AIC  (w environmental predictors) 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(kelp_growth ~ 1 + (1 | site)  , 
                      family = tweedie(), 
                      data = data)

#predictors 
predictor_list <- c( "percent_occuring_algae", "treatment", "site_density", "urchin_density",
                "gravel_size",  "mean_temperature_c", "mean_light_lux", 
                   "bottom_ph", "simp_div", "species_richness")


#initialize variables to store the results
selected_predictors <- c()  # To keep track of selected predictors
current_model <- base_model  # Start with the base model
current_aic <- AIC(base_model)  # AIC of the base model
aic_results <- data.frame(Model = "Base Model", AIC = current_aic)

#forward feature selection loop
for (predictor in predictor_list) {
  # Try adding each predictor to the current model
  formula <- as.formula(
    paste("kelp_growth ~", paste(c(selected_predictors, predictor), collapse = " + "), 
          "+ (1 | site)")
  )
  
  #fit the model with the new predictor
  new_model <- glmmTMB(
    formula,
    family = tweedie(),
    data = data
  )
  
  #compute AIC
  new_aic <- AIC(new_model)
  
  #store results
  aic_results <- rbind(aic_results, data.frame(Model = paste("Add", predictor), AIC = new_aic))
  
  #check if the AIC improves
  if (new_aic < current_aic) {
    # Update the current model and AIC
    current_model <- new_model
    current_aic <- new_aic
    selected_predictors <- c(selected_predictors, predictor)  # Add predictor to the list
  }
}

#print results
print(aic_results)

current_model 

#final selected model
summary(current_model)



```
```{r}
lm_model <- lm(kelp_growth ~  gravel_size + mean_temperature_c + 
                bottom_ph  + simp_div + site_density, data = data)

vif(lm_model) #so many >5
```


#Visualize 
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/growth_models/testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/growth_models/QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/growth_models/residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()
```
#co-efficient plot 
#create coefficient plot 
```{r}
# Load required libraries
# Extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(current_model, conf.int = TRUE)

# Combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Filter out 'cond.sd__(Intercept)' from the data if applicable
coef_tidy <- coef_tidy %>%
  filter(term != "cond.sd__(Intercept)")



# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) + 
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(title = "Model Coefficients with Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Predictor",
       color = "Component")

# Print the plot
print(dw)

#right of zero = positive effect 
#left of zero = negative effect 


#sve the dot-and-whisker plot as a PNG
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/growth_models/dot_and_whisker_plot.png", plot = dw, width = 8, height = 6, dpi = 300)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
