---
title: "Survivor_Models_WORKFLOW"
author: "Tessa Rehill"
date: "2024-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)

```

#load data
```{r}
transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_monitoring_transects.csv")

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```

#Kelp survival as a proportion 
```{r}

selected_columns <- plots[, c("site", "month", "plot_id", "n_kelp_gravel", "n_gravel")]

# Step 2: Merge (add) the selected columns to 'data' by 'site' and 'month'
data <- merge(data, selected_columns, by = c("site", "month", "plot_id"), all.x = TRUE)


data$kelp_prop <- data$kelp_survival/100

#transform response variable with small constant 
data$kelp_prop_adj <- (data$kelp_prop * (nrow(data) - 1) + 0.5) / nrow(data)

# Create new columns for the success (kelp) and failure (non-kelp)
data$kelp_success <- data$n_kelp_gravel
data$kelp_failure <- data$n_gravel - data$n_kelp_gravel

summary(data)
```

#Considering overdispersion and zero inflation 
```{r}
#check for overdispersion 
mean_survival <- mean(data$kelp_prop, na.rm = TRUE)
var_survival <- var(data$kelp_prop, na.rm = TRUE)
overdispersion <- var_survival / mean_survival 
print(overdispersion)  #38 - high dispersion

mean_survival <- mean(data$kelp_prop_adj, na.rm = TRUE)
var_survival <- var(data$kelp_prop_adj, na.rm = TRUE)
overdispersion <- var_survival / mean_survival 
print(overdispersion)  #37 - high dispersion


#check for zero inflation 
sum(data$kelp_prop == 0, na.rm = TRUE) / nrow(plots) #a lot of zeros (55%), evidence of zero inflation
  #rooted out using zero inflation model - similar outputs to standard beta models, zero inflation component does not add explanatory power

sum(data$kelp_prop_adj == 0, na.rm = TRUE) / nrow(plots) #no zeros 

```

#Checking correlation of predictos 
```{r}
#CHECK CORRELATION OF ALL NUMERIC REDICTORS ############################################
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "bottom_do", "urchin_density", "percent_occuring_algae", "site_density", "shannon_div", "simp_div", "species_richness")], use = "complete.obs")

corrplot(cor_matrix, method = "circle")

#Compute VIF factor 
#with all predictors 
lm_model <- lm(kelp_prop_adj ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m + bottom_temp + bottom_do + species_richness + 
               shannon_div + simp_div + kelp_growth + site_density, data = data)

vif(lm_model) #so many >5

lm_model <- lm(kelp_prop_adj ~ mean_light_lux + mean_temperature_c + bottom_salinity + bottom_ph +
               mean_depth_m + bottom_temp + bottom_do + species_richness + 
               shannon_div + simp_div + kelp_growth + site_density, data = data)

vif(lm_model) #so many >5

#pairwise scatterplot matrix
data$kelp_prop_adj <- as.numeric(data$kelp_prop_adj)
data$mean_light_lux <- as.numeric(data$mean_light_lux)
data$urchin_density <- as.numeric(data$urchin_density)
data$simp_div <- as.numeric(data$simp_div)

#log density 
data$urchin_density <- log(data$urchin_density + 1)
data$site_density <- log(data$site_density + 1)

#ggpairs(data[, c("kelp_prop_adj", "mean_light_lux", "urchin_density", "simp_div", "mean_temperature_c", "mean_depth_m", 
                # "percent_occuring_algae", "site_density", "shannon_div", "species_richness", "bottom_do")])



```

#Scaling predictors 
```{r}

#standardizing predictors
data$mean_temperature_c <- scale(data$mean_temperature_c)
data$bottom_ph <- scale(data$bottom_ph)
data$bottom_salinity <- scale(data$bottom_salinity)
data$bottom_do <- scale(data$bottom_do)
data$mean_light_lux <- scale(data$mean_light_lux)
data$mean_depth_m <- scale(data$mean_depth_m)
data$urchin_density <- scale(data$urchin_density)
data$site_density <- scale(data$site_density)
data$simp_div <- scale(data$simp_div)
data$percent_occuring_algae <- scale(data$percent_occuring_algae)

#month needs to be turned back into a factor
data$site <- factor(data$site)
data$month <- factor(data$month)
data$treatment <- as.factor(data$treatment)
data$gravel_size <- as.factor(data$gravel_size)
data$plot_id <- as.factor(data$plot_id)
```

#Define predictors and base model structure 
```{r}
#define predictors and initial formula
predictors <- c( "percent_occuring_algae", "treatment", "site_density", "urchin_density",
                "gravel_size",  "mean_temperature_c", "mean_light_lux", 
                "bottom_do", "bottom_temp", "bottom_salinity", "mean_depth_m", 
                 "shannon_div", "month")


#initialize the empty model formula (with no predictors)
base_formula <- "kelp_prop_adj ~ 1 + (1 | site) + (1 | plot_id)"  # Initial random effects
```

#Beta regression GLM - forward feature selection
```{r}

#predictors: site, month, plot_id, percent_occuring_algae, treatment, gravel_size, urchin_density, mean_temperature_c, mean_light_lux
#bottom_do, bottom_temp, bottom_salinity, mean_depth>m, site_density, species_richness, shannon_div, simp_div

#BETA structure
#beta_model <- glmmTMB(
#  kelp_prop_adj ~ predictor 1, predictor 2 + (1 | site), 
#  data = data,
#  family = beta_family(link = "logit")
#)


# Set up a function to perform forward feature selection
forward_feature_selection <- function(predictors, base_formula, data) {
  
  #initialize the best model to the base model
  best_model <- glmmTMB(as.formula(base_formula), data = data, family = beta_family(link = "logit"))
  best_aic <- AIC(best_model)
  best_formula <- base_formula
  selected_predictors <- character(0)  # No predictors selected initially
  
  #loop through the remaining predictors and add one at a time
  while (length(predictors) > 0) {
    
    #try adding each remaining predictor to the model
    candidate_aic <- numeric(length(predictors))
    for (i in 1:length(predictors)) {
      
      #create the new formula with the current predictor
      new_formula <- paste(best_formula, "+", predictors[i])
      
      #fit the new model
      model <- glmmTMB(as.formula(new_formula), data = data, family = beta_family(link = "logit"))
      
      #calculate the AIC of the new model
      candidate_aic[i] <- AIC(model)
    }
    
    #identify the model with the lowest AIC
    best_new_predictor <- predictors[which.min(candidate_aic)]
    if (min(candidate_aic) < best_aic) {
      
      #update the best model and the formula
      best_aic <- min(candidate_aic)
      best_formula <- paste(best_formula, "+", best_new_predictor)
      
      #add the best predictor to the selected predictors
      selected_predictors <- c(selected_predictors, best_new_predictor)
      
      #remove the selected predictor from the list of remaining predictors
      predictors <- setdiff(predictors, best_new_predictor)
      
      #print the updated model and AIC
      print(paste("Selected predictor:", best_new_predictor, "Current AIC:", best_aic))
    } else {
      #if no improvement, stop the selection process
      break
    }
  }
  
  #return the final model and selected predictors
  final_model <- glmmTMB(as.formula(best_formula), data = data, family = beta_family(link = "logit"))
  return(list(final_model = final_model, selected_predictors = selected_predictors, best_aic = best_aic))
}

# Run forward feature selection
result <- forward_feature_selection(predictors, base_formula, data)

# View the selected predictors and the final model's AIC
print(paste("Selected predictors:", paste(result$selected_predictors, collapse = ", ")))
print(paste("Final model AIC:", result$best_aic))

# View the summary of the final model
summary(result$final_model)


```

#Updated forward feature selection - based on AIC and REMSC
```{r}

# Modified forward feature selection function
forward_feature_selection <- function(predictors, base_formula, data, k = 5) {
  
  # Initialize the best model to the base model
  best_model <- glmmTMB(as.formula(base_formula), data = data, family = beta_family(link = "logit"))
  best_aic <- AIC(best_model)
  best_bic <- BIC(best_model)
  best_cv_rmse <- NA  # Cross-validated RMSE (initialized as NA)
  best_formula <- base_formula
  selected_predictors <- character(0)  # No predictors selected initially
  
  # Perform k-fold cross-validation to calculate RMSE for the given model
  calculate_cv_rmse <- function(model_formula, data, k) {
    folds <- createFolds(data$kelp_prop_adj, k = k, list = TRUE)
    rmse_values <- c()
    
    for (fold in folds) {
      train_data <- data[-fold, ]
      test_data <- data[fold, ]
      
      model <- glmmTMB(as.formula(model_formula), data = train_data, family = beta_family(link = "logit"))
      predictions <- predict(model, newdata = test_data, type = "response")
      rmse_values <- c(rmse_values, RMSE(predictions, test_data$kelp_prop_adj))
    }
    mean(rmse_values)
  }
  
  # Loop through the remaining predictors and add one at a time
  while (length(predictors) > 0) {
    
    candidate_metrics <- data.frame(predictor = predictors, aic = NA, bic = NA, cv_rmse = NA)
    
    for (i in 1:length(predictors)) {
      # Create the new formula with the current predictor
      new_formula <- paste(best_formula, "+", predictors[i])
      
      # Fit the new model
      model <- glmmTMB(as.formula(new_formula), data = data, family = beta_family(link = "logit"))
      
      # Calculate metrics
      candidate_metrics$aic[i] <- AIC(model)
      candidate_metrics$bic[i] <- BIC(model)
      candidate_metrics$cv_rmse[i] <- calculate_cv_rmse(new_formula, data, k)
    }
    
    # Select the predictor that optimizes a combination of metrics
    metric_ranks <- candidate_metrics %>%
      mutate(rank_aic = rank(aic),
             rank_bic = rank(bic),
             rank_rmse = rank(cv_rmse)) %>%
      rowwise() %>%
      mutate(combined_rank = mean(c(rank_aic, rank_bic, rank_rmse))) %>%
      arrange(combined_rank)
    
    best_new_predictor <- metric_ranks$predictor[1]
    best_new_aic <- metric_ranks$aic[1]
    best_new_bic <- metric_ranks$bic[1]
    best_new_rmse <- metric_ranks$cv_rmse[1]
    
    # Check if the new model improves based on AIC or RMSE
    if (best_new_aic < best_aic) {
      # Update the best model and the formula
      best_aic <- best_new_aic
      best_bic <- best_new_bic
      best_cv_rmse <- best_new_rmse
      best_formula <- paste(best_formula, "+", best_new_predictor)
      
      # Add the best predictor to the selected predictors
      selected_predictors <- c(selected_predictors, best_new_predictor)
      
      # Remove the selected predictor from the list of remaining predictors
      predictors <- setdiff(predictors, best_new_predictor)
      
      # Print the updated model and metrics
      print(paste("Selected predictor:", best_new_predictor, 
                  "AIC:", round(best_aic, 2), 
                  "BIC:", round(best_bic, 2), 
                  "CV RMSE:", round(best_cv_rmse, 2)))
    } else {
      # If no improvement, stop the selection process
      break
    }
  }
  
  # Return the final model and selected predictors
  final_model <- glmmTMB(as.formula(best_formula), data = data, family = beta_family(link = "logit"))
  return(list(final_model = final_model, 
              selected_predictors = selected_predictors, 
              best_aic = best_aic, 
              best_bic = best_bic, 
              best_cv_rmse = best_cv_rmse))
}

# Run the updated forward feature selection
result <- forward_feature_selection(predictors, base_formula, data)

# View the selected predictors and the final model's metrics
print(paste("Selected predictors:", paste(result$selected_predictors, collapse = ", ")))
print(paste("Final model AIC:", result$best_aic))
print(paste("Final model BIC:", result$best_bic))
print(paste("Final model CV RMSE:", result$best_cv_rmse))

# View the summary of the final model
summary(result$final_model)

```
#Best model 
```{r}
beta_model <- glmmTMB(
  kelp_prop_adj ~ gravel_size + mean_temperature_c + (1 | site) + (1 | plot_id) + (1 | month) , 
  data = data,
  family = beta_family(link = "logit")
)

```

Coefficient plot 
```{r}
# Load required libraries
library(broom.mixed)
library(dotwhisker)
library(ggplot2)

# Extract coefficients in tidy format
coef_tidy <- broom.mixed::tidy(final_model, conf.int = TRUE)

# Combine component and term for clarity
coef_tidy <- coef_tidy %>%
  mutate(term = paste(component, term, sep = "."))

# Add the 'model' column
coef_tidy$model <- "Model 1"

# Create the dot-and-whisker plot
dw <- dwplot(coef_tidy, conf.level = 0.95) + 
  geom_vline(xintercept = 0, linetype = "dashed") +  # Add reference line at 0
  theme_minimal() +
  labs(title = "Model Coefficients with Confidence Intervals",
       x = "Coefficient Estimate",
       y = "Predictor",
       color = "Component")

# Print the plot
print(dw)
```


#Visualize 
```{r}


#final model
fittedModel <- beta_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
