---
title: "GLM"
author: "Tessa Rehill"
date: "2025-01-20"
output: html_document
---

Run GLMs on relevant response variables of interest 

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries
```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(DHARMa)
library(glmmTMB)
library(car)
```

#Directory info
```{r}
wd=getwd()

username=as.vector(strsplit(wd,"/"));names(username)<-c("username")

data_dir = as.character(paste("C:/Users/", username$username[3], "/Dropbox/MSc/Exclosures", sep=""))

```

#Read in data
```{r}

data <- read.csv("../processed_data/cleaned_data.csv")

```

#Lauren's example code

glm.results.first.4 <- glmer(cbind(number_gravels_with_kelp, number_gravels_remaining - number_gravels_with_kelp) ~ parent_number +     source_population * site + (1|site_replicate), data.first, family = binomial)

#Let's test the predictors that were selected first 
#as single fixed and random effects, no stepwise selection
```{r}

#temperature only 
glm_result <- glmer(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ mean_temperature_c + (1|site), data, family = binomial)

summary(glm_result)
#AIC = 1385.7

#evaluate visually 

testDispersion(glm_result) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = glm_result, plot = T)
simulationOutput

```
#Evaluating models visualy 
```{r}
fittedModel <- glm_result

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```
#Using stepwise predictor selection 
  #base model + predictors 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ 1  +(1 | site), 
                      family = binomial, 
                      data = data)

#predictors 
predictor_list <- c("percent_occuring_algae", "treatment", "site_density", "urchin_density", "grazer_density",
                    "gravel_size", "mean_temperature_c", "mean_light_lux", "bottom_ph", "bottom_do", "bottom_salinity")

```

  
#MODEL ALGORITHM - forward and backward stepwise selection
```{r}

#initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

#stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~", 
            paste(remaining_predictors, collapse = " + "), 
            "+ (1 | site)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

#print results
print(selected_predictors)
summary(current_model)

```
#Visualize 
```{r}
#final model
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```
#Check collinearity 
```{r}
lm_model <- lm(n_gravel ~  treatment + gravel_size + 
                  mean_temperature_c + bottom_ph + bottom_salinity,  data = data)

vif(lm_model) #all below 5

lm_model <- lm(n_kelp_gravel ~  treatment + gravel_size + 
                  mean_temperature_c + bottom_ph + bottom_salinity,  data = data)

vif(lm_model) #all below 5


lm_model <- lm(kelp_survival ~  treatment + gravel_size + 
                  mean_temperature_c + bottom_ph + bottom_salinity,  data = data)

vif(lm_model) #all below 5

```
```{r}
glm_result <- glmer(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment + gravel_size + mean_temperature_c + bottom_ph + bottom_salinity + (1|site), data, family = binomial)

summary(glm_result)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
