---
title: "GLMMs"
author: "Tessa Rehill"
date: "2025-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 

```{r}
library(here)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)
library(GGally)
library(broom.mixed)
library(dotwhisker)
library(ggplot2)
library(broom)
library(lme4)
library(lmerTest)
```

#Load data 
```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
```

#Correlation among ABIOTIC predictors
```{r}
cor_matrix <- cor(data[, c("mean_light_lux", "mean_temperature_c", "max_temperature_c",  "mean_current", "max_current", "bottom_salinity",  "bottom_do", "bottom_ph" )], use = "complete.obs")

#rename variables
# Rename row and column names
colnames(cor_matrix) <- rownames(cor_matrix) <- c("Light (lux)", "Temp (mean)", "Temp (max)", 
                                                  "Current (mean)", "Current (max)", "Bottom salinity", "Bottom DO", 
                                                  "Bottom pH")

#create the correlation plot
plot <- ggcorrplot(
  cor_matrix, 
  method = "circle",   #use circles for visualization
  lab = TRUE           #add correlation coefficients as labels
)
plot

#save the plot
ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/abiotic_correlation_plot.png", plot = plot, width = 10, height = 8, dpi = 300)

```

# Investigating the effects of experimental variables on kelp survival and kelp growth 
##Kelp survival
```{r}
### What experimental variables are we interested in: treatment, gravel_size, urchin_density, urchin_site_density, ts_density, ts_site_density
### Also try using total grazer densities (urchin + ts)

survival_urchin_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment  + gravel_size + 
                                                                         urchin_density*urchin_site_density + 
                                                                         ts_density*ts_site_density + (1 | site) + (1 | date), 
                      family = binomial, 
                      data = data)

summary(survival_urchin_model)

survival_grazer_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ treatment  + gravel_size + 
                                                                         grazer_density * grazers_site_density + 
                                                                         (1 | site) + (1 | date), 
                      family = binomial, 
                      data = data)

summary(survival_grazer_model)


```
###Visualize with Dharma
```{r}
fittedModel <- survival_grazer_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

fittedModel <- survival_urchin_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput


qqnorm(residuals(survival_grazer_model))
qqline(residuals(survival_grazer_model))

qqnorm(residuals(survival_urchin_model))
qqline(residuals(survival_urchin_model))

```

##Growth models 
```{r}
#remove growth = 0 from dataset 
growth_data <- data %>% drop_na(kelp_growth)

growth_urchin_model <- glmmTMB(kelp_growth ~ treatment  + gravel_size + 
                                      urchin_density*urchin_site_density + 
                                      ts_density*ts_site_density + (1 | site) + (1 | date), 
                      family = Gamma(link="log"), 
                      data = growth_data)

summary(growth_urchin_model)

growth_grazer_model <- glmmTMB(kelp_growth ~ treatment  + gravel_size + 
                                        grazer_density * grazers_site_density + 
                                        (1 | site) + (1 | date), 
                      family = Gamma(link="log"), 
                      data = data)

summary(growth_grazer_model)

```
###Visualize with Dharma
```{r}

fittedModel <- growth_urchin_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

fittedModel <- growth_grazer_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput
```

#Investigating the abiotic variables for kelp survival and growth 
##Kelp survival
###base model + predictor 
```{r}
#starting with a minimal model (only random effects)
base_model <- glmmTMB(cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~ 1  + (1 | site) + (1 | date), 
                      family = binomial, 
                      data = data)

#predictors 
#predictor_list <- c("percent_occuring_algae", "treatment", "urchin_site_density*urchin_density",  "max_temperature_c", "max_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity","ts_density*ts_site_density" )

#predictor_list<- c("percent_occuring_algae", "treatment", "grazers_site_density*grazer_density",  "mean_temperature_c", "max_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity" )

predictor_list<- c("percent_occuring_algae", "treatment", "urchin_site_density*urchin_density",  "ts_site_density*ts_density","mean_temperature_c", "max_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity")
###
#predictors 
predictor_list<- c("percent_occuring_algae", "treatment", "urchin_site_density*urchin_density",  "ts_site_density*ts_density","mean_temperature_c", "max_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity", "max_temperature_c")

#
predictor_list<- c("percent_occuring_algae", "treatment", "urchin_site_density*urchin_density", "grazers_site_density*grazer_density", "ts_site_density*ts_density","mean_temperature_c", "mean_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity", "max_temperature_c")

predictor_list<- c("treatment", "urchin_site_density*urchin_density", "grazers_site_density*grazer_density", "ts_site_density*ts_density","mean_temperature_c", "mean_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity", "max_temperature_c")

```

###forward + backward predictor selection based on AIC 
```{r}
#initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

#stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site) + (1 | date)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("cbind(n_kelp_gravel, n_gravel - n_kelp_gravel) ~", 
            paste(remaining_predictors, collapse = " + "), 
            "+ (1 | site) + (1 | date)")
    )
    new_model <- glmmTMB(formula, family = binomial, data = data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

#print results
print(selected_predictors)
summary(current_model)


``` 

#visualize (+ save)
```{r}

#final model
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()

#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/beta_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```

  #collinearity 
```{r}
lm_model <- lm(n_gravel ~ grazers_site_density*grazer_density + gravel_size + mean_temperature_c + max_current + bottom_do, data = data)

vif(lm_model, type="predictor") 


```
```{r}
kelp_survival_model <- current_model
```


##Growth models 
  #base model + predictors 
```{r}


#starting with a minimal model (only random effects)
base_model <- glmmTMB(kelp_growth ~ 1 + (1 | site) + (1 | date), 
                      family = Gamma(link="log"), 
                      data = growth_data)


#predictors 
predictor_list<- c("percent_occuring_algae", "treatment", "grazers_site_density*grazer_density",  "mean_temperature_c", "max_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity" )

predictor_list<- c("percent_occuring_algae", "treatment", "urchin_site_density*urchin_density",  "ts_site_density*ts_density","mean_temperature_c", "max_current","gravel_size", "mean_light_lux", "bottom_do",   "bottom_salinity" )

```

  #chose model based on  forward and backward feature seleciton
```{r}
# Initialize variables
selected_predictors <- c()
current_model <- base_model
current_aic <- AIC(base_model)
converged <- FALSE

# Stepwise selection loop
while (!converged) {
  converged <- TRUE
  
  # Forward selection
  for (predictor in setdiff(predictor_list, selected_predictors)) {
    formula <- as.formula(
      paste("kelp_growth ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site) + (1 | date)")
    )
    new_model <- glmmTMB(formula, family = Gamma(link="log"), data = growth_data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- c(selected_predictors, predictor)
      converged <- FALSE
    }
  }
  
  # Backward elimination
  for (predictor in selected_predictors) {
    remaining_predictors <- setdiff(selected_predictors, predictor)
    formula <- as.formula(
      paste("kelp_growth ~", 
            paste(c(selected_predictors, predictor), collapse = " + "), 
            "+ (1 | site) + (1 | date)")
    )
    new_model <- glmmTMB(formula, family = Gamma(link="log"), data = growth_data)
    new_aic <- AIC(new_model)
    
    if (new_aic < current_aic) {
      current_model <- new_model
      current_aic <- new_aic
      selected_predictors <- remaining_predictors
      converged <- FALSE
    }
  }
}

# Print results
print(selected_predictors)
summary(current_model)

```

```{r}
lm_model <- lm(kelp_growth ~ grazers_site_density*grazer_density + mean_temperature_c + max_current + treatment, data = growth_data)

vif(lm_model, type="predictor") 
```

  #visualize 
```{r}
#simple model 
fittedModel <- current_model

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput


#save the testDispersion plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_testDispersion_plot.png", width = 800, height = 600)
testDispersion(fittedModel)
dev.off()


#save the QQ plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_QQ_plot.png", width = 800, height = 600)
plotQQunif(simulationOutput)  # Generate QQ plot
dev.off()

#save residual vs predictor plot
png("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/growth_plot_residuals_vs_predictor.png", width = 800, height = 600)
plotResiduals(simulationOutput)
dev.off()

```
```{r}
kelp_growth_model <- current_model
```

#PLOT ALL COEFFICIENT FIGURES 
```{r}
# Extract coefficients and confidence intervals for each model
kelp_model_tidy <- tidy(kelp_survival_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Survival GLMM")

growth_model_tidy <- tidy(kelp_growth_model, conf.int = TRUE) %>%
  mutate(model = "Kelp Growth GLMM")



# Combine all models into one data frame
all_models <- bind_rows(kelp_model_tidy, growth_model_tidy)

# Filter out random effects (if you only want fixed effects)
all_models <- all_models %>%
  filter(effect == "fixed" | is.na(effect))  # GLM doesn't have an 'effect' column, so allow NA

# Define the color scheme for each model
model_colors <- c(
  "Kelp Survival GLMM" = "#eacf62",
  "Kelp Growth GLMM" = "#FF5733"
)

# Define the desired order of predictors in reverse
predictor_order <- rev(c('treatmentopen', 'gravel_sizesmall', 'urchin_site_density','urchin_density','ts_density', "ts_site_density", 
                         "urchin_site_density:urchin_density", 'ts_site_density:ts_density',
                          'max_current', 'mean_temperature_c',  
                        'percent_occuring_algae', '(Intercept)'))

predictor_order <- rev(c('treatmentopen', 'gravel_sizesmall', 'grazers_site_density','grazer_density', 'grazers_site_density:grazer_density',
                          'max_current', 'mean_temperature_c',  
                        'percent_occuring_algae', '(Intercept)'))

# Modify the ggplot code to reorder the terms and increase font sizes
plot <- ggplot(all_models, aes(x = factor(term, levels = predictor_order),  
                               y = estimate, ymin = conf.low, ymax = conf.high,  
                               color = model)) +    
  geom_pointrange(position = position_dodge(width = 0.5)) +    
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +    
  coord_flip() +    
  scale_color_manual(values = model_colors) +    
  labs(
    title = "Coefficient and Confidence Interval Plot",
    x = "Predictors",
    y = "Estimate",
    color = "Model"
  ) +    
  theme_minimal() +
  theme(
    text = element_text(size = 18),         # Base text size
    axis.text = element_text(size = 16),    # Axis tick labels
    axis.title = element_text(size = 18),   # Axis titles
    legend.text = element_text(size = 16),  # Legend text
    legend.title = element_text(size = 18), # Legend title
    plot.title = element_text(size = 20, face = "bold") # Plot title
  )

plot

ggsave("/Users/tessarehill/Dropbox/MSc/Exclosures/plots/models/combined_dot_and_whisker_plot.png", 
       plot = plot, width = 15, height = 10, dpi = 300)

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
