---
title: "Kelp_Models"
author: "Tessa Rehill"
date: "2024-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries 
```{r}
library(here)
library(dplyr)
library(ggplot2)
library(corrplot)
library(car)
library(glmmTMB)
library(DHARMa)
library(caret)

```

#Read in files 
```{r}
transects <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_monitoring_transects.csv")

plots <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_green_gravel_monitoring.csv")

env <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_ysi_measurement.csv")

hobo <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/ENV/HOBO/summary_hobo.csv")

dens <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/density_urchin_transects.csv")
```


#Add site density to plots dataframe (to be used as a predictor in our models)
#And environmental variables (from hobo and env)
```{r}
#change column name urchin_density in dens to site_density
dens <- dens %>% rename(site_density = urchin_density) #describes urchin density at the site from transects 

#merge dens into plots by site and month
plots_glm <- merge(plots, dens[, c("site", "month", "mean_depth_m", "site_density")], 
               by = c("site", "month"), 
               all.x = TRUE)

#create env summary, for bottom temperature, do, ph, salinity
bottom_data <- env %>%
 group_by(site, month) %>%
  summarize(
    bottom_temp = temp[which.max(depth_m)],
    bottom_do = do[which.max(depth_m)],
    bottom_salinity = salinity[which.max(depth_m)],
    bottom_ph = ph[which.max(depth_m)],
    .groups = "drop" # Ungroups after summarization
  )

#make month a character
bottom_data <- bottom_data %>%
  mutate(month = as.character(month))

#merge environmental data and depth data by site/month to plot data 
plots_glm <- plots_glm %>%
  left_join(hobo, by = c("site", "month")) %>%
  left_join(bottom_data, by = c("site", "month"))

#select columns of interest for our models 
selected_columns <- c("site", "month", "plot_id", "date", "treatment", "gravel_size", "kelp_survival", "urchin_density", "kelp_growth", "mean_depth_m", "site_density", "mean_temperature_c", "mean_light_lux", "bottom_do", "bottom_salinity", "percent_occuring_algae")

model_df <- plots_glm[, selected_columns]

```

#Zero-inflated generalized linear mixed-effects models - SURVIVAL
  #zeros are due to natural processes 
```{r}
summary(model_df)

model_df$log_kelp_growth <- log(model_df$kelp_growth + 1)

#change month to integer 
  #convert 'month' to a factor
    model_df$month <- as.factor(model_df$month)

  #convert the factor levels to numeric
    model_df$month <- as.numeric(model_df$month)
    
#change plot ID to an integer 
      #convert 'month' to a factor
    model_df$plot_id <- as.factor(model_df$plot_id)

  #convert the factor levels to numeric
    model_df$plot_id <- as.numeric(model_df$plot_id)
    
#CHECK CORRELATION OF ALL PREDICTORS ############################################
cor_matrix <- cor(model_df[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "month","bottom_do", "urchin_density", "percent_occuring_algae", "site_density", "urchin_density", "plot_id")],  use = "complete.obs")

corrplot(cor_matrix, method = "circle")


#CHECK ENVIRONMENTAL PREDICTORS #################################################
#correlation matrix of predictors 
cor_matrix <- cor(model_df[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "month","bottom_do", "urchin_density", "percent_occuring_algae", "site_density", "urchin_density")],  use = "complete.obs")

print(cor_matrix)

corrplot(cor_matrix, method = "circle")

#check VIF factor 
lm_model <- lm(kelp_survival ~ mean_light_lux + percent_occuring_algae + 
               mean_temperature_c + bottom_salinity + mean_depth_m + 
               bottom_do + month + urchin_density + site_density + plot_id, data = model_df)

vif(lm_model) #salinity, bottom_do, mean_temp, is too high 

#remove salinity as a predictor variable 
lm_model <- lm(kelp_survival ~ mean_light_lux + percent_occuring_algae + mean_temperature_c + month + mean_depth_m + bottom_do + month + urchin_density + site_density, data = model_df)

vif(lm_model) #all below 5 

#standardizing predictors
model_df$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df$bottom_do <- scale(model_df$bottom_do)
model_df$mean_light_lux <- scale(model_df$mean_light_lux)
model_df$mean_depth_m <- scale(model_df$mean_depth_m)
model_df$urchin_density <- scale(model_df$urchin_density)
model_df$site_density <- scale(model_df$site_density)

#month needs to be turned back into a factor
model_df$month <- factor(model_df$month)
model_df$treatment <- as.factor(model_df$treatment)
model_df$gravel_size <- as.factor(model_df$gravel_size)

```


#Models - kelp survival 
  #kelp survival represented as a DISCRETE variable, calculated using count data (number        of kelps)
    #use poisson distribution

#Generalized Linear Model - Zero Inflation - Poisson 
```{r}
#for the first model, include all predictors 
#turn kelp survival into a integer 
model_df$kelp_survival <- as.integer(model_df$kelp_survival)

#base model (everything included)
model1 <- glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + bottom_do + mean_light_lux +
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model1)
AIC(model1) #1380


#checking random effects 
model2 <- glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + bottom_do + mean_light_lux + month +
                       urchin_density + site_density + (1 | site),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model2)
AIC(model2) #1459 - we should include plot_id



#checking fixed effects: treatment, month, gravel_size, mean temp, mean depth, bottom do, light lux, 
#urchin site densities, urchin plot densities 

#no month 
model3 <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + bottom_do + mean_light_lux +
                       urchin_density + site_density + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
summary(model3)
AIC(model3) #1436, worse than base model with no month, keep month 

#no site density 
model4 <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + bottom_do + mean_light_lux +
                       urchin_density  + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
summary(model4)
AIC(model4) #1440, worse, remove site density 

#no urchin either density
model5 <-   glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + bottom_do + mean_light_lux +
                       + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model5)
AIC(model5) #1405, lowest so far - but we WANT to test for urchin densities (so keep?)

#remove light lux 
model6 <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + bottom_do + 
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)

summary(model6)
AIC(model6) #1369.385  #best so far 

#remove bottom do
model7 <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + mean_depth_m + mean_light_lux +
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
summary(model7)
AIC(model7) #1361

#remove depth
model8 <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + bottom_do + mean_light_lux +
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model8)
AIC(model8) #1425, worse, remove depth 

#remove temp
model9 <-   glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_depth_m + bottom_do + mean_light_lux +
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model9)
AIC(model9) #1399 - winner as of now! 

#remove gravel_size 
model10 <- glmmTMB(kelp_survival ~ treatment +
                       mean_temperature_c + mean_depth_m + bottom_do + mean_light_lux +
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model10)
AIC(model10) #1375 - even better??

#no ENV variables involed 
model11 <- glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(model11)
AIC(model11) #1480 #not better



AIC <- AIC(model1, model2, model3, model4, model5, model6, model7, model8, model9, model10, model11)

#lowest AIC models WITH valid model (based on CV)
  #model 8 - AIC 1425


#FINAL MODEL BASED ON LOWEST AIC AND MODEL CV 
final  <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       mean_temperature_c + bottom_do + mean_light_lux +
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)

summary(final)
AIC(final) #1425

final_simple <-  glmmTMB(kelp_survival ~ treatment + gravel_size + 
                       urchin_density + site_density + month + (1 | site) + (1 | plot_id),
                       ziformula = ~ 1,
                       family = poisson(), 
                       data = model_df)
 
summary(final_simple)
AIC(final_simple) #1490

```

#Zero Inflated Negative Binomial GLMM 
```{r}
#include ALL predictors in first model 
model1 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) + (1 | plot_id) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model1)
AIC(model1) #943

#check rnadom effects 
model2 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model2)
AIC(model2) #941 - remove plot_id

model3 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) + (1 | month) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do,
                  family = nbinom2, 
                  data = model_df)
summary(model3)
AIC(model3) #942 - month stays as fixed effect 

#change combinations of fixed effects 
model4 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do,
                  family = nbinom2, 
                  data = model_df)
summary(model4)
AIC(model4) #943 - month stays 

model5 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density  + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density  + month,
                  family = nbinom2, 
                  data = model_df)
summary(model5)
AIC(model5) #943 - do stays 

model6 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density  + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux  + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model6)
AIC(model6) #939 - urchin density makes a better model, but its something we're interested in

model7 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c  + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model7)
AIC(model7) #940 - site density stays 

model8 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux  + site_density + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment  + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model8)
AIC(model8) #941 - temperature stays

model9 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                       urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model9)
AIC(model9) #928 - light removed? 

model10 <- glmmTMB(kelp_survival ~ treatment + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density  + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model10)
AIC(model10) #941 - gravel size stays 

model11 <- glmmTMB(kelp_survival ~  + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~  + mean_temperature_c + site_density  + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model11)
AIC(model11) #941 - treatment stays 

#Remove light lux and urchin density 
model12 <- glmmTMB(kelp_survival ~ treatment + gravel_size  + mean_temperature_c + site_density  + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        bottom_do + month,
                  family = nbinom2, 
                  data = model_df)
summary(model12)
AIC(model12) #935- lowest AIC 

#Testing ONLY things we are testing (irrelevant of environment)

model13 <- glmmTMB(kelp_survival ~ treatment + gravel_size + site_density + urchin_density + month + (1 | site),
                  zi = ~ treatment + site_density + gravel_size + month,
                  family = nbinom2, 
                  data = model_df)

summary(model13)
AIC(model13) #937 - remove plot_id

AIC <- AIC(model1, model2, model3, model4, model5, model6, model7, model8, model9, model10, model11, model12, model13)
AIC


#I want to CV test 
  #model 13
  #model 2

final2 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  (1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month,
                  family = nbinom2, 
                  data = model_df)

summary(final2)
  
simple_final2 <- glmmTMB(kelp_survival ~ treatment + gravel_size + site_density + urchin_density + month + (1 | site),
                  zi = ~ treatment + site_density + gravel_size + month,
                  family = nbinom2, 
                  data = model_df)

summary(simple_final2)

```

#Validation of models

#using DHARMA 
```{r}
library(DHARMa)

#final model - poisson
fittedModel <- final

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#simple final model - poisson
fittedModel <- simple_final2

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput

#Negative binomial 
fittedModel <- final2

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput


fittedModel <- simple_final2

testDispersion(fittedModel) #check for overdispersion

simulationOutput <- simulateResiduals(fittedModel = fittedModel, plot = T)
simulationOutput



```

#Other model parameters
```{r}
# Load necessary libraries
library(DHARMa)
library(caret)


# 1. Residual Diagnostics for both models
# Residuals vs Fitted plot for final model (Poisson)
plot(fitted(final), residuals(final), main = "Residuals vs Fitted (final)", 
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")


# Residuals vs Fitted plot for final2 model (Negative Binomial)
plot(fitted(final2), residuals(final2), main = "Residuals vs Fitted (final2)", 
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

# 2. Zero-Inflation Check
# Predicted probabilities of zero inflation for the final model
pred_zeros_final <- predict(final, type = "response", se.fit = TRUE)
plot(pred_zeros_final$fit, type = "l", main = "Predicted Zero Probabilities (final)", 
     xlab = "Observation", ylab = "Predicted Probability of Zero")

# Predicted probabilities of zero inflation for the final2 model
pred_zeros_final2 <- predict(final2, type = "response", se.fit = TRUE)
plot(pred_zeros_final2$fit, type = "l", main = "Predicted Zero Probabilities (final2)", 
     xlab = "Observation", ylab = "Predicted Probability of Zero")

# 3. Model Comparison (AIC)
cat("AIC for final model (Poisson):", AIC(final), "\n")
cat("AIC for final2 model (Negative Binomial):", AIC(final2), "\n")

# 4. Dispersion Check for Overdispersion
# For final model (Poisson)
simulationOutput_final <- simulateResiduals(fittedModel = final, plot = TRUE)
testDispersion(simulationOutput_final)  # Check for overdispersion

# For final2 model (Negative Binomial)
simulationOutput_final2 <- simulateResiduals(fittedModel = final2, plot = TRUE)
testDispersion(simulationOutput_final2)  # Check for overdispersion

# 5. Check Convergence
# Check model summaries to see if they converged properly
summary(final)
summary(final2)

# 6. Fitted vs Observed Plot
# Fitted vs. Observed plot for final model (Poisson)
fitted_values_final <- fitted(final)
plot(model_df$kelp_survival, fitted_values_final, 
     main = "Observed vs Fitted (final)", xlab = "Observed", ylab = "Fitted")
abline(a = 0, b = 1, col = "red")  # 45-degree line

# Fitted vs. Observed plot for final2 model (Negative Binomial)
fitted_values_final2 <- fitted(final2)
plot(model_df$kelp_survival, fitted_values_final2, 
     main = "Observed vs Fitted (final2)", xlab = "Observed", ylab = "Fitted")
abline(a = 0, b = 1, col = "red")  # 45-degree line

# 7. Cross-validation (optional)
set.seed(123)

# Split data into training and testing sets
trainIndex <- createDataPartition(model_df$kelp_survival, p = 0.8, list = FALSE)
trainData <- model_df[trainIndex, ]
testData <- model_df[-trainIndex, ]

# Train both models on training data
model_final_train <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_temperature_c + 
                            mean_depth_m + bottom_do + mean_light_lux + urchin_density + 
                            site_density + month + (1 | site) + (1 | plot_id),
                            ziformula = ~1, family = poisson(), data = trainData)

model_final2_train <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_temperature_c + 
                             mean_depth_m + bottom_do + mean_light_lux + urchin_density + 
                             site_density + month + (1 | site) + (1 | plot_id),
                             ziformula = ~1, family = nbinom2(), data = trainData)

# Predict on test data
predictions_final <- predict(model_final_train, newdata = testData)
predictions_final2 <- predict(model_final2_train, newdata = testData)

# Calculate RMSE (Root Mean Squared Error)
RMSE_final <- sqrt(mean((predictions_final - testData$kelp_survival)^2))
RMSE_final2 <- sqrt(mean((predictions_final2 - testData$kelp_survival)^2))

cat("RMSE for final model (Poisson):", RMSE_final, "\n")
cat("RMSE for final2 model (Negative Binomial):", RMSE_final2, "\n")

# 8. Model Comparison (Likelihood Ratio Test)
anova(final, final2)  # Compare models using likelihood ratio test


fittedModel <- final2
summary(fittedModel)
AIC(fittedModel)
simulated_residuals <- simulateResiduals(fittedModel)
plotQQunif(simulated_residuals)
plot(simulated_residuals)

#ran this script comparing final2 and simple_final2 and was similar results 

```


#Final 2 is the best model 


```{r}
data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/data.csv")

model_df <- data 

#change month to integer 
  #convert 'month' to a factor
    model_df$month <- as.factor(model_df$month)

  #convert the factor levels to numeric
    model_df$month <- as.numeric(model_df$month)
    
#change plot ID to an integer 
      #convert 'month' to a factor
    model_df$plot_id <- as.factor(model_df$plot_id)

  #convert the factor levels to numeric
    model_df$plot_id <- as.numeric(model_df$plot_id)
    


#standardizing predictors
model_df$mean_temperature_c <- scale(model_df$mean_temperature_c)
model_df$bottom_do <- scale(model_df$bottom_do)
model_df$mean_light_lux <- scale(model_df$mean_light_lux)
model_df$mean_depth_m <- scale(model_df$mean_depth_m)
model_df$urchin_density <- scale(model_df$urchin_density)
model_df$site_density <- scale(model_df$site_density)
model_df$shannon_div <- scale(model_df$shannon_div)
model_df$simp_div <- scale(model_df$simp_div)
model_df$species_richness <- scale(model_df$species_richness)

#month needs to be turned back into a factor
model_df$month <- factor(model_df$month)
model_df$treatment <- as.factor(model_df$treatment)
model_df$gravel_size <- as.factor(model_df$gravel_size)


model_df$kelp_survival <- as.integer(model_df$kelp_survival)

#shannon 
model1 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  shannon_div +(1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month + shannon_div,
                  family = nbinom2, 
                  data = model_df)

summary(model1)
AIC(model1) #945

#simp
model2 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  simp_div +(1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month + simp_div,
                  family = nbinom2, 
                  data = model_df)

summary(model2)
AIC(model2) #935

#richness
model3 <- glmmTMB(kelp_survival ~ treatment + gravel_size + mean_light_lux + mean_temperature_c + site_density + urchin_density + bottom_do + month +
                  species_richness +(1 | site) ,
                  zi = ~ treatment + mean_temperature_c + site_density + gravel_size + 
                        mean_light_lux + urchin_density + bottom_do + month + species_richness,
                  family = nbinom2, 
                  data = model_df)

summary(model3)
AIC(model3) #943


#

#correlation matrix of predictors 
cor_matrix <- cor(model_df[, c("mean_light_lux", "mean_temperature_c", "bottom_salinity", "mean_depth_m", "month","bottom_do", "urchin_density", "percent_occuring_algae", "site_density", "urchin_density", "simp_div")],  use = "complete.obs")

print(cor_matrix)

corrplot(cor_matrix, method = "circle")

#check VIF factor 
lm_model <- lm(kelp_survival ~ mean_light_lux + percent_occuring_algae + 
               mean_temperature_c + mean_depth_m + bottom_do + month + urchin_density + site_density + plot_id + simp_div, 
               data = model_df)

vif(lm_model) #salinity, bottom_do, mean_temp, is too high 

```

