---
title: "PCA"
author: "Tessa Rehill"
date: "2024-12-12"
output: html_document
---

#notes 
  - need to make PCA predictors ALL the same (including non-numeric predictor -> factors -> numeric), include ALLLLL 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load libraries 
```{r}

library(dplyr)
library(vegan)
library(tidyr)

```

#load data 
```{r}

data <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/cleaned_data.csv")
sites <- read.csv("/Users/tessarehill/Dropbox/MSc/Exclosures/raw_data/metadata/site_coordinates.csv")

data <- merge(data, sites, by = "site", all.x = TRUE) #merge site data with data 
data$density <- as.factor(data$density)


```


#RUN PCA w/ site 
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)

# Prepare the data: Remove rows with NA values in both the numeric columns and site
data_clean <- data %>%
  select(kelp_survival, kelp_growth, urchin_density, mean_temperature_c, 
         mean_light_lux, bottom_temp, bottom_do, bottom_salinity, 
         bottom_ph, mean_depth_m, site_density, species_richness, 
         shannon_div, simp_div, site) %>%
  drop_na()  # Drop rows with NA values in any of the selected columns

# Perform PCA
pca <- prcomp(data_clean %>% select(-site), scale. = TRUE)  # Remove 'site' column for PCA

# Create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$site <- data_clean$site  # Add 'site' from cleaned data

# Plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(site))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Site", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Site")

# Plot PCA with urchin density (low, med, high)
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(site))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Site", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Site")
```
Group by treatment 
```{r}

library(tidyverse)
library(ggplot2)

# Prepare the data: Remove rows with NA values in both the numeric columns and site
data_clean <- data %>%
  select(kelp_survival, kelp_growth, urchin_density, mean_temperature_c, 
         mean_light_lux, bottom_temp, bottom_do, bottom_salinity, 
         bottom_ph, mean_depth_m, site_density, species_richness, 
         shannon_div, simp_div, treatment) %>%
  drop_na()  # Drop rows with NA values in any of the selected columns

# Perform PCA
pca <- prcomp(data_clean %>% select(-treatment), scale. = TRUE)  # Remove 'site' column for PCA

# Create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$treatment <- data_clean$treatment  # Add 'site' from cleaned data

# Plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(treatment))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Treatment", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Treatment")



```

Group by density 
```{r}

library(tidyverse)
library(ggplot2)

# Prepare the data: Remove rows with NA values in both the numeric columns and site
data$site <- as.factor(data$site)
data$site <- as.numeric(data$site)

data_clean <- data %>%
  select(kelp_survival, site, kelp_growth,  mean_temperature_c, 
         mean_light_lux,  bottom_do, bottom_salinity, 
         bottom_ph, mean_depth_m,  species_richness, 
         shannon_div, simp_div, density) %>%
  drop_na()  # Drop rows with NA values in any of the selected columns

# Perform PCA
pca <- prcomp(data_clean %>% select(-density), scale. = TRUE)  # Remove 'density' column for PCA

# Create a data frame with PCA scores and the 'site' variable
pca_scores <- as.data.frame(pca$x)
pca_scores$density <- data_clean$density  

# Plot PCA with sites as colors
ggplot(pca_scores, aes(x = PC1, y = PC2, color = as.factor(density))) +
  geom_point(size = 3) +
  labs(title = "PCA - Grouped by Density", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  scale_color_discrete(name = "Density")

#NEED TO FIGURE OUT WHY GREEN IS SO GROUPED ON ONE SIDE - 
```
#PCA


#THE GRAVEYARD
#create species richness and diversity indices for transect data
```{r}
data <- dens
# Assuming your dataframe is named 'data'
div_data <- data %>%
  rowwise() %>%
  mutate(
    # Combine species counts into a vector
    species_counts = list(c(total_n_ab, total_n_ru, total_n_gu, total_n_pu, 
                            total_n_sn, total_n_pyc, total_n_cuc)),
    # Calculate Species Richness
    species_richness = sum(species_counts > 0, na.rm = TRUE),
    # Shannon Diversity Index
    shannon_div = diversity(species_counts, index = "shannon"),
    # Simpson Diversity Index
    simp_div = diversity(species_counts, index = "simpson")
  ) %>%
  ungroup()

#rename urchin density to site density 
div_data <- div_data %>%
  rename(site_density = urchin_density)



```

#format env data
```{r}
#create env summary, for bottom temperature, do, ph, salinity
bottom_data <- env %>%
 group_by(site, month) %>%
  summarize(
    bottom_temp = temp[which.max(depth_m)],
    bottom_do = do[which.max(depth_m)],
    bottom_salinity = salinity[which.max(depth_m)],
    bottom_ph = ph[which.max(depth_m)],
    .groups = "drop" # Ungroups after summarization
  )

#make month a character
bottom_data <- bottom_data %>%
  mutate(month = as.character(month))
```

#Merge dataframe 
```{r}
# Merge hobo with plots
plots <- plots %>%
  left_join(hobo, by = c("site", "month"))

# Merge bottom_data with plots
plots <- plots %>%
  left_join(bottom_data, by = c("site", "month"))

# Merge div_data with plots
plots <- plots %>%
  left_join(div_data, by = c("site", "month"))

data <- plots

#select columns 
data <- data %>%
  select(site, plot_id, date, percent_occuring_algae, month, treatment, 
         gravel_size, plot, kelp_survival, urchin_density, kelp_growth, 
         mean_temperature_c, mean_light_lux, bottom_temp, bottom_do, 
         bottom_salinity, bottom_ph, mean_depth_m, site_density, 
         species_richness, shannon_div, simp_div)

write.csv(data, "/Users/tessarehill/Dropbox/MSc/Exclosures/processed_data/data.csv", row.names = FALSE)

env_data <- data %>%
  select(site,  date,  month, 
         mean_temperature_c, mean_light_lux, bottom_temp, bottom_do, 
         bottom_salinity, bottom_ph, mean_depth_m,  
         species_richness, shannon_div, simp_div)

#make numeric
data <- env_data %>%
  mutate(across(c(site, plot_id, date, month, treatment, gravel_size, plot), 
                ~ as.numeric(as.factor(.))))

summary(data)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
